### 2011-12-01 plos_template.tex

% Template for PLoS
% Version 1.0 January 2009
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{openSNP - Crowdsourcing Genome Wide Association Studies}
}
% Insert Author names, affiliations and corresponding author email.
\\
Bastian Greshake$^{1,\ast}$, 
Philipp Bayer$^{2}$, 
Fabian Zimmer$^{3,}$
\\
\bf{1} Bastian Greshake Frankfurt am Main, Germany
\\
\bf{2} Philipp Bayer somewhere, Australia
\\
\bf{3} Fabian Zimmer M\"unster, Germany
\\
$\ast$ E-mail: info@opensnp.org
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

\section*{Introduction}
Genome Wide Association Studies (GWAS) are an easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance. SNPs found through GWAS can be used to find candidate genes for a closer inspection or to predict disease risks. Genome Wide Association Studies make use of statistics to compare the alleles of patients to the alleles of healthy controls. By this the method does not allow to find causal differences but mere correlations. The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group (doi:10.1126/science.1109557). Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed (doi:10.1186/1471-2350-10-6.) and over 5000 SNPs have been linked to different diseases and traits in those studies %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table).

Since 2006 there are different companies like 23andMe, deCODEme or FamilyTreeDNA on the market which offer Direct-To-Consumer genetic testing. Those companies use DNA micro arrays to screen for around 1 million SNPs which are spread over the human genome. In return customers get an analysis of the results, as well as a raw file that includes the SNP-IDs and their respective allele for the customer. In 2011 23andMe alone had over 100.000 customers (http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/) and the company also recognizes the potential to do GWAS with that amount of data. They provide surveys to their customers that ask about traits and genetic diseases. With the consent of the customer those data will then be used for association studies. 23andMe published 3(?) articles in 2011 in which they replicate known findings but also find new associations for Parkinson's Disease. By activating their customer base they achieved to have over 30.000 individuals enrolled in those association studies.

Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to get hold of the data. This arises mainly due to privacy concerns of the customers. A possible solution to this can be a crowdsourcing platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent. In our study we investigated if there would be interest in such a crowd sourcing platform, how many people would be willing to share their genetic and phenotypic information with the public and built such a platform.

% Results and Discussion can be combined.
\section*{Results}

\subsection*{Subsection 1}

\subsection*{Subsection 2}

\section*{Discussion}

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

%\section*{References}
% The bibtex filename
\bibliography{template}

\section*{Figure Legends}
%\begin{figure}[!ht]
%\begin{center}
%%\includegraphics[width=4in]{figure_name.2.eps}
%\end{center}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption 
%should be left justified, as specified by the options to the caption 
%package.
%}
%\label{Figure_label}
%\end{figure}

\section*{Tables}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

\end{document}

### 2011-12-01 plos_template.tex

@@
 This arises mainly due to privacy concerns of the customers.
-A possible solution to this can be a crowdsourcing platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent.
+Nevertheless there are individual customers who are willingly sharing their data.
+Most do though by uploading it to their personal website or to software repositories like GitHub.
+While this is makes it possible to use the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
+While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
+Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data.
+
+A possible solution to this can be a community-driven platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent.
 In our study we investigated if there would be interest in such a crowd sourcing platform, how many people would be willing to share their genetic and phenotypic information with the public and built such a platform.

### 2011-12-02 plos_template.tex
started writing the results

@@
 
-Since 2006 there are different companies like 23andMe, deCODEme or FamilyTreeDNA on the market which offer Direct-To-Consumer genetic testing.
+Since 2006 there are different companies like 23andMe, deCODEme or FamilyTreeDNA on the market which offer Direct-To-Consumer (DTC) genetic testing.
 Those companies use DNA micro arrays to screen for around 1 million SNPs which are spread over the human genome.
@@
 
-\subsection*{Subsection 1}
-
+\subsection*{Survey on Sharing Genetic Information}
+229 people, 169 with a self-reported chromosomal sex of XY, 59 with a self-reported chromosomal sex of XX, participated in the survey.
+The mean age of the participants is 33 and over 80 \% reported their ethnicity as caucasian.
+40 \% of the participants are already customer of at least one DTC genetic testing company and further 30 \% of them plan to become one in the future.
+30 \% don't plan to become a DTC customer.
 \subsection*{Subsection 2}

### 2011-12-03 plos_template.tex
more work on the statistics-stuff

--- 2011-12-01 plos_template.tex
@@
 
-\subsection*{Subsection 1}
+\subsection*{Survey on Sharing Genetic Information}
+229 people, 180 with a self-reported chromosomal sex of XY, 56 with a self-reported chromosomal sex of XX, participated in the survey.
+The mean age of the participants is 33 (SD = 11,29) and over 81.7 \% reported their ethnicity as caucasian.
+39.7 \% of the participants are already customer of at least one DTC genetic testing company and further 30.1 \% of them plan to become one in the future.
+29.7 \% don't plan to become a DTC customer.
+There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).
 
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so, if the company does not share the data with third parties.
+6.6 \% of the participants would not share their data.
+There is no significant difference between sharing-habits between both chromosomal sexes (Somers-d).
+Those who are a customer of a DTC company or are planing to become one in the future are more likely to share their results, compared to those who don't plan to get themselves genotyped (Somers-d).
 \subsection*{Subsection 2}

### 2011-12-04 paper_draft.tex
Moved the paper draft into its own file, plos_template.tex should only contain the template

--- 2011-12-03 plos_template.tex
@@
 \section*{Introduction}
+
+Paragraph1:
+
+Investigate the link between genetic variation and phenotypic change
+Important because of desease stuff.
+
+Tool used: Genome Wide Association Studies.
+--> Describe in one sentence.
+--> 2005 first study
+--> Today 1200 GWAS performed
+
+genetic markers usually SNPs.
+SNPs definition. Allele definition.
+SNP can be linked to a phenotypic trait (around 5000 linked to desease)
+
+Paragraph2:
+
+How to obtain data for GWAS?
+--> SNPChip, used in many different species
+--> Now different companies for DTC genetic testing.
+--> Test results provided to customer and raw data
+--> Raw data can be used for GWAS, but closed
+
+Paragraph3:
+
+RAW data from 1000 of individuals could be used for multiple scientific
+purposes
+---> Large potential dataset Are the numbers available for dataset size used
+in scientic studies? 2005 93, 2008 500, 2011 23andme 100000? Plot?
+
+Resolution of the problem: Open Science
+Create repository for sharing this information. openSNP
+
+Paragraph4:
+
+Describe openSNP:
+Open Source, non-commercial
+Open Data
+Community driven --> Crowd sourcing
+
+Parses data from 23andMe deCODEme....
+--> Publications crawler (Maybe mention mendeley price?)
+        --PLOS and Mendeley
+
+Advantages for scientists: See paragraph 3
+
+Advantages for customers: 
+-->Make sense of the data
+-->Contribute to the public domain
+
 Genome Wide Association Studies (GWAS) are an easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance.

### 2011-12-04 plos_template.tex
finished survey-results

@@
 Those who are a customer of a DTC company or are planing to become one in the future are more likely to share their results, compared to those who don't plan to get themselves genotyped (Somers-d).
+
+There are significant differences between those people who are already genotyped and those who don't plan to get genotyped: (All those numbers are tukey-hsd test).
+The first group is more likely to agree to share their information because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
+
+On the other hand those hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
+
+Similarly those people who would share data with their DTC provider are more likely to agree on sharing the data, because they want to help scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
+
+Those participants who are not planning to get genotyped are more likely to agree to not share their data, because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
 \subsection*{Subsection 2}

### 2011-12-04 paper_draft.tex
added some keywords for the first paragraph of the discussion

@@
 
+Plot: Maybe plot the amount of new added datasets and the number of new phenotypes per
+
+Paragraph 1:
+
+First focus on biological issues:
+
+Polymorphisms to phenotypes, correlation only.
+Bias in the samples collected by openSNP?
+Sample size to small?
+Quality check for phenotypes?
+Mapping to haplotype blocks?
+Combination with the hapmap
+Manage increasing amount of data?
+
+Paragraph 2:
+from great power follows great responsibility!
+
+Imagine a world where all data is free...
+What privacy issues do arise? 
+Is there an effective mechanism to avoid abuse or
+genetic discrimination?
+How should governments treat the additional  data?
+
 % You may title this section "Methods" or "Models". 

### 2011-12-11 plos_template.tex
ups, added statsmonkey
wrote draft on methods of survey and better draft on software-methods

@@
 Philipp Bayer$^{2}$, 
-Fabian Zimmer$^{3,}$
+Fabian Zimmer$^{3}$,
+Julia Reda$^{4}$
 \\
@@
 \bf{3} Fabian Zimmer M\"unster, Germany
+\\
+\bf{4} Julia Reda, Mainz, Germany
 \\
@@
 While this is makes it possible to use the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
-While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
+While projects like the SNPedia (10.1093/nar/gkr798) try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
 Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data.
@@
 Those participants who are not planning to get genotyped are more likely to agree to not share their data, because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
-\subsection*{Subsection 2}
+\subsection*{openSNP platform}
 
@@
 \section*{Materials and Methods}
+\subsection*{Survey on Sharing Genetic Information}
+Done in Google Docs, questions on age, chromosomal sex, ethnicity, if they are already customer of a DTC company/planning to become one/don't plan to become one.
+If they are already a customer: Are they already sharing and are they already sharing phenotypic information?
+How strong on a scale of 1 = strongly disagree to  5 = strongly agree how do they feel about reasons to share their genetic data because...
+a) they want to help scientists b) get a personal benefit out of it (new treatments for illnesses) c) because of curiosity d) because it might bring relevant, personalized advertising.
+How strong on a scale of 1 = strongly disagree to 5 = strongly agree would they feel about not sharing their data because...
+a) they fear discrimination b) it is a breach of their privacy c) because of negative side-effects for their next of kin d) because they fear personalized spam.
+
+The results were analyzed in somestatisticstool.
+
+\subsection*{Technical realization of the platform}
+The main platform is realized using the web-framework Ruby on Rails.
+Postgres is used as the main database backend for Rails.
+The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia.
+
+The literature database of Mendeley is queried using the REST API, which delivers results in JSON.
+The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in XML.
+Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.
+
+Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are be mailed to users and queries to external resources are handled using the ruby-gem Resque and a Redis-server.
+Search-features on the platform itself are implemented using SOLR and the ruby-gem sunspot.
+Additionally data can be requested from openSNP using the Distributed Annotation System.
+The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer, which was modified by Gel et al.
+for use in easyDAS.
+
+The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr.
+The genetical and phenotypical data is licensed under Creative Commons Zero.
 

### 2011-12-11 plos_template.tex
small change
minor work on technical methods

@@
 Genome Wide Association Studies make use of statistics to compare the alleles of patients to the alleles of healthy controls.
-By this the method does not allow to find causal differences but mere correlations.
+By this the method it is not possbile to find causal differences but mere correlations.
 The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group (doi:10.1126/science.1109557).
@@
 The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia.
-
 The literature database of Mendeley is queried using the REST API, which delivers results in JSON.
@@
 Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.
+
+SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number.
+This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources.
 

### 2011-12-11 plos_template.tex
some more writing-stuff

@@
 \subsection*{Survey on Sharing Genetic Information}
-Done in Google Docs, questions on age, chromosomal sex, ethnicity, if they are already customer of a DTC company/planning to become one/don't plan to become one.
-If they are already a customer: Are they already sharing and are they already sharing phenotypic information?
-How strong on a scale of 1 = strongly disagree to  5 = strongly agree how do they feel about reasons to share their genetic data because...
-a) they want to help scientists b) get a personal benefit out of it (new treatments for illnesses) c) because of curiosity d) because it might bring relevant, personalized advertising.
-How strong on a scale of 1 = strongly disagree to 5 = strongly agree would they feel about not sharing their data because...
-a) they fear discrimination b) it is a breach of their privacy c) because of negative side-effects for their next of kin d) because they fear personalized spam.
+The survey was done with Google Docs and included questions on the age, chromosomal sex and ethnicity of the participants.
+Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one.
+If they are already a customer they also got asked if they already share their genetic and phenotypic data.
+All participants got asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".
 
-The results were analyzed in somestatisticstool.
+The survey also asked some scale-questions which measured how strong participants agree/disagree on different reasons to share or not to share their information with third parties.
+The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree.
+Motivations given to share data were "because you want to help scientists with their research", "because of possible personal benefits (e.g.
+getting treatments for a disease you have, possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity".
+Motivations given not to share data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", "because of the breach of your privacy" and " because of the fear of discrimination (e.g.
+by the employer, the state, some insurance company)".
+Additionally participants had the possibility to give own reasons to share or not to share their data.
+
+The results were analyzed with SPSS 19.
 

### 2011-12-13 paper_draft.tex
Added bib-file with two papers, started work on abstract and did some minor changes in introduction

--- 2011-12-04 paper_draft.tex
@@
 \\
-\bf{2} Philipp Bayer somewhere, Australia
+\bf{2} Philipp Bayer Gold Coast, Australia
 \\
@@
 \section*{Abstract}
+Genome wide association studies (GWAS) are a cheap and quick way to assess health risks by comparing Single Nucleotide Polymorphisms (SNPs) between groups of participants.
+Direct-To-Consumer Companies like 23andme offer their customers to sequence their SNPs alongside with an evaluation of the customer's genetic risks.
+However, the data 23andme and other companies generate is not accessible for other scientists, and withholds some information from their customers for various reasons.
+In this paper, we present an open approach to GWAS by introducing openSNP, a webpage which allows GWAS-customers to openly share their SNPs with scientists for free.
+% misses survey
 
@@
 
-Since 2006 there are different companies like 23andMe, deCODEme or FamilyTreeDNA on the market which offer Direct-To-Consumer (DTC) genetic testing.
-Those companies use DNA micro arrays to screen for around 1 million SNPs which are spread over the human genome.
+Since 2006 companies like 23andMe, deCODEme or FamilyTreeDNA offer Direct-To-Consumer (DTC) genetic testing.
+Those companies use DNA micro arrays to screen for around 1 million SNPs spread over the human genome.
 In return customers get an analysis of the results, as well as a raw file that includes the SNP-IDs and their respective allele for the customer.
-In 2011 23andMe alone had over 100.000 customers (http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/) and the company also recognizes the potential to do GWAS with that amount of data.
-They provide surveys to their customers that ask about traits and genetic diseases.
-With the consent of the customer those data will then be used for association studies.
-23andMe published 3(?) articles in 2011 in which they replicate known findings but also find new associations for Parkinson's Disease.
-By activating their customer base they achieved to have over 30.000 individuals enrolled in those association studies.
+In 2011 23andMe alone had over 100.000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} - the company recognizes the potential to perform GWAS with that amount of data by using surveys to ask their customers about traits and diseases.
+With the consent of the customer those data is used for association studies.
+23andMe published several articles in which they replicate known findings but also find new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}.
+Over 30,000 23andme-customers participated in those association studies.
 
-Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to get hold of the data.
-This arises mainly due to privacy concerns of the customers.
+Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data.
+This is mainly due to privacy concerns of the customers.
 Nevertheless there are individual customers who are willingly sharing their data.
-Most do though by uploading it to their personal website or to software repositories like GitHub.
-While this is makes it possible to use the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
+Most do so by uploading their data to their personal website or to open software repositories like GitHub.
+While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
 While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
 Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data.
+%WHY?
 
 A possible solution to this can be a community-driven platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent.
-In our study we investigated if there would be interest in such a crowd sourcing platform, how many people would be willing to share their genetic and phenotypic information with the public and built such a platform.
+We designed a survey to assess interest in such a crowd sourcing platform, in which we asked how many people would be willing to share their genetic and phenotypic information with the public.
 
@@
 \subsection*{Survey on Sharing Genetic Information}
-229 people, 180 with a self-reported chromosomal sex of XY, 56 with a self-reported chromosomal sex of XX, participated in the survey.
+229 participants filled out the survey.
+Of these 229, 180 had a self-reported chromosomal sex of XY and 56 carried a self-reported chromosomal sex of XX.
 The mean age of the participants is 33 (SD = 11,29) and over 81.7 \% reported their ethnicity as caucasian.
@@
 
-67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so, if the company does not share the data with third parties.
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would share if the company does not share the data with third parties.
 6.6 \% of the participants would not share their data.
 There is no significant difference between sharing-habits between both chromosomal sexes (Somers-d).
-Those who are a customer of a DTC company or are planing to become one in the future are more likely to share their results, compared to those who don't plan to get themselves genotyped (Somers-d).
-\subsection*{Subsection 2}
+Those who are a customer of a DTC company or are planning to become one in the future are more likely to share their results, as compared to those who don't plan to get themselves genotyped (Somers-d).
+
+\subsection*{Introducing openSNP}
+We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license.
+Users are encouraged to list as many phenotypes as possible due to a simple achievement-system similar to systems found in computer games.
 
 \section*{Discussion}
+to-do list together with outlook of personalised medicine
 
@@
 % The bibtex filename
-\bibliography{template}
+\bibliography{papers}
 

### 2011-12-17 paper_draft.tex
added some text, moved outline to outline.txt

--- 2011-12-13 paper_draft.tex
@@
 However, the data 23andme and other companies generate is not accessible for other scientists, and withholds some information from their customers for various reasons.
-In this paper, we present an open approach to GWAS by introducing openSNP, a webpage which allows GWAS-customers to openly share their SNPs with scientists for free.
+In this paper, we present an open approach to GWAS by introducing openSNP, a web platform which allows GWAS-customers to openly share their SNPs with scientists for free.
 % misses survey
@@
 \section*{Introduction}
-
-Paragraph1:
-
-Investigate the link between genetic variation and phenotypic change
-Important because of desease stuff.
-
-Tool used: Genome Wide Association Studies.
---> Describe in one sentence.
---> 2005 first study
---> Today 1200 GWAS performed
-
-genetic markers usually SNPs.
-SNPs definition. Allele definition.
-SNP can be linked to a phenotypic trait (around 5000 linked to desease)
-
-Paragraph2:
-
-How to obtain data for GWAS?
---> SNPChip, used in many different species
---> Now different companies for DTC genetic testing.
---> Test results provided to customer and raw data
---> Raw data can be used for GWAS, but closed
-
-Paragraph3:
-
-RAW data from 1000 of individuals could be used for multiple scientific
-purposes
----> Large potential dataset Are the numbers available for dataset size used
-in scientic studies? 2005 93, 2008 500, 2011 23andme 100000? Plot?
-
-Resolution of the problem: Open Science
-Create repository for sharing this information. openSNP
-
-Paragraph4:
-
-Describe openSNP:
-Open Source, non-commercial
-Open Data
-Community driven --> Crowd sourcing
-
-Parses data from 23andMe deCODEme....
---> Publications crawler (Maybe mention mendeley price?)
-        --PLOS and Mendeley
-
-Advantages for scientists: See paragraph 3
-
-Advantages for customers: 
--->Make sense of the data
--->Contribute to the public domain
 
@@
 While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
-Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data.
-%WHY?
+Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.
 
@@
 We designed a survey to assess interest in such a crowd sourcing platform, in which we asked how many people would be willing to share their genetic and phenotypic information with the public.
+Additionally we built a platform which allows customers of DTC genetic testing to publicate of genetic and phenotypic information and gives researchers multiple ways to reuse the data.
 
@@
 \subsection*{Survey on Sharing Genetic Information}
-229 participants filled out the survey.
-Of these 229, 180 had a self-reported chromosomal sex of XY and 56 carried a self-reported chromosomal sex of XX.
+229 people, 180 with a self-reported chromosomal sex of XY, 56 with a self-reported chromosomal sex of XX, participated in the survey.
 The mean age of the participants is 33 (SD = 11,29) and over 81.7 \% reported their ethnicity as caucasian.
@@
 
-67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would share if the company does not share the data with third parties.
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so, if the company does not share the data with third parties.
 6.6 \% of the participants would not share their data.
 There is no significant difference between sharing-habits between both chromosomal sexes (Somers-d).
-Those who are a customer of a DTC company or are planning to become one in the future are more likely to share their results, as compared to those who don't plan to get themselves genotyped (Somers-d).
+Those who are a customer of a DTC company or are planing to become one in the future are more likely to share their results, compared to those who don't plan to get themselves genotyped (Somers-d).
 
-\subsection*{Introducing openSNP}
-We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license.
-Users are encouraged to list as many phenotypes as possible due to a simple achievement-system similar to systems found in computer games.
+There are significant differences, tested by Tukey's HSD test, between those people who are already genotyped and those who don't plan to get genotyped: The first group is more likely to agree to share their information because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
+
+On the other hand those hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
+
+Similarly those people who would share data with their DTC provider are more likely to agree on sharing the data, because they want to help scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
+
+Those participants who are not planning to get genotyped are more likely to agree to not share their data, because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
+
+\subsection*{openSNP}
+We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
+Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges for their efforts.
+The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking.
+
+On the side of getting access to the data users can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation or can access a single download that includes all genotyping-files and all phenotypes in a comma separated table.
+Additionally users can access the genetic data through the Distributed Annotation System, which allows to get all data for specific chromosomes and specific positions on single chromosomes.
+
+Between the start of openSNP on 09/29/2011 and 12/17/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping results.
+In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes.
+The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users.
+There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.
 
 \section*{Discussion}
-to-do list together with outlook of personalised medicine
 
@@
 \section*{Materials and Methods}
+\subsection*{Survey on Sharing Genetic Information}
+The survey was done with Google Docs and included questions on the age, chromosomal sex and ethnicity of the participants.
+Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one.
+If they are already a customer they also got asked if they already share their genetic and phenotypic data.
+All participants got asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".
 
+The survey also asked some scale-questions which measured how strong participants agree/disagree on different reasons to share or not to share their information with third parties.
+The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree.
+Motivations given to share data were "because you want to help scientists with their research", "because of possible personal benefits (e.g.
+getting treatments for a disease you have, possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity".
+Motivations given not to share data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", "because of the breach of your privacy" and " because of the fear of discrimination (e.g.
+by the employer, the state, some insurance company)".
+Additionally participants had the possibility to give own reasons to share or not to share their data.
+
+The data of the survey was analyzed with SPSS 19.
+
+\subsection*{Technical realization of the platform}
+The main platform is realized using the web-framework Ruby on Rails.
+Postgres is used as the main database backend for Rails.
+The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia.
+The literature database of Mendeley is queried using the REST API, which delivers results in JSON.
+The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format.
+Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.
+
+SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number.
+This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources.
+
+Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are be mailed to users and queries to external resources are handled using the ruby-gem Resque and a Redis-server.
+Search-features on the platform itself are implemented using SOLR and the ruby-gem sunspot.
+Additionally data can be requested from openSNP using the Distributed Annotation System.
+The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer, which was modified by Gel et al.
+for use in easyDAS.
+
+The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr.
+The genetical and phenotypical data is licensed under Creative Commons Zero.
 % Do NOT remove this, even if you are not including acknowledgments

### 2011-12-19 paper_draft.tex

@@
 Philipp Bayer$^{2}$, 
-Fabian Zimmer$^{3,}$
+Fabian Zimmer$^{3}$,
+Julia Reda$^{4}$
 \\
-\bf{1} Bastian Greshake Frankfurt am Main, Germany
+\bf{1} Goethe University, Frankfurt am Main, Germany
 \\
-\bf{2} Philipp Bayer Gold Coast, Australia
+\bf{2} Bond University, Gold Coast, Australia
 \\
-\bf{3} Fabian Zimmer M\"unster, Germany
+\bf{3} Westf\"alische Wilhelms Universit\"at, M\"unster, Germany
+\\
+\bf{4} Johannes-Gutenberg University, Mainz, Germany
 \\
@@
 
-Genome Wide Association Studies (GWAS) are an easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance.
+Genome Wide Association Studies (GWAS) are a comparatively easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance.
 SNPs found through GWAS can be used to find candidate genes for a closer inspection or to predict disease risks.
@@
 
-On the other hand those hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
+On the other hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
 
@@
 \subsection*{openSNP}
-We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
-Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges for their efforts.
+We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
+Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile page.
+
 The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking.
+To reduce the amount of manual data curation openSNP tries to avoid the entry of the same phenotype or variation, but with a slightly different spelling by helping users at entering data by an autocompletion-feature which lists similar entries which are already in the openSNP-database.
 
@@
 
-Between the start of openSNP on 09/29/2011 and 12/17/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping results.
+Between the start of openSNP on 09/27/2011 and 12/18/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping files.
+Through this the openSNP database lists 69486471 SNPs which are distributed over 1938604 unique Rs-IDs.
 In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes.
+See figure n for a distribution of data acquisition over time.
+
+The mean number of users that have entered their variation for a single phenotype is 14.36 (SD 12.65), the median is 10.
+The distribution of how many users have entered their data per phenotype can be seen in figure n+1.
 The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users.
 There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.
+
+A total number of 15229 documents relevant to the SNP-IDs which are listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia.
+Of the primary literature 25 \% are released in Open Access-journals and can be freely free of charge by every user (Figure n+2).
+For usability reasons SNPs are ranked by the amount of information gathered by the external services.
+
+The external services themselves are ranked by how easy users can access information out of these sources.
+The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science.
+Lowest values are given to the Mendeley-results, as those aren't necessary freely available for every user.
+SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley-entry.
 
@@
 \subsection*{Survey on Sharing Genetic Information}
-The survey was done with Google Docs and included questions on the age, chromosomal sex and ethnicity of the participants.
+The survey was done with Google Docs and was distributed to possible participants through the 23andMe-community, the DIYBiology mailinglist, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
+Because of this the results are biased in a way that customers of DTC genetic testing are overrepresented.
+As we wanted to find out how people who have purchased DTC genetic testing care about sharing their data with third parties this does not lead to further problems.
+
+The survey included questions on the age, chromosomal sex and ethnicity of the participants.
 Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one.
@@
 Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.
+All databases are queried using the unique identifier of each SNP as search term.
 
@@
 This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources.
+Publications out of the different databases and the genotypes of the users are associated with single SNPs by the Rs-ID.
+Allele and genotype frequencies are are updated regularly, based on the data present in openSNP.
 
@@
 
+A flowchart of all services incorporated in openSNP and how users can upload or access the data is given in figure n+3.
 The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr.

### 2011-12-19 paper_draft.tex
added more citations

@@
 By this the method does not allow to find causal differences but mere correlations.
-The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group (doi:10.1126/science.1109557).
-Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed (doi:10.1186/1471-2350-10-6.) and over 5000 SNPs have been linked to different diseases and traits in those studies %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table).
+The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}.
+Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table).
 
@@
 While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
-While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
+While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
 Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.
@@
 Search-features on the platform itself are implemented using SOLR and the ruby-gem sunspot.
-Additionally data can be requested from openSNP using the Distributed Annotation System.
-The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer, which was modified by Gel et al.
-for use in easyDAS.
+Additionally data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}.
+The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al.
+for use in easyDAS \cite{GelMoreno2011}.
 

### 2011-12-19 paper_draft.tex
some more references

@@
 The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}.
-Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table).
+Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}.
+%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
 

### 2011-12-19 paper_draft.tex
started working on discussion

@@
 Nevertheless there are individual customers who are willingly sharing their data.
-Most do so by uploading their data to their personal website or to open software repositories like GitHub.
+Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}.
 While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
@@
 \section*{Discussion}
+Although prices for exome or even full genome sequencing are dropping rapidly, in comparison to this GWAS tend to stay cheaper and offer a possibility to get insights on genetic variations on a population level and allow an analysis of SNPs which are linked to different traits.
+Despite this benefit in terms of costs it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow the detection of the causes for this and needs a large enough sample which has to be statistically analysed by using sound methods.
+
+One way to bring costs for GWAS further down is to make use of already available genotyping results and datasets.
+Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for those genotypings by themselves.
+As we have found in the survey on the sharing such results many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.
+
+With openSNP we've build a platform that can be used by customers of DTC genetic testing to share their data to easily share their genetic and phenotypic data with a wide audience as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
+Customers of DTC genetic testing also benefit of an easy access to primary literature on SNPs and genetic variation they carry.
+While we don't have collected enough data to perform full scale GWAS yet this might be possible in the future, as user numbers are rising.
+
+By crowdsourcing the acquisition of genetic and phenotypic data openSNP faces the same problems as any other open platform on the Internet, namely we have to trust users about the data they upload and enter on openSNP.
+Additionally the quality of the data varies, especially in terms of accuracy on the phenotypic variation.
+While we try to suggest similar entries to users there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.
+
+There are two possible solutions to this problem: We could only allow some (trusted) users to enter new phenotypes or we could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start on.
+Both methods have their own disadvantage: In either case it makes it harder for users to enter their data and by this highers the bar for participation, which ultimately could lead to less data entered.
+Facing this trade-off we decided to keep entering data easy, at the cost that users who want to perform GWAS with the data need to perform more quality control.
+
+Another thing that should kept in mind is a possible bias, as we can't rule out the possibility that only a biased subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
 

### 2011-12-19 paper_draft.tex
finished first draft of discussion, added some dtc-privacy-pubs

@@
 
-Genome Wide Association Studies (GWAS) are a comparatively easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance.
-SNPs found through GWAS can be used to find candidate genes for a closer inspection or to predict disease risks.
+Genome Wide Association Studies (GWAS) are a comparatively easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which are of interest because of their medical relevance.
+Such SNPs of interest can be used to find candidate genes for a closer inspection or to predict disease risks or other traits.
 Genome Wide Association Studies make use of statistics to compare the alleles of patients to the alleles of healthy controls.
@@
 Those companies use DNA micro arrays to screen for around 1 million SNPs spread over the human genome.
-In return customers get an analysis of the results, as well as a raw file that includes the SNP-IDs and their respective allele for the customer.
+In return customers get an analysis of the results, as well as a raw file that includes the individual genotypes of the customer.
 In 2011 23andMe alone had over 100.000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} - the company recognizes the potential to perform GWAS with that amount of data by using surveys to ask their customers about traits and diseases.
@@
 Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data.
-This is mainly due to privacy concerns of the customers.
+This is mainly due to concerns about privacy, liability and consent.
 Nevertheless there are individual customers who are willingly sharing their data.
@@
 While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
-Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.
+Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.
 
-A possible solution to this can be a community-driven platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent.
+A possible solution to this can be a community-driven platform that aggregates genetic and phenotypic information of people who are willing to share their data with the general public and have given their informed consent.
 We designed a survey to assess interest in such a crowd sourcing platform, in which we asked how many people would be willing to share their genetic and phenotypic information with the public.
@@
 Additionally users can access the genetic data through the Distributed Annotation System, which allows to get all data for specific chromosomes and specific positions on single chromosomes.
+Users can discuss SNPs and phenotypes on the platform using a simple commenting-system or private messages.
 
@@
 Lowest values are given to the Mendeley-results, as those aren't necessary freely available for every user.
-SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley-entry.
+A entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley-entry.
 
@@
 Despite this benefit in terms of costs it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow the detection of the causes for this and needs a large enough sample which has to be statistically analysed by using sound methods.
+Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe?
+we could look at the twitter-timeline of @opensnporg, i publish papers i find there).
 
-One way to bring costs for GWAS further down is to make use of already available genotyping results and datasets.
+One way to bring costs for GWAS even further down is to make use of already available genotyping results and datasets.
 Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for those genotypings by themselves.
@@
 
-There are two possible solutions to this problem: We could only allow some (trusted) users to enter new phenotypes or we could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start on.
-Both methods have their own disadvantage: In either case it makes it harder for users to enter their data and by this highers the bar for participation, which ultimately could lead to less data entered.
+There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start on.
+Both methods have a disadvantage: In either case it makes it harder for users to enter their data and by this highers the bar for participation, which ultimately could lead to less data entered.
 Facing this trade-off we decided to keep entering data easy, at the cost that users who want to perform GWAS with the data need to perform more quality control.
 
-Another thing that should kept in mind is a possible bias, as we can't rule out the possibility that only a biased subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
+Another thing that should kept in mind is a possible bias in which data is available on openSNP, as we can't rule out the possibility that only a biased subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
+
+The advent of DTC genetic testing has lead to new ethical and social issues.
+Much of the critique on DTC genetic testing focusses on the practice of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and help them to put the new knowledge to good use.
+Less attention has been given to the privacy implications that come with this cheap way of obtaining genetical information.
+
+By making the own genetic and medical information public it becomes clear that the issue of privacy needs to get a closer view.
+Our survey has shown that people are concerned about their privacy and are concerned that stakeholders like employers, insurance companies, governments or advertisers misuse this information.
+Policy makers start to react to those changes by introducing laws like the \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact widely available genetic information.
+DTC genetic testing companies on their side also try to educate their customers about the risks of releasing genetic data.
+
+We transparently address the problem of privacy implications that come with releasing the data twice, during registering for openSNP and during the upload of the DTC results.
+Users have to check that they have read and understood the disclaimer about possible side-effects that can arise by making their data public.
+To further improve this process and get informed consent we are working to implement the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).
 

### 2011-12-19 paper_draft.tex
first rework of draft :)

@@
 {\Large
-\textbf{openSNP - Crowdsourcing Genome Wide Association Studies}
+\textbf{openSNP - Crowdsourcing Genome-Wide Association Studies}
 }
@@
 \section*{Abstract}
-Genome wide association studies (GWAS) are a cheap and quick way to assess health risks by comparing Single Nucleotide Polymorphisms (SNPs) between groups of participants.
-Direct-To-Consumer Companies like 23andme offer their customers to sequence their SNPs alongside with an evaluation of the customer's genetic risks.
-However, the data 23andme and other companies generate is not accessible for other scientists, and withholds some information from their customers for various reasons.
-In this paper, we present an open approach to GWAS by introducing openSNP, a web platform which allows GWAS-customers to openly share their SNPs with scientists for free.
+Genome-wide association studies are a cheap and quick way of assessing health risks by comparing Single Nucleotide Polymorphisms between groups of participants.
+Direct-To-Consumer Companies like 23andme offer SNP sequencing to their customers alongside with an evaluation of the customer's genetic risks.
+However, the data 23andme and other companies generate is not accessible to other scientists, and withholds some information from customers for various reasons.
+In this paper, we present an open approach to Genome-wide association studies by introducing openSNP, a web platform which allows customers to openly share their polymorphisms with scientists for free.
 % misses survey
@@
 
-Genome Wide Association Studies (GWAS) are a comparatively easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which are of interest because of their medical relevance.
+Genome-Wide Association Studies (GWAS) are a comparatively easy and cheap way of finding Single Nucleotide Polymorphisms (SNPs) of medical relevance.
 Such SNPs of interest can be used to find candidate genes for a closer inspection or to predict disease risks or other traits.
-Genome Wide Association Studies make use of statistics to compare the alleles of patients to the alleles of healthy controls.
-By this the method does not allow to find causal differences but mere correlations.
+Genome-Wide Association Studies make use of statistics for comparing the alleles of patients to the alleles of healthy controls.
+This method does not allow inference of causal differences but merely identifies correlations.
 The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}.
-Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}.
+Since the beginning, the number of participants in such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}.
 %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
 
-Since 2006 companies like 23andMe, deCODEme or FamilyTreeDNA offer Direct-To-Consumer (DTC) genetic testing.
-Those companies use DNA micro arrays to screen for around 1 million SNPs spread over the human genome.
-In return customers get an analysis of the results, as well as a raw file that includes the individual genotypes of the customer.
-In 2011 23andMe alone had over 100.000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} - the company recognizes the potential to perform GWAS with that amount of data by using surveys to ask their customers about traits and diseases.
-With the consent of the customer those data is used for association studies.
-23andMe published several articles in which they replicate known findings but also find new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}.
-Over 30,000 23andme-customers participated in those association studies.
+Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing.
+Those companies use DNA microarrays to screen for around 1 million SNPs spread over the human genome.
+In return customers get an analysis of the results, as well as a raw file that includes the customer's individual genotypes.
+In 2011, 23andMe alone had over 100,000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} - the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about traits and diseases.
+With the consent of the customer the data is used for association studies.
+23andMe has published several articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}.
+Over 30,000 23andme-customers have participated in those association studies.
 
-Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data.
+Although companies like 23andMe are willing to contribute to science, it is not easy for individual scientists to access the data.
 This is mainly due to concerns about privacy, liability and consent.
-Nevertheless there are individual customers who are willingly sharing their data.
+Nevertheless, there are individual customers who are willingly sharing their data.
 Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}.
-While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
-While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information.
+While this makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
+While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is not attached to the genetic information.
 Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.
 
-A possible solution to this can be a community-driven platform that aggregates genetic and phenotypic information of people who are willing to share their data with the general public and have given their informed consent.
-We designed a survey to assess interest in such a crowd sourcing platform, in which we asked how many people would be willing to share their genetic and phenotypic information with the public.
-Additionally we built a platform which allows customers of DTC genetic testing to publicate of genetic and phenotypic information and gives researchers multiple ways to reuse the data.
+A possible solution to this is a community-driven platform that aggregates genetic and phenotypic information of people who are willing to share their data with the general public and have given their informed consent.
+We designed an online survey to assess interest in such a crowdsourcing platform, in which we asked an audience interested in biology topics about their willingness to share their genetic and phenotypic information with the public.
+The aim of the survey was not to determine the percentage of the population interested in sharing genotypic information, but to get an idea of whether a crowdsourcing platform could get enough support to go forward with the project.
+Additionally, we built a platform which allows customers of DTC genetic testing to publicize genetic and phenotypic information and gives researchers multiple ways of reusing the data.
 
@@
 229 people, 180 with a self-reported chromosomal sex of XY, 56 with a self-reported chromosomal sex of XX, participated in the survey.
-The mean age of the participants is 33 (SD = 11,29) and over 81.7 \% reported their ethnicity as caucasian.
-39.7 \% of the participants are already customer of at least one DTC genetic testing company and further 30.1 \% of them plan to become one in the future.
-29.7 \% don't plan to become a DTC customer.
+The mean age of the participants is 33 (SD = 11,29) and 81.7 \% reported their ethnicity as caucasian.
+39.7 \% of the participants are already customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future.
+29.7 \% don't plan on becoming a DTC customer.
 There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).
 
-67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so, if the company does not share the data with third parties.
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company didn't share the data with third parties.
 6.6 \% of the participants would not share their data.
-There is no significant difference between sharing-habits between both chromosomal sexes (Somers-d).
-Those who are a customer of a DTC company or are planing to become one in the future are more likely to share their results, compared to those who don't plan to get themselves genotyped (Somers-d).
+There is no significant difference in readiness to share between chromosomal sexes (Somers-d).
+Those who are a customers of a DTC company or are planning on becoming one in the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d).
 
-There are significant differences, tested by Tukey's HSD test, between those people who are already genotyped and those who don't plan to get genotyped: The first group is more likely to agree to share their information because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
+There are significant differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped and those who are not planning on getting genotyped.
+The first group is likely to agree more strongly, on a five-point scale, with the following motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
 
-On the other hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
+On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because they fear negative consequences for their family members (mean difference = 0.733, SE = 0.21, p = 0.002) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
 
-Similarly those people who would share data with their DTC provider are more likely to agree on sharing the data, because they want to help scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
+Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with the following motivations for sharing than those who would not share their data with their DTC company: because they want to help scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
 
-Those participants who are not planning to get genotyped are more likely to agree to not share their data, because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their familiy members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
 
@@
 \subsection*{Survey on Sharing Genetic Information}
-The survey was done with Google Docs and was distributed to possible participants through the 23andMe-community, the DIYBiology mailinglist, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
-Because of this the results are biased in a way that customers of DTC genetic testing are overrepresented.
-As we wanted to find out how people who have purchased DTC genetic testing care about sharing their data with third parties this does not lead to further problems.
+The survey was done with Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
+Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
+As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
 
-The survey included questions on the age, chromosomal sex and ethnicity of the participants.
-Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one.
-If they are already a customer they also got asked if they already share their genetic and phenotypic data.
-All participants got asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".
+The survey included demographics such as age, chromosomal sex and ethnicity of the participants.
+Furthermore, it included questions on their (planned) customership of a DTC company.
+If they already were customers, they were also asked if they were already sharing their genetic and phenotypic data.
+All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".
 
-The survey also asked some scale-questions which measured how strong participants agree/disagree on different reasons to share or not to share their information with third parties.
+The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their information with third parties.
 The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree.
-Motivations given to share data were "because you want to help scientists with their research", "because of possible personal benefits (e.g.
-getting treatments for a disease you have, possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity".
-Motivations given not to share data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", "because of the breach of your privacy" and " because of the fear of discrimination (e.g.
+Motivations queried for sharing data were "because I want to help scientists with their research", "because of possible personal benefits (e.g.
+getting treatments for a disease I have, possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity".
+Motivations queried for not sharing data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", "because of the breach of my privacy" and " because of the fear of discrimination (e.g.
 by the employer, the state, some insurance company)".
-Additionally participants had the possibility to give own reasons to share or not to share their data.
+Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
 
-The data of the survey was analyzed with SPSS 19.
+The survey data was analyzed with SPSS 19.
 

### 2011-12-20 paper_draft.tex
reworking of the draft, more references

@@
 We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
-Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile page.
+Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages.
 
-The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking.
-To reduce the amount of manual data curation openSNP tries to avoid the entry of the same phenotype or variation, but with a slightly different spelling by helping users at entering data by an autocompletion-feature which lists similar entries which are already in the openSNP-database.
+The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories.
+Every user can add completely new phenotypes that are not yet queried by openSNP.
+To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database.
 
-On the side of getting access to the data users can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation or can access a single download that includes all genotyping-files and all phenotypes in a comma separated table.
-Additionally users can access the genetic data through the Distributed Annotation System, which allows to get all data for specific chromosomes and specific positions on single chromosomes.
-Users can discuss SNPs and phenotypes on the platform using a simple commenting-system or private messages.
+OpenSNP offers extensive access to the data uploaded by users.
+Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table.
+Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes.
+Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.
 
-Between the start of openSNP on 09/27/2011 and 12/18/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping files.
-Through this the openSNP database lists 69486471 SNPs which are distributed over 1938604 unique Rs-IDs.
-In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes.
+Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files.
+The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
+In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes.
 See figure n for a distribution of data acquisition over time.
 
-The mean number of users that have entered their variation for a single phenotype is 14.36 (SD 12.65), the median is 10.
+The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10.
 The distribution of how many users have entered their data per phenotype can be seen in figure n+1.
-The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users.
-There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.
+The phenotype provided by the most users is the eye color, which has been entered by 54 users.
+There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.
 
-A total number of 15229 documents relevant to the SNP-IDs which are listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia.
-Of the primary literature 25 \% are released in Open Access-journals and can be freely free of charge by every user (Figure n+2).
-For usability reasons SNPs are ranked by the amount of information gathered by the external services.
+In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs.
+A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia.
+Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2).
+For usability reasons, SNPs are ranked by the amount of information gathered by the external services.
 
-The external services themselves are ranked by how easy users can access information out of these sources.
+The external services themselves are ranked by how easily users can access information from these sources.
 The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science.
-Lowest values are given to the Mendeley-results, as those aren't necessary freely available for every user.
-A entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley-entry.
+Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user.
+An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
 
 \section*{Discussion}
-Although prices for exome or even full genome sequencing are dropping rapidly, in comparison to this GWAS tend to stay cheaper and offer a possibility to get insights on genetic variations on a population level and allow an analysis of SNPs which are linked to different traits.
-Despite this benefit in terms of costs it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow the detection of the causes for this and needs a large enough sample which has to be statistically analysed by using sound methods.
+Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits.
+Despite this benefit in terms of cost, it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow inference on the cause for any correlation.
+Furthermore, GWAS need a large enough sample which has to be statistically analysed by using sound methods.
 Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe?
@@
 
-One way to bring costs for GWAS even further down is to make use of already available genotyping results and datasets.
-Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for those genotypings by themselves.
-As we have found in the survey on the sharing such results many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.
+One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets.
+Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for the genotyping by themselves.
+As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.
 
-With openSNP we've build a platform that can be used by customers of DTC genetic testing to share their data to easily share their genetic and phenotypic data with a wide audience as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
-Customers of DTC genetic testing also benefit of an easy access to primary literature on SNPs and genetic variation they carry.
-While we don't have collected enough data to perform full scale GWAS yet this might be possible in the future, as user numbers are rising.
+With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
+Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry.
+While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
 
-By crowdsourcing the acquisition of genetic and phenotypic data openSNP faces the same problems as any other open platform on the Internet, namely we have to trust users about the data they upload and enter on openSNP.
-Additionally the quality of the data varies, especially in terms of accuracy on the phenotypic variation.
-While we try to suggest similar entries to users there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.
+By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP.
+Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation.
+While we try to suggest similar entries to the users, there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.
 
-There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start on.
-Both methods have a disadvantage: In either case it makes it harder for users to enter their data and by this highers the bar for participation, which ultimately could lead to less data entered.
-Facing this trade-off we decided to keep entering data easy, at the cost that users who want to perform GWAS with the data need to perform more quality control.
+There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start.
+Both methods have a disadvantage: In either case it makes it harder for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered.
+Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform more quality control.
 
-Another thing that should kept in mind is a possible bias in which data is available on openSNP, as we can't rule out the possibility that only a biased subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
+Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
 
-The advent of DTC genetic testing has lead to new ethical and social issues.
-Much of the critique on DTC genetic testing focusses on the practice of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and help them to put the new knowledge to good use.
-Less attention has been given to the privacy implications that come with this cheap way of obtaining genetical information.
+The advent of DTC genetic testing has led to new ethical and social issues.
+Much of the critique on DTC genetic testing focusses on the practice of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.
+Less attention has been paid to the privacy implications that come with this cheap way of obtaining genetical information \cite{Caulfield2011,Joh2011}.
 
-By making the own genetic and medical information public it becomes clear that the issue of privacy needs to get a closer view.
-Our survey has shown that people are concerned about their privacy and are concerned that stakeholders like employers, insurance companies, governments or advertisers misuse this information.
-Policy makers start to react to those changes by introducing laws like the \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact widely available genetic information.
-DTC genetic testing companies on their side also try to educate their customers about the risks of releasing genetic data.
+With a move towards making one's own genetic and medical information public, it becomes clear that the issue of privacy needs to get a closer look.
+Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information.
+Policy makers start to react to those changes by introducing laws like the \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of widely available genetic information.
+DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
 
-We transparently address the problem of privacy implications that come with releasing the data twice, during registering for openSNP and during the upload of the DTC results.
-Users have to check that they have read and understood the disclaimer about possible side-effects that can arise by making their data public.
-To further improve this process and get informed consent we are working to implement the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).
+We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during the upload of the DTC genetic testing results.
+Users have to confirm that they have read and understood the disclaimer about possible side-effects that can arise by publishing their data.
+To further improve this process and get informed consent, we are working on implementing the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).
 
@@
 \subsection*{Technical realization of the platform}
-The main platform is realized using the web-framework Ruby on Rails.
+The main platform is realized using the web framework Ruby on Rails.
 Postgres is used as the main database backend for Rails.
-The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia.
+The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs which can be found in SNPedia.
 The literature database of Mendeley is queried using the REST API, which delivers results in JSON.
 The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format.
-Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.
-All databases are queried using the unique identifier of each SNP as search term.
+Summaries on SNPs are provided by SNPedia, through querying the content via the MediaWiki API.
+All databases are queried using the unique identifier of each SNP as the search term.
 
 SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number.
-This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources.
-Publications out of the different databases and the genotypes of the users are associated with single SNPs by the Rs-ID.
-Allele and genotype frequencies are are updated regularly, based on the data present in openSNP.
+This is a common format, which is employed by the NCBI dbSNP database and is also widely used and easily parsed from different literature sources.
+Publications from the different databases as well as the users' genotypes are associated with individual SNPs by the Rs-ID.
+Allele and genotype frequencies are updated regularly, based on the data present in openSNP.
 
-Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are be mailed to users and queries to external resources are handled using the ruby-gem Resque and a Redis-server.
-Search-features on the platform itself are implemented using SOLR and the ruby-gem sunspot.
-Additionally data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}.
-The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al.
+Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources are handled using the ruby gem Resque and a Redis server.
+Search features on the platform itself are implemented using SOLR and the ruby gem Sunspot.
+Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}.
+The required data is stored in a mySQL database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al.
 for use in easyDAS \cite{GelMoreno2011}.
 
-A flowchart of all services incorporated in openSNP and how users can upload or access the data is given in figure n+3.
+A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure n+3.
 The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr.

### 2012-01-22 paper_draft.tex
Added a new introduction, structured the results and discussion, moved a lot of paragraphs

@@
 %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
+\\
+\\
+FZ REWRITE 
+\\
+The comparison of DNA sequences from indivduals in a population reveals sites that are variable. 
+These variations are called Single Nucleotide Polymorphisms (SNP). Each different nucleotide 
+at a specific site is defined as an allele. FZ: Explain genotype!!
+Different methods have been developed to measure/find/access/sequence SNPs in an individual, a
+process called genotyping. Microarrays are a cheap and fast way to genotype individuals, in comparison
+to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
+allele frequencies can be used to associated SNPs with for example diseass. Genome-Wide Assocation Studies
+(GWAS) make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This
+enables GWAS to find significant overrepresantations of alleles in patients and associate them wir a disease.
+This method does not allow inference of causal differences but merely identifies correlations. 
+The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
+to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
+such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
+5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
+%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
 
@@
 Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}.
-While this makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public.
+
+FZ: The data is scattered and unorganized, making it hard to use this infornomation.
 While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is not attached to the genetic information.
@@
 
-A possible solution to this is a community-driven platform that aggregates genetic and phenotypic information of people who are willing to share their data with the general public and have given their informed consent.
-We designed an online survey to assess interest in such a crowdsourcing platform, in which we asked an audience interested in biology topics about their willingness to share their genetic and phenotypic information with the public.
-The aim of the survey was not to determine the percentage of the population interested in sharing genotypic information, but to get an idea of whether a crowdsourcing platform could get enough support to go forward with the project.
-Additionally, we built a platform which allows customers of DTC genetic testing to publicize genetic and phenotypic information and gives researchers multiple ways of reusing the data.
+We designed an online survey to evaluate the support in the community for a crowdsourced online plattform.
+Here we present the results of the survey in combination with the openSNP project.
+The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources.
+The data is made available to research via a Creative-Commones-Zero licence.
 
@@
 \subsection*{openSNP}
-We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
+
+The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers.
+DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context.
+
+\subsubsection*{Sharing genotypic information}
+Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
+Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files.
+The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
+\subsubsection*{Crowdsourcing phenotypes}
 Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages.
-
 The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories.
@@
 To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database.
-
+In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes.
+See figure n for a distribution of data acquisition over time.
+The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10.
+The distribution of how many users have entered their data per phenotype can be seen in figure n+1.
+The phenotype provided by the most users is the eye color, which has been entered by 54 users.
+There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.
+\subsubsection*{Connection to external services}
+In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs.
+A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia.
+Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2).
+For usability reasons, SNPs are ranked by the amount of information gathered by the external services.The external services themselves are ranked by how easily users can access information from these sources.
+The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science.
+Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user.
+An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
+\subsubsection*{Data access}
 OpenSNP offers extensive access to the data uploaded by users.
@@
 
-Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files.
-The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
-In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes.
-See figure n for a distribution of data acquisition over time.
+\section*{Discussion}
 
-The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10.
-The distribution of how many users have entered their data per phenotype can be seen in figure n+1.
-The phenotype provided by the most users is the eye color, which has been entered by 54 users.
-There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.
-
-In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs.
-A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia.
-Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2).
-For usability reasons, SNPs are ranked by the amount of information gathered by the external services.
-
-The external services themselves are ranked by how easily users can access information from these sources.
-The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science.
-Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user.
-An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
-
-\section*{Discussion}
+\subsubsection*{Survey issues}
+Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
+As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
+\subsubsection*{GWAS and open data}
 Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits.
@@
 Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for the genotyping by themselves.
-As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.
-
-With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
-Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry.
-While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
 
@@
 Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
+\subsubsection*{Why should a user publish their genetic informationß Pro vs. Con}
+With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
+Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry.
+While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
+\subsubsection*{Privacy and health implications and ethical considerations}
+
+As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.
 
@@
 The survey was done with Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
-Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
-As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
 

### 2012-01-22 paper_draft.tex
some changes in the introduction

@@
 
-Genome-Wide Association Studies (GWAS) are a comparatively easy and cheap way of finding Single Nucleotide Polymorphisms (SNPs) of medical relevance.
-Such SNPs of interest can be used to find candidate genes for a closer inspection or to predict disease risks or other traits.
-Genome-Wide Association Studies make use of statistics for comparing the alleles of patients to the alleles of healthy controls.
-This method does not allow inference of causal differences but merely identifies correlations.
-The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}.
-Since the beginning, the number of participants in such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}.
-%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
-\\
-\\
-FZ REWRITE 
-\\
-The comparison of DNA sequences from indivduals in a population reveals sites that are variable. 
-These variations are called Single Nucleotide Polymorphisms (SNP). Each different nucleotide 
-at a specific site is defined as an allele. FZ: Explain genotype!!
-Different methods have been developed to measure/find/access/sequence SNPs in an individual, a
-process called genotyping. Microarrays are a cheap and fast way to genotype individuals, in comparison
+The comparison of DNA sequences from individuals in a population reveals sites which are variable. 
+If differences between individuals span only over a single nucleotide then these variations are also called Single Nucleotide Polymorphisms (SNP) and different nucleotides 
+at a specific site are defined as an allele. For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
+Different methods have been developed to read SNPs in an individual, a
+process called genotyping. Microarrays are a cheap, fast and widespread used way to genotype individuals, in comparison
 to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
-allele frequencies can be used to associated SNPs with for example diseass. Genome-Wide Assocation Studies
+allele frequencies can be used to associated SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
 (GWAS) make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This
-enables GWAS to find significant overrepresantations of alleles in patients and associate them wir a disease.
+enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
 This method does not allow inference of causal differences but merely identifies correlations. 
@@
 
-FZ: The data is scattered and unorganized, making it hard to use this infornomation.
-While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is not attached to the genetic information.
-Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.
+This data is scattered and unorganized, making it hard to use this information.
+While projects like SNPedia try to keep track of all the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is most often not attached to the genetic information.
+Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.
 
@@
 
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their familiy members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
 
@@
 Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
-\subsubsection*{Why should a user publish their genetic informationß Pro vs. Con}
+\subsubsection*{Why should a user publish their genetic information? Pro vs. Con}
 With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.

### 2012-01-23 paper_draft.tex
New structure for the sharing genotypes and the crowdsourcing phenotypes section. The verbatim package is used to create multi-line comments

@@
 
+%FZ: Enable the comment command
+\usepackage{verbatim}
 % Text layout
@@
 DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context.
+\begin{comment}
+FZ: That is a good summary, but as Bastian pointed out the reuse is not only
+enabled in a scientific context?
+\end{comment}
 
+\begin{comment}
+FZ:
+\end{comment}
 \subsubsection*{Sharing genotypic information}
-Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
-Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files.
-The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
+Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree to/via a webinterface to the openSNP 
+project. The uploaded data is published under the Creative Commons Zero-license, 
+which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
+allows a complete reuse of the data without any constraints.
+\begin{comment}
+FZ: If that is true, rephrase sentence above
+FZ: Made two sentences out of one.
+\end{comment}
+Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed 
+up with openSNP, 79 of which have uploaded their genotyping files. The openSNP 
+database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
+
 \subsubsection*{Crowdsourcing phenotypes}
-Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages.
-The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories.
-Every user can add completely new phenotypes that are not yet queried by openSNP.
-To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database.
-In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes.
-See figure n for a distribution of data acquisition over time.
-The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10.
-The distribution of how many users have entered their data per phenotype can be seen in figure n+1.
-The phenotype provided by the most users is the eye color, which has been entered by 54 users.
-There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.
+Beside uploading genotypic data, users may list new phenotypes that are not yet 
+queried by openSNP. The specification of these phenotypes  is open and not limited 
+to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
+the expression and spelling of the same phenotype or variation. We implemented an 
+autocompletion-feature which helps users to reuse already entered phenotypes.
+Users are encouraged to list as many phenotypes as possible through a simple 
+achievement system which rewards users that upload their data and enter phenotypic 
+information with small badges that are shown on their profile pages.
+\begin{comment}
+FZ: I changed the structure of the paragraph to make it a little bit more
+concise.
+I think its best to connect the phenotypes to the above paragraph, explain
+the procedure and then talk about the achievments
+\end{comment}
+In the same timeframe that was mentioned above, all users combined have 
+entered a total of 675 variations on 47 different phenotypes. See figure n for a 
+distribution of data acquisition over time.
+The mean number of users that have entered their variations for a single phenotype 
+is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
+entered their data per phenotype can be seen in figure n+1. The phenotype provided 
+y the most users is the eye color, which has been entered by 54 users. There are 
+two phenotypes which have so far only been provided by a single user: 
+The SAT Writing score and triglyceride-levels.
+\begin{comment}
+FZ: I think we should clarify the difference between variaton and phenotype!
+\end{comment}
+
 \subsubsection*{Connection to external services}

### 2012-01-23 paper_draft.tex
Added new commenting system taken from here: http://stackoverflow.com/questions/539791/tips-for-collaboratively-editing-a-latex-document/539848#539848; changed already present comments

@@
 %% PLEASE INCLUDE ALL MACROS BELOW
+\long\def\authornote#1{%
+        \leavevmode\unskip\raisebox{-3.5pt}{\rlap{$\scriptstyle\diamond$}}%
+        \marginpar{\raggedright\hbadness=10000
+        \def\baselinestretch{0.8}\tiny
+        \it #1\par}}
+\newcommand{\bastian}[1]{\authornote{BG: #1}}
+\newcommand{\fabian}[1]{\authornote{FZ: #1}}
 
@@
 DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context.
-\begin{comment}
-FZ: That is a good summary, but as Bastian pointed out the reuse is not only
-enabled in a scientific context?
-\end{comment}
-
-\begin{comment}
-FZ:
-\end{comment}
+\fabian{That is a good summary, but as Bastian pointed out the reuse is not only
+enabled in a scientific context?}
 \subsubsection*{Sharing genotypic information}
@@
 allows a complete reuse of the data without any constraints.
-\begin{comment}
-FZ: If that is true, rephrase sentence above
-FZ: Made two sentences out of one.
-\end{comment}
+\fabian{If that is true, rephrase sentence above. Made two sentences out of one.}
 Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed 
@@
 information with small badges that are shown on their profile pages.
-\begin{comment}
-FZ: I changed the structure of the paragraph to make it a little bit more
-concise.
-I think its best to connect the phenotypes to the above paragraph, explain
-the procedure and then talk about the achievments
-\end{comment}
+\fabian{
+I changed the structure of the paragraph to make it a little bit more
+concise.I think its best to connect the phenotypes to the above paragraph, explain
+the procedure and then talk about the achievments.}
+
 In the same timeframe that was mentioned above, all users combined have 
@@
 The SAT Writing score and triglyceride-levels.
-\begin{comment}
-FZ: I think we should clarify the difference between variaton and phenotype!
-\end{comment}
+\fabian{I think we should clarify the difference between variaton and phenotype!}
 

### 2012-01-29 paper_draft.tex
worked on fabians comments

@@
 The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers.
-DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context.
-\fabian{That is a good summary, but as Bastian pointed out the reuse is not only
-enabled in a scientific context?}
+DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers but the data can also be used for other research and in any other context.
+\bastian{Changed this to make it more true ;)}
 \subsubsection*{Sharing genotypic information}
@@
 allows a complete reuse of the data without any constraints.
-\fabian{If that is true, rephrase sentence above. Made two sentences out of one.}
 Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed 
@@
 information with small badges that are shown on their profile pages.
-\fabian{
-I changed the structure of the paragraph to make it a little bit more
-concise.I think its best to connect the phenotypes to the above paragraph, explain
-the procedure and then talk about the achievments.}
 
 In the same timeframe that was mentioned above, all users combined have 
-entered a total of 675 variations on 47 different phenotypes. See figure n for a 
+entered a total of 675 variations on 47 different phenotypes with those variations being the different values on a given trait or phenotype. 
+\bastian{Clarification of variation/phenotype?}
+See figure n for a 
 distribution of data acquisition over time.
@@
 The SAT Writing score and triglyceride-levels.
-\fabian{I think we should clarify the difference between variaton and phenotype!}
 

### 2012-02-03 paper_draft.tex
Fixed some typos, introduced JSON, changed some formulations

@@
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
+
 \subsubsection*{Data access}
@@
 Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table.
-Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes.
-Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.
+The genetic data is also accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes.
+The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format.
+Users can discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.
 
@@
 By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP.
-Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation.
-While we try to suggest similar entries to the users, there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.
+Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, with users entering data in different measurement systems.
+While we try to suggest similar entries to the users, there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
 
@@
 
-Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
-\subsubsection*{Why should a user publish their genetic information? Pro vs. Con}
+Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, along with phenotypic information of medical relevance.
+
+\subsubsection*{Why should a user publish their genetic information? Pros vs. Cons}
 With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
@@
 \subsection*{Survey on Sharing Genetic Information}
-The survey was done with Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
+The survey was performed using Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
 
@@
 
-Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources are handled using the ruby gem Resque and a Redis server.
+Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources are handled using the ruby gem Resque and the standalone server Redis.
 Search features on the platform itself are implemented using SOLR and the ruby gem Sunspot.
 Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}.
-The required data is stored in a mySQL database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al.
+The required data is stored in a PostgreSQL database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al.
 for use in easyDAS \cite{GelMoreno2011}.
+% probably no proserver, more own stuff
 

### 2012-02-08 paper_draft.tex
Wrote paragraph about the current DAS-methods

@@
 for use in easyDAS \cite{GelMoreno2011}.
-% probably no proserver, more own stuff
+Requested data is delivered in XML-format to facilitate parsing.
+% kick out proserver?
+what's the citation Finn2007 for?
+
+OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented.
+Currently, two methods are implemented: \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, based on the user's query.
+The second method is \textit{sources}, which advertises all DAS-sources for all genotypes present in openSNP.
 

### 2012-02-27 paper_draft.tex
Added some details on DAS, changed some statistics (phenotypes missing, as user_snps is still counting), fixed some types, changed some sentences.

@@
 Genome-wide association studies are a cheap and quick way of assessing health risks by comparing Single Nucleotide Polymorphisms between groups of participants.
-Direct-To-Consumer Companies like 23andme offer SNP sequencing to their customers alongside with an evaluation of the customer's genetic risks.
+Direct-To-Consumer (DTC) Companies like 23andme offer SNP sequencing to their customers alongside with an evaluation of the customer's genetic risks.
 However, the data 23andme and other companies generate is not accessible to other scientists, and withholds some information from customers for various reasons.
 In this paper, we present an open approach to Genome-wide association studies by introducing openSNP, a web platform which allows customers to openly share their polymorphisms with scientists for free.
-% misses survey
+We also present results of a survey to assess the willingness of DTC-customers to openly share their genotyping and polymorphisms.
 
@@
 allele frequencies can be used to associated SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
-(GWAS) make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This
+(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
 enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
@@
 Those companies use DNA microarrays to screen for around 1 million SNPs spread over the human genome.
-In return customers get an analysis of the results, as well as a raw file that includes the customer's individual genotypes.
+In return, customers get an analysis of the results, as well as a file that includes the customer's raw individual genotypes.
 In 2011, 23andMe alone had over 100,000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} - the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about traits and diseases.
@@
 \subsection*{Survey on Sharing Genetic Information}
-229 people, 180 with a self-reported chromosomal sex of XY, 56 with a self-reported chromosomal sex of XX, participated in the survey.
-The mean age of the participants is 33 (SD = 11,29) and 81.7 \% reported their ethnicity as caucasian.
+229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in the survey.
+The mean age of the participants is 33 (SD = 11,29).
+81.7 \% reported their ethnicity as caucasian.
 39.7 \% of the participants are already customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future.
@@
 allows a complete reuse of the data without any constraints.
-Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed 
-up with openSNP, 79 of which have uploaded their genotyping files. The openSNP 
-database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
+Between the start of openSNP on 09/27/2011 and 27/02/2012, 286 people have signed 
+up with openSNP, and 117 genotyping files were made available. The openSNP 
+database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.
 
@@
 As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
+\philipp{"Because of this approach", what approach?}
+
 \subsubsection*{GWAS and open data}
@@
 \subsubsection*{Why should a user publish their genetic information? Pros vs. Cons}
+\philipp{Should this be a subsubsection, not rather a subsection?}
 With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
@@
 \subsection*{Technical realization of the platform}
-The main platform is realized using the web framework Ruby on Rails.
-Postgres is used as the main database backend for Rails.
+The main platform is realized using the web framework Ruby on Rails 3.0.10.
+Postgres 9.2 is used as the main database backend for Rails.
 The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs which can be found in SNPedia.
@@
 
-SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number.
+SNPs are catalogued by their unique identifier, which consists of a prefix (mostly \textit{rs}, rarely \textit{i}) and a unique number.
 This is a common format, which is employed by the NCBI dbSNP database and is also widely used and easily parsed from different literature sources.
@@
 Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources are handled using the ruby gem Resque and the standalone server Redis.
-Search features on the platform itself are implemented using SOLR and the ruby gem Sunspot.
+Search features on the platform itself are implemented using Solr and the ruby gem Sunspot.
 Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}.
-The required data is stored in a PostgreSQL database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al.
-for use in easyDAS \cite{GelMoreno2011}.
+The required data is stored in a PostgreSQL database.
 Requested data is delivered in XML-format to facilitate parsing.
-% kick out proserver?
-what's the citation Finn2007 for?
+Additionally, users can request data in the JSON-format, using a system not specified in any standard.
 

### 2012-03-03 paper_draft.tex
Reverted some stats to be congruent with pics
rewored draft with fabian

@@
 \newcommand{\fabian}[1]{\authornote{FZ: #1}}
-
+\newcommand{\philipp}[1]{\authornote{PB: #1}}
 %% END MACROS SECTION
-
 \begin{document}
@@
 \section*{Abstract}
-Genome-wide association studies are a cheap and quick way of assessing health risks by comparing Single Nucleotide Polymorphisms between groups of participants.
-Direct-To-Consumer (DTC) Companies like 23andme offer SNP sequencing to their customers alongside with an evaluation of the customer's genetic risks.
-However, the data 23andme and other companies generate is not accessible to other scientists, and withholds some information from customers for various reasons.
-In this paper, we present an open approach to Genome-wide association studies by introducing openSNP, a web platform which allows customers to openly share their polymorphisms with scientists for free.
-We also present results of a survey to assess the willingness of DTC-customers to openly share their genotyping and polymorphisms.
+Genome-wide association studies are a cheap and quick way of assessing health risks by comparing Single Nucleotide Polymorphisms 
+between groups of participants. Direct-To-Consumer (DTC) Companies like 23andme offer SNP sequencing to their customers alongside 
+with an evaluation of the customer's genetic risks. However, the data 23andme and other companies generate is not accessible 
+to other scientists, and withholds some information from customers for various reasons. In this paper, we present an open 
+approach to Genome-wide association studies by introducing openSNP, a web platform which allows customers to openly share 
+their polymorphisms with scientists for free. We also present results of a survey to assess the willingness of DTC-customers 
+to openly share their genotyping and polymorphisms.
 
@@
 
-The comparison of DNA sequences from individuals in a population reveals sites which are variable. 
-If differences between individuals span only over a single nucleotide then these variations are also called Single Nucleotide Polymorphisms (SNP) and different nucleotides 
-at a specific site are defined as an allele. For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
+The availabilty of new DNA sequencing techniques has shifted the focus of biological data acuisiton towards new biomedical applications.
+Many diseases – for example Alzheimers, Parkinsons or different types of cancers – are at least partly heritable and thus the genome 
+of patients can be used for diagnostic purposes. Using the genetic information of patients is made possible through the sharp decrease in costs for analysing genetic information. 
+Thus the importance genetic information in biomedicine will/may increase over the next years and may become a standard tool for diagnostics.
+
+The comparison of DNA sequences from individuals in a population can reveal variable sites, which are of major interest in investigating diseases. 
+A variation of only one nucleotide length at a given site is called a single Nucleotide Polymorphisms (SNP). 
+The different nucleotides at this site are defined as an allele.
+For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
 Different methods have been developed to read SNPs in an individual, a
-process called genotyping. Microarrays are a cheap, fast and widespread used way to genotype individuals, in comparison
-to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
-allele frequencies can be used to associated SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
+process called genotyping. A widespread used method for genotyping individuals is the use of microarrays.
+In comparison to whole genome sequencing only sites which are known to be variable are analysed, making microarrays cheaper and faster \cite{Brown1999}.
+
+If the genotypes for more then one individual are known, the analysis of
+allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
 (GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
@@
 Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
-Those companies use DNA microarrays to screen for around 1 million SNPs spread over the human genome.
-In return, customers get an analysis of the results, as well as a file that includes the customer's raw individual genotypes.
-In 2011, 23andMe alone had over 100,000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} - the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about traits and diseases.
-With the consent of the customer the data is used for association studies.
-23andMe has published several articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}.
+Those companies use DNA microarrays to screen for around 1 million SNPs spread over the human genome. In return, customers 
+get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
+alone had over 100,000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} 
+- the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
+traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
+articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
 Over 30,000 23andme-customers have participated in those association studies.
 
-Although companies like 23andMe are willing to contribute to science, it is not easy for individual scientists to access the data.
-This is mainly due to concerns about privacy, liability and consent.
-Nevertheless, there are individual customers who are willingly sharing their data.
-Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}.
+Although companies like 23andMe are willing to use the data for scientific research,
+it is not easy for individual scientists to access the data. 
+The generation of biomedical data by private companies raises concerns about privacy \footnote{https://www.23andme.com/about/privacy/}, 
+liabilty and consent \footnote{addcitations}. 
+Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
+their personal website or to open software repositories like \textit{GitHub}. 
+This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
+the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
+most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
+like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
+or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.
 
-This data is scattered and unorganized, making it hard to use this information.
-While projects like SNPedia try to keep track of all the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is most often not attached to the genetic information.
-Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.
-
-We designed an online survey to evaluate the support in the community for a crowdsourced online plattform.
+We designed an online survey to evaluate the support in the personal genetics community for a crowdsourced online plattform.
 Here we present the results of the survey in combination with the openSNP project.
@@
 \subsection*{Survey on Sharing Genetic Information}
-229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in the survey.
-The mean age of the participants is 33 (SD = 11,29).
-81.7 \% reported their ethnicity as caucasian.
-39.7 \% of the participants are already customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future.
-29.7 \% don't plan on becoming a DTC customer.
-There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).
+In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
+The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
+customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on 
+becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).
 
-67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company didn't share the data with third parties.
-6.6 \% of the participants would not share their data.
-There is no significant difference in readiness to share between chromosomal sexes (Somers-d).
-Those who are a customers of a DTC company or are planning on becoming one in the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d).
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
+didn't share the data with third parties. 6.6 \% of the participants would not share their data. There is no significant difference in 
+readiness to share between chromosomal sexes (Somers-d). Those who are a customers of a DTC company or are planning on becoming one in 
+the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d).
 
-There are significant differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped and those who are not planning on getting genotyped.
-The first group is likely to agree more strongly, on a five-point scale, with the following motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
+There are significant differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
+and those who are not planning on getting genotyped. The first group is likely to agree more strongly, on a five-point scale, with the following 
+motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), 
+because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
 
-On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because they fear negative consequences for their family members (mean difference = 0.733, SE = 0.21, p = 0.002) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
+On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations 
+for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they feel 
+it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because they fear negative consequences for their 
+family members (mean difference = 0.733, SE = 0.21, p = 0.002) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
 
-Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with the following motivations for sharing than those who would not share their data with their DTC company: because they want to help scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
+Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
+the following motivations for sharing than those who would not share their data with their DTC company: because they want to help 
+scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), 
+and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
 
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
+for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
+fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), 
+because they fear consequences for their family members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
 
@@
 The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. 
-DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers but the data can also be used for other research and in any other context.
-\bastian{Changed this to make it more true ;)}
+DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers,
+but the data can also be used for other research and in any other context.  
+Users can create a personal profile, discuss SNPs and phenotypes on the 
+platform using a simple commenting system, or send each other private messages.
+
 \subsubsection*{Sharing genotypic information}
-Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree to/via a webinterface to the openSNP 
+Currently users can upload their genotypings from the companies 23andme, deCODEme and FamilyTreeDNA via a webinterface to the openSNP 
 project. The uploaded data is published under the Creative Commons Zero-license, 
@@
 allows a complete reuse of the data without any constraints.
-Between the start of openSNP on 09/27/2011 and 27/02/2012, 286 people have signed 
-up with openSNP, and 117 genotyping files were made available. The openSNP 
+Between the start of openSNP on 09/27/2011 and 12/18/2012, 214 people have signed 
+up with openSNP, and 79 genotyping files were made available. The openSNP 
 database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.
+Figure n depicts the increase of users and genotyping files over time.
 
 \subsubsection*{Crowdsourcing phenotypes}
-Beside uploading genotypic data, users may list new phenotypes that are not yet 
-queried by openSNP. The specification of these phenotypes  is open and not limited 
+Beside uploading genotypic data, users may create new phenotypes that are not yet 
+listed by openSNP. 
+The specification of these phenotypes is open and not limited 
 to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
 the expression and spelling of the same phenotype or variation. We implemented an 
-autocompletion-feature which helps users to reuse already entered phenotypes.
+autocompletion-feature, which helps users to reuse already entered phenotypes.
 Users are encouraged to list as many phenotypes as possible through a simple 
-achievement system which rewards users that upload their data and enter phenotypic 
+achievement system, rewarding users that upload their data and enter phenotypic 
 information with small badges that are shown on their profile pages.
@@
 In the same timeframe that was mentioned above, all users combined have 
-entered a total of 675 variations on 47 different phenotypes with those variations being the different values on a given trait or phenotype. 
-\bastian{Clarification of variation/phenotype?}
-See figure n for a 
-distribution of data acquisition over time.
+entered a total of 675 variations on 47 different phenotypes with those variations being 
+the different values on a given trait or phenotype. See figure n for the increase of phenotypic information over time.
+
 The mean number of users that have entered their variations for a single phenotype 
@@
 entered their data per phenotype can be seen in figure n+1. The phenotype provided 
-y the most users is the eye color, which has been entered by 54 users. There are 
+by the most users is the eye color, which has been entered by 54 users. There are 
 two phenotypes which have so far only been provided by a single user: 
@@
 A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
-Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2).
-For usability reasons, SNPs are ranked by the amount of information gathered by the external services.The external services themselves are ranked by how easily users can access information from these sources.
-The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science.
-Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user.
+Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, 
+SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily users can access information 
+from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of 
+the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. 
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
@@
 \subsubsection*{Data access}
-OpenSNP offers extensive access to the data uploaded by users.
-Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table.
-The genetic data is also accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes.
+OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
+grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. The genetic data is also 
+accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. 
 The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format.
-Users can discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.
 
@@
 
-\subsubsection*{Survey issues}
-Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
-As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
-\philipp{"Because of this approach", what approach?}
+\subsection*{Survey issues}
+As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, 
+the results do not reflect the general population, but overrepresent those people most likely to be interested in
+a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology. 
+As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic 
+testing felt about sharing their data with third parties, this approach was warranted.
 
-\subsubsection*{GWAS and open data}
-Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits.
-Despite this benefit in terms of cost, it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow inference on the cause for any correlation.
-Furthermore, GWAS need a large enough sample which has to be statistically analysed by using sound methods.
-Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe?
-we could look at the twitter-timeline of @opensnporg, i publish papers i find there).
+\subsection*{GWAS and open data}
+Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper 
+and offer a possibility of getting insight into genetic variations, for example SNPs which are linked to different traits 
+on a population level.
+
+Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and don't allow 
+inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
+which has to be statistically analysed by using sound methods. 
+Nevertheless, GWAS are still frequently used and new associations are found. 
+(we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).
 
 One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. 
-Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for the genotyping by themselves.
+Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high 
+numbers of customers which are willing to pay for the genotyping by themselves.
 
-By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP.
-Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, with users entering data in different measurement systems.
-While we try to suggest similar entries to the users, there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
+By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other 
+open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP. 
+Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, 
+with users entering data in different measurement systems. While we try to suggest similar entries to the users, 
+there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
 
-There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start.
-Both methods have a disadvantage: In either case it makes it harder for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered.
-Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform more quality control.
+There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one 
+could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add 
+variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder 
+for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered. 
+Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS 
+with the data to perform more quality control.
 
-Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, along with phenotypic information of medical relevance.
+Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot 
+rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing, from which an even smaller subset 
+is willing to publish the results, along with phenotypic information of medical relevance.
 
-\subsubsection*{Why should a user publish their genetic information? Pros vs. Cons}
-\philipp{Should this be a subsubsection, not rather a subsection?}
-With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS.
+With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
+data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. 
 Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
 While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
-\subsubsection*{Privacy and health implications and ethical considerations}
 
-As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.
+\subsection*{Privacy and health implications and ethical considerations}
+As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and 
+the public to help scientific progress, although people who have taken DTC genetic testing are considering the privacy implications that come 
+with openly sharing those results.
 
-The advent of DTC genetic testing has led to new ethical and social issues.
-Much of the critique on DTC genetic testing focusses on the practice of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.
-Less attention has been paid to the privacy implications that come with this cheap way of obtaining genetical information \cite{Caulfield2011,Joh2011}.
+The advent of DTC genetic testing has led to new ethical and social issues. Much of the critique on DTC genetic testing focusses on the practice 
+of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information 
+and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}. Less attention has been paid to the privacy implications 
+that come with this cheap way of obtaining genetical information \cite{Caulfield2011,Joh2011}.
 
 With a move towards making one's own genetic and medical information public, it becomes clear that the issue of privacy needs to get a closer look. 
-Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information.
-Policy makers start to react to those changes by introducing laws like the \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of widely available genetic information.
-DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
+Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
+or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
+\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
+widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
 
-We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during the upload of the DTC genetic testing results.
-Users have to confirm that they have read and understood the disclaimer about possible side-effects that can arise by publishing their data.
-To further improve this process and get informed consent, we are working on implementing the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).
+We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
+the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
+that can arise by publishing their data. To further improve this process we are working on implementing informed consent-processes,
+like the one by \textit{Consent for Research} (http://www.weconsent.us).
 
@@
 \subsection*{Survey on Sharing Genetic Information}
-The survey was performed using Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
+The survey was performed using Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, 
+blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
 
-The survey included demographics such as age, chromosomal sex and ethnicity of the participants.
-Furthermore, it included questions on their (planned) customership of a DTC company.
-If they already were customers, they were also asked if they were already sharing their genetic and phenotypic data.
-All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".
+The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
+(planned) customership of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
+All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
+but only if they did not share my medical information with anybody else" and No".
 
-The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their information with third parties.
-The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree.
-Motivations queried for sharing data were "because I want to help scientists with their research", "because of possible personal benefits (e.g.
-getting treatments for a disease I have, possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity".
-Motivations queried for not sharing data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", "because of the breach of my privacy" and " because of the fear of discrimination (e.g.
-by the employer, the state, some insurance company)".
+The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
+information with third parties. The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree. Motivations queried for sharing data 
+were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
+possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
+data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
+"because of the breach of my privacy" and " because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
 Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
@@
 \subsection*{Technical realization of the platform}
-The main platform is realized using the web framework Ruby on Rails 3.0.10.
-Postgres 9.2 is used as the main database backend for Rails.
-The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs which can be found in SNPedia.
-The literature database of Mendeley is queried using the REST API, which delivers results in JSON.
-The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format.
-Summaries on SNPs are provided by SNPedia, through querying the content via the MediaWiki API.
-All databases are queried using the unique identifier of each SNP as the search term.
+The main platform is realized using the web framework Ruby on Rails 3.0.10. Postgres 9.2 is used as the main database backend for Rails. 
+The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs 
+which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of 
+the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, 
+through querying the content via the MediaWiki API. All databases are queried using the unique identifier of each SNP as the search term.
 
-SNPs are catalogued by their unique identifier, which consists of a prefix (mostly \textit{rs}, rarely \textit{i}) and a unique number.
-This is a common format, which is employed by the NCBI dbSNP database and is also widely used and easily parsed from different literature sources.
-Publications from the different databases as well as the users' genotypes are associated with individual SNPs by the Rs-ID.
-Allele and genotype frequencies are updated regularly, based on the data present in openSNP.
+SNPs are catalogued by their unique identifier, which consists of a prefix (mostly \textit{rs}, rarely \textit{i}) and a unique number. This is a common format, 
+which is employed by the NCBI dbSNP database and is also widely used and easily parsed from different literature sources. Publications from the different databases as 
+well as the users' genotypes are associated with individual SNPs by the Rs-ID. Allele and genotype frequencies are updated regularly, based on the data present in openSNP.
 
-Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources are handled using the ruby gem Resque and the standalone server Redis.
-Search features on the platform itself are implemented using Solr and the ruby gem Sunspot.
-Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}.
-The required data is stored in a PostgreSQL database.
-Requested data is delivered in XML-format to facilitate parsing.
-Additionally, users can request data in the JSON-format, using a system not specified in any standard.
+Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources 
+are handled using the ruby gem Resque and the standalone server Redis. Search features on the platform itself are implemented using Solr and the ruby gem Sunspot. 
+Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}. The required data is stored in a PostgreSQL database.  
+Requested data is delivered in XML-format to facilitate parsing. Additionally, users can request data in the JSON-format, using a system not specified in any standard.
 
-OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented.
-Currently, two methods are implemented: \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, based on the user's query.
-The second method is \textit{sources}, which advertises all DAS-sources for all genotypes present in openSNP.
+OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
+two methods are implemented: \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
+based on the user's query. The second method is \textit{sources}, which advertises all DAS-sources for all genotypes present in openSNP.
 
-A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure n+3.
-The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr.
-The genetical and phenotypical data is licensed under Creative Commons Zero.
+A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure n+3. The source code of openSNP is 
+published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
 % Do NOT remove this, even if you are not including acknowledgments

### 2012-03-04 paper_draft.tex
added citations

@@
 The generation of biomedical data by private companies raises concerns about privacy \footnote{https://www.23andme.com/about/privacy/}, 
-liabilty and consent \footnote{addcitations}. 
+liabilty and consent \cite{Caulfield2011}. 
 Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
@@
 customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on 
-becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).
+becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).
 
 67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
-didn't share the data with third parties. 6.6 \% of the participants would not share their data. There is no significant difference in 
-readiness to share between chromosomal sexes (Somers-d). Those who are a customers of a DTC company or are planning on becoming one in 
-the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d).
+didn't share the data with third parties.
+6.6 \% of the participants would not share their data.
+Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221).
+Those who are a customers of a DTC company or are planning on becoming one in
+the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d = 0.331).
 
-There are significant differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
+There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
 and those who are not planning on getting genotyped. The first group is likely to agree more strongly, on a five-point scale, with the following 
-motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), 
-because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).
+motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128), 
+because they think of personal benefits (mean difference = 0.448, SE = 0.183) and because they are curious (mean difference = 1.159, SE = 0.193).
 
 On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations 
-for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they feel 
-it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because they fear negative consequences for their 
-family members (mean difference = 0.733, SE = 0.21, p = 0.002) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).
+for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195), because they feel 
+it is a breach of their privacy (mean difference = 0.821, SE = 0.225), because they fear negative consequences for their 
+family members (mean difference = 0.733, SE = 0.21) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208).
 
@@
 the following motivations for sharing than those who would not share their data with their DTC company: because they want to help 
-scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), 
-and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).
+scientists (mean difference = 1.57, SE = 0.199), because they think of personal benefits (mean difference = 0.951, SE = 0.308), 
+and because they are curious (mean difference = 1.99, SE = 0.321).
+In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).
 
@@
 for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
-fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), 
-because they fear consequences for their family members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).
+fear discrimination (mean difference = 1.52, SE = 0.322), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324), 
+because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357).
+In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist: Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346).
+Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).
 
@@
 which has to be statistically analysed by using sound methods. 
-Nevertheless, GWAS are still frequently used and new associations are found. 
-(we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).
+Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
 

### 2012-03-20 paper_draft.tex
added images, wrote abstract

@@
 \section*{Abstract}
-Genome-wide association studies are a cheap and quick way of assessing health risks by comparing Single Nucleotide Polymorphisms 
-between groups of participants. Direct-To-Consumer (DTC) Companies like 23andme offer SNP sequencing to their customers alongside 
-with an evaluation of the customer's genetic risks. However, the data 23andme and other companies generate is not accessible 
-to other scientists, and withholds some information from customers for various reasons. In this paper, we present an open 
-approach to Genome-wide association studies by introducing openSNP, a web platform which allows customers to openly share 
-their polymorphisms with scientists for free. We also present results of a survey to assess the willingness of DTC-customers 
-to openly share their genotyping and polymorphisms.
+Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants like Single Nucleotide Polymorphisms.
+The advent of Direct-To-Consumer genetic testing offers the analysis of Single Nucleotide Polymorphisms for an individual at a low cost without the need to consult a physician or geneticist.
+Genome-wide association studies compare the genetic variation between two groups to find correlations between membership in a group and the genetic variation.
+To find significant correlations it is necessary to have a large enough sample size.
+Over 100.000 people already got themselves genotyped through Direct-To-Consumer genetic testing and could be a valuable data source for association studies.
+However, generally this data is not public for various reasons and thus can’t be used by other scientists.
+Here we present a survey which evaluates if people are willing to publicly share their genetic information.
+In the light of those results we also present the web-platform openSNP.
+It is an open approach which allows participants of Direct-To-Consumer genetic testing to publish their genetic data along with phenotypic information into the public domain.
+Through this crowdsourced effort of collecting genetic and phenotypic information we create a valuable resource which can be used in a wide array of studies, included Genome-wide association studies.
 
@@
 \section*{Author Summary}
+Missing for now.
 
@@
 database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.
-Figure n depicts the increase of users and genotyping files over time.
+Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.
 
@@
 entered a total of 675 variations on 47 different phenotypes with those variations being 
-the different values on a given trait or phenotype. See figure n for the increase of phenotypic information over time.
+the different values on a given trait or phenotype. See figure \ref{Figure1_label} for the increase of phenotypic information over time.
 
@@
 is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
-entered their data per phenotype can be seen in figure n+1. The phenotype provided 
+entered their data per phenotype can be seen in figure \ref{Figure2_label}. The phenotype provided 
 by the most users is the eye color, which has been entered by 54 users. There are 
@@
 A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
-Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, 
+Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure \ref{Figure3_label}). For usability reasons, 
 SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily users can access information 
@@
 
-A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure n+3. The source code of openSNP is 
+A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
 published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
@@
 \section*{Figure Legends}
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[scale=0.35]{chart_growth.png}
+	\end{center}
+	\caption{
+	{\bf Growth of openSNP.} The increase in numbers for users, genotyping-files, phenotypes and their variation from 27.09.2011 to 16.12.2011 is shown.} 
+	\label{Figure1_label}
+\end{figure}
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[scale=0.40]{histogram_phenotypes.png}
+	\end{center}
+	\caption{
+	{\bf Histogram of users/phenotype-distribution.} The x-axis shows the minimum number of users who provide information for a phenotype, the y-axis shows how many phenotypes have at least that many users.}
+	\label{Figure2_label}
+\end{figure}
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[scale=0.50]{paper_distribution.png}
+	\end{center}
+	\caption{
+	{\bf Distribution of external information gathered for SNPs in the openSNP-database.} Data on PLoS and SNPedia is openly available for every user.
+Publications on Mendeley are either Open Access (OA) or Closed Access (CA).}
+	\label{Figure3_label}
+\end{figure}
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[scale=0.60]{uml_diagram.png}
+	\end{center}
+	\caption{
+	{\bf Flow of data inside openSNP.} External databases and user-provided data are used as input. Output of data is done using the website, the \emph{Distributed Annotation System} and a JSON-API.} 
+	\label{Figure4_label}
+\end{figure}
+
 %\begin{figure}[!ht]

### 2012-03-25 paper_draft.tex
Replaced all don'ts etc by do not etc., fixed typos, started work on more coherent abstract, changed DAS-citation to earlier in text, fixed all of Manuel's small tips

@@
 }
+% Alternative titles: 
+% openSNP - a new, open data-source for personalised medicine
+% What kind of person would share genotyping-data? Presenting a survey and an open data-source for personalised medicine
+% 
 % Insert Author names, affiliations and corresponding author email.
@@
 \\
-\bf{2} Bond University, Gold Coast, Australia
+\bf{2} University of Queensland, Brisbane, Australia
 \\
@@
 To find significant correlations it is necessary to have a large enough sample size.
-Over 100.000 people already got themselves genotyped through Direct-To-Consumer genetic testing and could be a valuable data source for association studies.
-However, generally this data is not public for various reasons and thus can’t be used by other scientists.
+Over 100.000 people were already genotyped through Direct-To-Consumer genetic testing and could be a valuable data source for association studies.
+However, generally this data is not public for various reasons and thus cannot be used by other scientists.
+It is vital that there is a central open data-repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
 Here we present a survey which evaluates if people are willing to publicly share their genetic information.
-In the light of those results we also present the web-platform openSNP.
-It is an open approach which allows participants of Direct-To-Consumer genetic testing to publish their genetic data along with phenotypic information into the public domain.
+In the light of those results we also present the web-platform openSNP, which is a open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data along with phenotypic information into the public domain.
 Through this crowdsourced effort of collecting genetic and phenotypic information we create a valuable resource which can be used in a wide array of studies, included Genome-wide association studies.
@@
 
-The availabilty of new DNA sequencing techniques has shifted the focus of biological data acuisiton towards new biomedical applications.
+The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
 Many diseases – for example Alzheimers, Parkinsons or different types of cancers – are at least partly heritable and thus the genome 
@@
 
-The comparison of DNA sequences from individuals in a population can reveal variable sites, which are of major interest in investigating diseases. 
-A variation of only one nucleotide length at a given site is called a single Nucleotide Polymorphisms (SNP). 
-The different nucleotides at this site are defined as an allele.
-For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
-Different methods have been developed to read SNPs in an individual, a
-process called genotyping. A widespread used method for genotyping individuals is the use of microarrays.
-In comparison to whole genome sequencing only sites which are known to be variable are analysed, making microarrays cheaper and faster \cite{Brown1999}.
+%Kicked out the following due to Corpas' argument
+%The comparison of DNA sequences from individuals in a population can reveal variable sites, which are of major interest in investigating diseases. 
+%A variation of only one nucleotide length at a given site is called a single Nucleotide Polymorphisms (SNP). 
+%The different nucleotides at this site are defined as an allele.
+%For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
+%Different methods have been developed to read SNPs in an individual, a
+%process called genotyping. A widespread used method for genotyping individuals is the use of microarrays.
+%In comparison to whole genome sequencing only sites which are known to be variable are analysed, making microarrays cheaper and faster \cite{Brown1999}.
 
@@
 most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
-like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
+like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
 or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.
 
-We designed an online survey to evaluate the support in the personal genetics community for a crowdsourced online plattform.
+We designed an online survey to evaluate the support in the personal genetics community for a crowdsourced online platform.
 Here we present the results of the survey in combination with the openSNP project.
 The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources. 
-The data is made available to research via a Creative-Commones-Zero licence.
+The data is made available to research via a Creative-Commons-Zero licence.
 
@@
 The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
-customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on 
+customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
 becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).
@@
 Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221).
-Those who are a customers of a DTC company or are planning on becoming one in
-the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d = 0.331).
+Those who are customers of a DTC company or are planning on becoming one in
+the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331).
 
@@
 from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of 
-the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. 
+the Public Library of Science. Lowest values are given to the Mendeley results, as those are not necessary freely available to every user. 
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
@@
 grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. The genetic data is also 
-accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. 
+accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
 The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format.
@@
 
-Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and don't allow 
+Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and do not allow 
 inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
@@
 with users entering data in different measurement systems. While we try to suggest similar entries to the users, 
-there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
+there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
 
 There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one 
-could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add 
+could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
 variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder 
@@
 Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
-While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
+While we do not have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
 
-\subsection*{Privacy and health implications and ethical considerations}
+\subsection*{Privacy, health implications and ethical considerations}
 As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and 
@@
 
-The advent of DTC genetic testing has led to new ethical and social issues. Much of the critique on DTC genetic testing focusses on the practice 
+%The advent of DTC genetic testing has led to new ethical and social issues. 
+Much of the critique on DTC genetic testing focusses on the practice 
 of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information 
@@
 
-\subsection*{Technical realization of the platform}
-The main platform is realized using the web framework Ruby on Rails 3.0.10. Postgres 9.2 is used as the main database backend for Rails. 
+\subsection*{Technical implementation of the platform}
+The main platform is implemented using the web framework Ruby on Rails 3.0.10. Postgres 9.2 is used as the main database backend for Rails. 
 The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs 
@@
 are handled using the ruby gem Resque and the standalone server Redis. Search features on the platform itself are implemented using Solr and the ruby gem Sunspot. 
-Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}. The required data is stored in a PostgreSQL database.  
+Additionally, data can be requested from openSNP using the Distributed Annotation System. The required data is stored in a PostgreSQL database.  
 Requested data is delivered in XML-format to facilitate parsing. Additionally, users can request data in the JSON-format, using a system not specified in any standard.

### 2012-03-25 paper_draft.tex
Added v2-chip numbers. Removed space in superscript page 2 paragraph 3 (we have to move footnotes to references anyway) Added dbSNP-citation to papers.bib and text

@@
 Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
-Those companies use DNA microarrays to screen for around 1 million SNPs spread over the human genome. In return, customers 
+These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
 get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
@@
 it is not easy for individual scientists to access the data. 
-The generation of biomedical data by private companies raises concerns about privacy \footnote{https://www.23andme.com/about/privacy/}, 
+The generation of biomedical data by private companies raises concerns about privacy\footnote{https://www.23andme.com/about/privacy/}, 
 liabilty and consent \cite{Caulfield2011}. 
@@
 SNPs are catalogued by their unique identifier, which consists of a prefix (mostly \textit{rs}, rarely \textit{i}) and a unique number. This is a common format, 
-which is employed by the NCBI dbSNP database and is also widely used and easily parsed from different literature sources. Publications from the different databases as 
+which is employed by the NCBI dbSNP database \cite{Sherry2001} and is also widely used and easily parsed from different literature sources. Publications from the different databases as 
 well as the users' genotypes are associated with individual SNPs by the Rs-ID. Allele and genotype frequencies are updated regularly, based on the data present in openSNP.

### 2012-03-25 paper_draft.tex
Fixed colons and lowercases. Put footnotes into references according to http://blogs.plos.org/mfenner/2011/07/21/how-to-formally-cite-a-blog-post/

@@
 to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
-such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
+such studies has been rising. Over 1200 GWAS have been performed \cite{Johnson2009} and over 
 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
@@
 get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
-alone had over 100,000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} 
+alone had over 100,000 customers \cite{23andMe2011}
 - the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
@@
 it is not easy for individual scientists to access the data. 
-The generation of biomedical data by private companies raises concerns about privacy\footnote{https://www.23andme.com/about/privacy/}, 
+The generation of biomedical data by private companies raises concerns about privacy \cite{23andMe2012}, 
 liabilty and consent \cite{Caulfield2011}. 
@@
 because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357).
-In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist: Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346).
+In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist.
+Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346).
 Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).
@@
 two phenotypes which have so far only been provided by a single user: 
-The SAT Writing score and triglyceride-levels.
+the SAT Writing score and triglyceride-levels.
 
@@
 
-There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one 
+There are two possible solutions to this problem: one could only allow some (trusted) users to enter new phenotypes or one 
 could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 

### 2012-07-17 paper_draft.tex
abstract redone

@@
 \section*{Abstract}
-Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants like Single Nucleotide Polymorphisms.
-The advent of Direct-To-Consumer genetic testing offers the analysis of Single Nucleotide Polymorphisms for an individual at a low cost without the need to consult a physician or geneticist.
+Genome-wide association studies are used to correlate phenotypic traits with genetic variants like Single Nucleotide Polymorphisms (SNPs).
 Genome-wide association studies compare the genetic variation between two groups to find correlations between membership in a group and the genetic variation.
 To find significant correlations it is necessary to have a large enough sample size.
-Over 100.000 people were already genotyped through Direct-To-Consumer genetic testing and could be a valuable data source for association studies.
-However, generally this data is not public for various reasons and thus cannot be used by other scientists.
-It is vital that there is a central open data-repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
-Here we present a survey which evaluates if people are willing to publicly share their genetic information.
-In the light of those results we also present the web-platform openSNP, which is a open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data along with phenotypic information into the public domain.
-Through this crowdsourced effort of collecting genetic and phenotypic information we create a valuable resource which can be used in a wide array of studies, included Genome-wide association studies.
+The advent of Direct-To-Consumer genetic testing offers the analysis of SNPs for an individual at low cost without the need to consult a physician or geneticist.
+Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing.
+This data could be a valuable source for association studies.
+However, this data is not public for various reasons and thus cannot be used by other scientists.
+It seems reasonable to create a central open data-repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
+Here we present a survey which evaluates whether people are willing to publicly share their genetic information.
+In the light of those results we also present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data along with phenotypic information into the public domain.
+Through this "crowdsourced" effort of collecting genetic and phenotypic information we create a valuable resource which can be used in a wide array of studies, included Genome-wide association studies.
+OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at github.com/gedankenstuecke/snpr.
 
@@
 The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases – for example Alzheimers, Parkinsons or different types of cancers – are at least partly heritable and thus the genome 
+Many diseases – for example Alzheimers, Parkinsons or different types of cancers – are at least partially heritable so the genome 
 of patients can be used for diagnostic purposes. Using the genetic information of patients is made possible through the sharp decrease in costs for analysing genetic information.
-Thus the importance genetic information in biomedicine will/may increase over the next years and may become a standard tool for diagnostics.
 
@@
 
-If the genotypes for more then one individual are known, the analysis of
+If the genotypes for more than one individual are known, the analysis of
 allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies

### 2012-07-17 paper_draft.tex
Ironed out sentences, shortened a lot, re-arranged some stuff

@@
 to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
-such studies has been rising. Over 1200 GWAS have been performed \cite{Johnson2009} and over 
-5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
+such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
+5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}. 
 %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
 
-Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
+GWAS are not limited to the scientific community. Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
 These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
-get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
+receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
 alone had over 100,000 customers \cite{23andMe2011}
-- the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
+- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
 traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
 articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
-Over 30,000 23andme-customers have participated in those association studies.
+So far, over 30,000 23andme-customers have participated in those association studies, which proves that this data-source has potential for other researchers.
 
-Although companies like 23andMe are willing to use the data for scientific research,
-it is not easy for individual scientists to access the data. 
 The generation of biomedical data by private companies raises concerns about privacy \cite{23andMe2012}, 
-liabilty and consent \cite{Caulfield2011}. 
-Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
+liability and consent \cite{Caulfield2011}. 
+Nevertheless, in some instances individual customers are willingly sharing their data. Most do so by uploading their data to
 their personal website or to open software repositories like \textit{GitHub}. 
-This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
-the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
-most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
+This data is scattered and unorganized, making it hard to use. While projects like SNPedia try to keep track of all 
+the publicly available genotyping files \cite{Cariaso2011}, they usually do not provide the information necessary to perform GWAS, as the phenotypic information is 
+often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
 like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
-or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.
+or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.
 
-We designed an online survey to evaluate the support in the personal genetics community for a crowdsourced online platform.
-Here we present the results of the survey in combination with the openSNP project.
-The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources. 
-The data is made available to research via a Creative-Commons-Zero licence.
+Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform. 
+We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes. 
+The genotypes are made available to researchers via the open Creative-Commons-Zero licence.
 
@@
 
-The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. 
-DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers,
-but the data can also be used for other research and in any other context.  
+The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers interested in performing GWAS. 
+DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable access for researchers.
 Users can create a personal profile, discuss SNPs and phenotypes on the 
@@
 Currently users can upload their genotypings from the companies 23andme, deCODEme and FamilyTreeDNA via a webinterface to the openSNP 
-project. The uploaded data is published under the Creative Commons Zero-license, 
+project. There is experimental support for openSNP-exomes. The uploaded data is published under the Creative Commons Zero-license, 
 which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
@@
 up with openSNP, and 79 genotyping files were made available. The openSNP 
-database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.
+database lists 69,486,471 SNPs of which 1,938,603 are unique SNPs.
 Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.
@@
 \subsubsection*{Crowdsourcing phenotypes}
-Beside uploading genotypic data, users may create new phenotypes that are not yet 
+Users are able to create new phenotypes that are not yet 
 listed by openSNP. 
@@
 
-In the same timeframe that was mentioned above, all users combined have 
+In the same timeframe as above, all users combined have 
 entered a total of 675 variations on 47 different phenotypes with those variations being 
@@
 A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
-Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure \ref{Figure3_label}). For usability reasons, 
-SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily users can access information 
-from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of 
-the Public Library of Science. Lowest values are given to the Mendeley results, as those are not necessary freely available to every user. 
+Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}). For usability reasons, 
+SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily non-scientists can understand information 
+from these sources. The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of 
+the Public Library of Science. Lowest values are given to the Mendeley results, as those are not necessarily freely available to every user. 
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
@@
 As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, 
-the results do not reflect the general population, but overrepresent those people most likely to be interested in
+the results do not reflect the general population, but over-represent those people most likely to be interested in
 a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
-As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic 
-testing felt about sharing their data with third parties, this approach was warranted.
+
+\subsection*{Privacy, health implications and ethical considerations}
+
+%The advent of DTC genetic testing has led to new ethical and social issues. 
+Much of the critique on DTC genetic testing focusses on the practice 
+of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
+and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.
+
+As we have found in the survey on sharing such results, many DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
+There is a variety of ethical and privacy implications when it comes to DTC genetic testing\cite{Caulfield2011,Joh2011}.
+
+Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
+or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
+\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
+widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
+
+OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
+the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
+that can arise by publishing their data. To further improve this process we are working on implementing an informed consent-processes,
+similar to \textit{Consent for Research} (http://www.weconsent.us).
 
 \subsection*{GWAS and open data}
-Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper 
-and offer a possibility of getting insight into genetic variations, for example SNPs which are linked to different traits 
-on a population level.
-
-Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and do not allow 
-inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
-which has to be statistically analysed by using sound methods. 
+Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper.
+However, GWAS can only detect correlations of SNPs with those traits and do not allow
+inference on the cause for any correlation.
+Furthermore, for a statistically sound analysis GWAS need a large enough sample size.
 Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
@@
 Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, 
-with users entering data in different measurement systems. While we try to suggest similar entries to the users, 
+with users entering data in different measurement systems.
+Another problem with user-entered data is the frequent switching between categorical and continuous phenotypes of users - for example, some users entered the specific value of their height, while other users entered their height according to a category like "150cm to 160cm".
+
+While we try to suggest similar entries to the users, 
 there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
+There are two possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes.
+The other one
+would be to make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
+variations that have not been available from the start.
 
-There are two possible solutions to this problem: one could only allow some (trusted) users to enter new phenotypes or one 
-could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
-variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder 
-for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered. 
-Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS 
-with the data to perform more quality control.
+In both cases it makes it harder 
+for users to enter their data which raises the bar for participation.
+We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform more quality control.
 
-Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot 
-rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing, from which an even smaller subset 
-is willing to publish the results, along with phenotypic information of medical relevance.
+Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP: only a subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, which can potentially lead to skewed GWAS-results.
 
 With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
-data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. 
+data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to use in GWAS. 
 Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
-While we do not have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.
-
-\subsection*{Privacy, health implications and ethical considerations}
-As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and 
-the public to help scientific progress, although people who have taken DTC genetic testing are considering the privacy implications that come 
-with openly sharing those results.
-
-%The advent of DTC genetic testing has led to new ethical and social issues. 
-Much of the critique on DTC genetic testing focusses on the practice 
-of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information 
-and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}. Less attention has been paid to the privacy implications 
-that come with this cheap way of obtaining genetical information \cite{Caulfield2011,Joh2011}.
-
-With a move towards making one's own genetic and medical information public, it becomes clear that the issue of privacy needs to get a closer look. 
-Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
-or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
-\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
-widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
-
-We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
-the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-that can arise by publishing their data. To further improve this process we are working on implementing informed consent-processes,
-like the one by \textit{Consent for Research} (http://www.weconsent.us).
+While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise.
 
@@
 The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
-(planned) customership of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
+(planned) customer-ship of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
 All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
@@
 The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
-information with third parties. The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree. Motivations queried for sharing data 
+information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
 were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
@@
 A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
-published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
+published under the MIT-license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
 % Do NOT remove this, even if you are not including acknowledgments

### 2012-07-17 paper_draft.tex
I'm trapped in GitHub, help, I have to live off too-long sentences

@@
 traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
-articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
-So far, over 30,000 23andme-customers have participated in those association studies, which proves that this data-source has potential for other researchers.
+articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
+So far, over 30,000 23andme-customers have participated in 23andme's association studies, which proves that this data-source has a lot of potential for other researchers.
 
@@
 
-OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
+OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, once during registration for openSNP and once during 
 the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-that can arise by publishing their data. To further improve this process we are working on implementing an informed consent-processes,
+of publishing their data. To further improve this process we are working on implementing an informed consent-processes 
 similar to \textit{Consent for Research} (http://www.weconsent.us).
@@
 With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
-data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to use in GWAS. 
+data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to freely use in their studies.
 Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
@@
 \section*{Acknowledgments}
+We thank Dr. Manuel Corpas for constructive advice in grammar, spelling and structure of this study.
 

### 2012-07-17 paper_draft.tex
Minimal mistake fixed, stopping for today

@@
 with users entering data in different measurement systems.
-Another problem with user-entered data is the frequent switching between categorical and continuous phenotypes of users - for example, some users entered the specific value of their height, while other users entered their height according to a category like "150cm to 160cm".
+Another problem with user-entered data is the frequent switching between categorical and continuous phenotypes - for example, some users entered the specific value of their height, while other users entered their height according to a category like "150cm to 160cm".
 

### 2012-07-18 paper_draft.tex
Introduced tables for survey-results. Made transition survey/openSNP easier.

@@
 There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
-and those who are not planning on getting genotyped. The first group is likely to agree more strongly, on a five-point scale, with the following 
-motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128), 
-because they think of personal benefits (mean difference = 0.448, SE = 0.183) and because they are curious (mean difference = 1.159, SE = 0.193).
+and those who are not planning on getting genotyped.
+The first group is likely to agree more strongly, on a five-point scale, with motivations for sharing genotypic information.
+On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations
+for not sharing their data. For easier comparison both of these groups are compared in table \ref{tab:motivations1}.
 
-On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations 
-for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195), because they feel 
-it is a breach of their privacy (mean difference = 0.821, SE = 0.225), because they fear negative consequences for their 
-family members (mean difference = 0.733, SE = 0.21) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208).
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who are already genotyped} & & \\ \hline 
+... curious & 1.159 & 0.193 \\ 
+... want to help scientists & 0.465 & 0.128 \\
+... for personal benefits & 0.448 & 0.183 \\ \hline
+\textbf{Motivation for not sharing in participants} & & \\ 
+\textbf{who are not planning to get genotyped} & & \\ \hline
+... fear of discrimination & 1.06 & 0.195 \\
+... breach of privacy & 0.821 & 0.225 \\
+... fear of personalized advertising & 0.848 & 0.208 \\ 
+... negative consequences for family members & 0.733 & 0.21 \\ \hline 
+\end{tabular}
+\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
+}
+\label{tab:motivations1}
+\end{table}
+
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would share with their DTC provider} & & \\ \hline
+... curiosity & 1.99 & 0.321 \\
+... want to help science & 1.57 & 0.199 \\
+... for personal benefits & 0.951 & 0.308 \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would not share with their DTC provider} & & \\ \hline
+... fear of discrimination & 1.52 & 0.322 \\
+... fear of consequences for family members & 1.146 & 0.32 \\
+... fear of personalized advertising & 1.112 & 0.357 \\ \hline
+\end{tabular}
+\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with their DTC provider with participants who would not share their data with their DTC provider.}
+\label{tab:motivations2}
+\end{table}
 
 Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
-the following motivations for sharing than those who would not share their data with their DTC company: because they want to help 
-scientists (mean difference = 1.57, SE = 0.199), because they think of personal benefits (mean difference = 0.951, SE = 0.308), 
-and because they are curious (mean difference = 1.99, SE = 0.321).
+the following motivations for sharing than those who would not share their data with their DTC company.
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the some motivations 
+for not sharing their data when compared to those who would share their data with their DTC company under any circumstances, for an overview of the motivations of both groups, see table \ref{tab:motivations2}.
+
 In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).
 
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
-for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
-fear discrimination (mean difference = 1.52, SE = 0.322), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324), 
-because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357).
 In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist.
@@
 Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).
+
+These survey-results indicate that there is a definite interest in customers of DTC-companies to share their results with other scientists.
+We've created the web-platform openSNP to help DTC-customers share their genotyping results with scientists, for free and under the open Creative Commons Zero-license.
 

### 2012-08-22 paper_draft.tex
restructure of paper itself

--- 2012-07-17 paper_draft.tex
@@
 The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases – for example Alzheimers, Parkinsons or different types of cancers – are at least partially heritable so the genome 
-of patients can be used for diagnostic purposes. Using the genetic information of patients is made possible through the sharp decrease in costs for analysing genetic information.
+Many diseases – for example Alzheimers, Parkinsons or different types of cancers (needs citations) – are at least partially heritable so the genome 
+of patients can be used for diagnostic purposes.
+Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.
 
@@
 
-If the genotypes for more than one individual are known, the analysis of
-allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
-(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
-enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
+If genetic information for more than one individual is known, the analysis of
+allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
+(GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
+enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or disease.
 This method does not allow inference of causal differences but merely identifies correlations. 
@@
 5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}.
-%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
 
-GWAS are not limited to the scientific community. Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
+GWAS are not only performed inside the traditional scientific community. 
+Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
 These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
@@
 their personal website or to open software repositories like \textit{GitHub}. 
-This data is scattered and unorganized, making it hard to use. While projects like SNPedia try to keep track of all 
+This data is scattered and unorganized, making it hard to use in studies. While projects like SNPedia try to keep track of all 
 the publicly available genotyping files \cite{Cariaso2011}, they usually do not provide the information necessary to perform GWAS, as the phenotypic information is 
 often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
-like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
+like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an application programming interface (API) 
 or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.
 
-Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform. 
-We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes. 
-The genotypes are made available to researchers via the open Creative-Commons-Zero licence.
+As a follow up of a survey, which evaluated whether customers of DTC genetic testing would be willing to share their genotypic and phenotypic information (see Supplementary Materials), we designed openSNP, an online platform which enables DTC customers to freely share their genotypic and phenotypic data under a Creative Commons-Zero license.
+The phenotypic and genotypic data can be retrieved from openSNP using different methods, which also include different API standards to allow automated processing of the data.
+Additionally openSNP also enables users of DTC genetic testing to receive further information on their genotypes, through the crowdsourced resource SNPedia and primary publications, like those of the Public Library of Science journals.
 
@@
 
-\subsection*{Survey on Sharing Genetic Information}
-In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
-The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
-customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
-becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).
+We created the openSNP project (http://opensnp.org) as an open, crowdsourced online platform for DTC customers interested in sharing their raw data and for researchers interested in performing GWAS or perform other types of analysis with the data.
+Customers of DTC testing are encouraged to share their genotyping results along with their phenotypic traits to enable easy access for researchers.
+Users of openSNP can create a personal profile, discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.
 
-67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
-didn't share the data with third parties.
-6.6 \% of the participants would not share their data.
-Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221).
-Those who are customers of a DTC company or are planning on becoming one in
-the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331).
+People interested in using the data of openSNP can download complete dumps of the genotypic and phenotypic information or use query API endpoints utilizing JavaScript Object Notation (JSON) objects or the Distributed Annotation System (DAS) \cite{Dowell2001}.
 
-There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
-and those who are not planning on getting genotyped. The first group is likely to agree more strongly, on a five-point scale, with the following 
-motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128), 
-because they think of personal benefits (mean difference = 0.448, SE = 0.183) and because they are curious (mean difference = 1.159, SE = 0.193).
-
-On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations 
-for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195), because they feel 
-it is a breach of their privacy (mean difference = 0.821, SE = 0.225), because they fear negative consequences for their 
-family members (mean difference = 0.733, SE = 0.21) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208).
-
-Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
-the following motivations for sharing than those who would not share their data with their DTC company: because they want to help 
-scientists (mean difference = 1.57, SE = 0.199), because they think of personal benefits (mean difference = 0.951, SE = 0.308), 
-and because they are curious (mean difference = 1.99, SE = 0.321).
-In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).
-
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
-for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
-fear discrimination (mean difference = 1.52, SE = 0.322), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324), 
-because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357).
-In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist.
-Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346).
-Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).
-
-\subsection*{openSNP}
-
-The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers interested in performing GWAS. 
-DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable access for researchers.
-Users can create a personal profile, discuss SNPs and phenotypes on the 
-platform using a simple commenting system, or send each other private messages.
-
-\subsubsection*{Sharing genotypic information}
-Currently users can upload their genotypings from the companies 23andme, deCODEme and FamilyTreeDNA via a webinterface to the openSNP 
-project. There is experimental support for openSNP-exomes. The uploaded data is published under the Creative Commons Zero-license, 
-which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
+\subsection*{Sharing genotypic information}
+Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a webinterface to the openSNP 
+project.
+There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers.
+So far only the SNPs of the exome data sets are visualized on openSNP, but the downloads include all variation found in the exome.
+The uploaded data is published under the Creative Commons Zero-license, 
+which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
 allows a complete reuse of the data without any constraints.
@@
 up with openSNP, and 79 genotyping files were made available. The openSNP 
-database lists 69,486,471 SNPs of which 1,938,603 are unique SNPs.
-Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.
+database lists 69,486,471 genotypes which are distributed over 1,938,603 unique SNPs.
+Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.\bastian{update all numbers}
 
-\subsubsection*{Crowdsourcing phenotypes}
+\subsection*{Crowdsourcing phenotypes}
 Users are able to create new phenotypes that are not yet 
@@
 two phenotypes which have so far only been provided by a single user: 
-the SAT Writing score and triglyceride-levels.
+the SAT Writing score and triglyceride-levels.\bastian{update all numbers and graphs}
 
-\subsubsection*{Connection to external services}
+\subsection*{Connection to external services}
 In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
-A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
+A total number of 15,229 documents \bastian{number needs to get updated}relevant to the SNPs listed in openSNP could be found in the publication databases of Mendeley, the Public Library of Science and in the crowdsourced SNPedia.
 Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}). For usability reasons, 
-SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily non-scientists can understand information 
-from these sources. The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of 
-the Public Library of Science. Lowest values are given to the Mendeley results, as those are not necessarily freely available to every user. 
+SNPs are ranked by the amount of information gathered through the external services. The external services themselves are ranked by how easily non-scientists can understand information 
+from these sources and available this information.
+The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of
+the Public Library of Science. Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments. 
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
 
-\subsubsection*{Data access}
+\subsection*{Data access}
 OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
-grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. The genetic data is also 
+grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
 accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
+An example of how the DAS can be used can be found on openSNP where users genotypes are visualized inside a genome browser.
+So far all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now.
+\bastian{should this maybe include an image as well?}
+
 The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format.
-
+The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.
 \section*{Discussion}
-
-\subsection*{Survey issues}
-As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, 
-the results do not reflect the general population, but over-represent those people most likely to be interested in
-a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
 
@@
 
-As we have found in the survey on sharing such results, many DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
+As we have found in our survey on sharing such results (see supplementary methods), many DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
 There is a variety of ethical and privacy implications when it comes to DTC genetic testing\cite{Caulfield2011,Joh2011}.
@@
 the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-of publishing their data. To further improve this process we are working on implementing an informed consent-processes 
-similar to \textit{Consent for Research} (http://www.weconsent.us).
+of publishing their data. To further improve this process we are looking forward to implement an informed consent-processes 
+by a provider like \textit{Consent for Research} (http://www.weconsent.us).
 
-\subsection*{GWAS and open data}
+\subsection*{GWAS and Open Data}
 Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper.
@@
 for users to enter their data which raises the bar for participation.
-We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform more quality control.
+We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform additional quality control.
 
@@
 While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise.
+By including the option of uploading exome data sets the platform already is capable of adjusting for changes in the type of data generated by DTC genetic testing.
 
@@
 \section*{Materials and Methods}
-\subsection*{Survey on Sharing Genetic Information}
-The survey was performed using Google Docs and was distributed to possible participants through the 23andMe community, the DIYBiology mailing list, 
-blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook.
-
-The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
-(planned) customer-ship of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
-All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
-but only if they did not share my medical information with anybody else" and No".
-
-The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
-information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
-were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
-possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
-data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
-"because of the breach of my privacy" and " because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
-Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
-
-The survey data was analyzed with SPSS 19.
 

### 2012-08-28 paper_draft.tex
Reworked abstract a little

--- 2012-07-18 paper_draft.tex
@@
 \section*{Abstract}
-Genome-wide association studies are used to correlate phenotypic traits with genetic variants like Single Nucleotide Polymorphisms (SNPs).
-Genome-wide association studies compare the genetic variation between two groups to find correlations between membership in a group and the genetic variation.
-To find significant correlations it is necessary to have a large enough sample size.
+Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants, for example Single Nucleotide Polymorphisms (SNPs).
+These studies compare the genetic variation between two groups to single out certain SNPs that are linked to a phenotypic variation in one of the groups.
+However, it is necessary to have a large enough sample size to find statistically significant correlations.
 The advent of Direct-To-Consumer genetic testing offers the analysis of SNPs for an individual at low cost without the need to consult a physician or geneticist.
-Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing.
+Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies.
 This data could be a valuable source for association studies.
-However, this data is not public for various reasons and thus cannot be used by other scientists.
+However, this data is not public for a variety of reasons and thus cannot be used in research.
 It seems reasonable to create a central open data-repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
 Here we present a survey which evaluates whether people are willing to publicly share their genetic information.
-In the light of those results we also present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data along with phenotypic information into the public domain.
-Through this "crowdsourced" effort of collecting genetic and phenotypic information we create a valuable resource which can be used in a wide array of studies, included Genome-wide association studies.
-OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at github.com/gedankenstuecke/snpr.
+In the light of those results we present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publicly publish at no cost their genetic data along with phenotypic information.
+Through this "crowdsourced" effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide array of studies, including Genome-wide association studies.
+OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at github.com/gedankenstuecke/snpr
 

### 2012-08-28 paper_draft.tex
Now the abstract flows better

@@
 \section*{Abstract}
-Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants, for example Single Nucleotide Polymorphisms (SNPs).
-These studies compare the genetic variation between two groups to single out certain SNPs that are linked to a phenotypic variation in one of the groups.
+Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants.
+These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups.
 However, it is necessary to have a large enough sample size to find statistically significant correlations.
-The advent of Direct-To-Consumer genetic testing offers the analysis of SNPs for an individual at low cost without the need to consult a physician or geneticist.
+Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist.
 Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies.
-This data could be a valuable source for association studies.
 However, this data is not public for a variety of reasons and thus cannot be used in research.
@@
 In the light of those results we present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publicly publish at no cost their genetic data along with phenotypic information.
-Through this "crowdsourced" effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide array of studies, including Genome-wide association studies.
+Through this "crowdsourced" effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide area of studies, including Genome-wide association studies.
 OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at github.com/gedankenstuecke/snpr

### 2012-10-25 25_10_2012_Graphs/plotPhenotypesAndUserPhenotypes.R
Replaced old graphs by newest versions, added R-code, input-data to reproduce

@@ -1,0 +1,10 @@
+svg("phenotypes_vs_userphenotypes.svg")
+par(mar=c(5, 4, 4, 6))
+plot(sort(strptime(pheno[,2], "%d.%m.%Y %H:%M")), pheno[,1],type="l", xlab="Time", ylab="Amount of unique phenotypes",xaxt="n")
+axis(1, at = seqYear, labels = format(seqYear, "%b-%y"), las = 1)
+par(new=T)
+plot(sort(strptime(usersp[,2], "%d.%m.%Y %H:%M")), usersp[,1],type="l",xlab="", yaxt="n",col="red",xaxt="n",ylab="")
+axis(4,col.axis="red")
+mtext("Amount of entered phenotypes", side=4,line=3,col="red")
+dev.off()
+

### 2012-10-25 25_10_2012_Graphs/plotUsers.r

@@ -1,0 +1,8 @@
+svg("users.svg")
+users <- read.table("users.txt", sep="\t")
+plot(strptime(users[,2], format="%d.%m.%Y %H:%M"), users[,1], type="l", ylab="Amount of user-accounts",xlab="Time", xaxt="n")
+dmin <- min(strptime(users[,2], format="%d.%m.%Y %H:%M"))
+dminjan <-  as.POSIXct(format(dmin, "%Y-1-1"))
+seqYear <- seq(dminjan, by = "1 month", to = max(strptime(users[,2], format="%d.%m.%Y %H:%M")) + (86400 * 365))
+axis(1, at = seqYear, labels = format(seqYear, "%b-%y"), las = 1)
+dev.off()

### 2012-10-25 timePlot.R

@@ -1,10 +1,0 @@
-svg("NormedUserPhenoPheno.svg")
-par(mar=c(5, 4, 4, 6))
-plot(sort(strptime(pheno[,1], "%d.%m.%Y %H:%M")), pheno[,2],type="l", xlab="Time", ylab="Count Phenotypes",xaxt="n")
-axis.POSIXct(1,sort(strptime(pheno[,1], "%d.%m.%Y %H:%M")),format="%b-%y")
-par(new=T)
-plot(sort(strptime(usersp[,1], "%d.%m.%Y %H:%M")), usersp[,2],type="l",xlab="", yaxt="n",col="red",xaxt="n",ylab="")
-axis(4,col.axis="red")
-mtext("Count User-Phenotypes",side=4,line=3,col="red")
-dev.off()
-

### 2012-10-25 paper_draft.tex
Added acknowledgment to openSNP's users

@@
 Fabian Zimmer$^{3}$,
-Julia Reda$^{4}$
+Julia Reda$^{4}$,
+Helge Rausch$^{5}$
 \\
@@
 \bf{4} Johannes-Gutenberg University, Mainz, Germany
+\\
+\bf{5} Hochschule für Technik und Wirtschaft, Berlin, Germany
 \\
@@
 \section*{Acknowledgments}
-We thank Dr. Manuel Corpas for constructive advice in grammar, spelling and structure of this study.
+We thank Dr.
+Manuel Corpas and Prof.
+Dr.
+Dave Edwards for constructive advice in grammar, spelling and structure of this study.
+We would also like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
 

### 2012-10-26 paper_draft.tex
Re-introduced Survey into main-text

@@
 
-As a follow up of a survey, which evaluated whether customers of DTC genetic testing would be willing to share their genotypic and phenotypic information (see Supplementary Materials), we designed openSNP, an online platform which enables DTC customers to freely share their genotypic and phenotypic data under a Creative Commons-Zero license.
-The phenotypic and genotypic data can be retrieved from openSNP using different methods, which also include different API standards to allow automated processing of the data.
-Additionally openSNP also enables users of DTC genetic testing to receive further information on their genotypes, through the crowdsourced resource SNPedia and primary publications, like those of the Public Library of Science journals.
+Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform.
+We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes.
+The genotypes are made available to researchers via the open Creative-Commons-Zero licence.
 
@@
 \section*{Results}
+\subsection*{Survey on Sharing Genetic Information}
+In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
+The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
+customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
+becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).
+
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
+didn't share the data with third parties.
+6.6 \% of the participants would not share their data.
+Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221).
+Those who are customers of a DTC company or are planning on becoming one in
+the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331).
+
+There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
+and those who are not planning on getting genotyped.
+The first group is likely to agree more strongly, on a five-point scale, with motivations for sharing genotypic information.
+On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations
+for not sharing their data, see table \ref{tab:motivations1}.
+
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who are already genotyped} & & \\ \hline 
+... curious & 1.159 & 0.193 \\ 
+... want to help scientists & 0.465 & 0.128 \\
+... for personal benefits & 0.448 & 0.183 \\ \hline
+\textbf{Motivation for not sharing in participants} & & \\ 
+\textbf{who are not planning to get genotyped} & & \\ \hline
+... fear of discrimination & 1.06 & 0.195 \\
+... breach of privacy & 0.821 & 0.225 \\
+... fear of personalized advertising & 0.848 & 0.208 \\ 
+... negative consequences for family members & 0.733 & 0.21 \\ \hline 
+\end{tabular}
+\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
+}
+\label{tab:motivations1}
+\end{table}
+
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would share with their DTC provider} & & \\ \hline
+... curiosity & 1.99 & 0.321 \\
+... want to help science & 1.57 & 0.199 \\
+... for personal benefits & 0.951 & 0.308 \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would not share with their DTC provider} & & \\ \hline
+... fear of discrimination & 1.52 & 0.322 \\
+... fear of consequences for family members & 1.146 & 0.32 \\
+... fear of personalized advertising & 1.112 & 0.357 \\ \hline
+\end{tabular}
+\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with their DTC provider with participants who would not share their data with their DTC provider.}
+\label{tab:motivations2}
+\end{table}
+
+Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
+the following motivations for sharing than those who would not share their data with their DTC company.
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the some motivations 
+for not sharing their data when compared to those who would share their data with their DTC company under any circumstances, for an overview of the motivations of both groups, see table \ref{tab:motivations2}.
+
+In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).
+
+In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist.
+Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346).
+Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).
+
+These survey-results indicate that there is a definite interest in customers of DTC-companies to share their results with other scientists.
+
+\subsubsection*{Sharing genotypic information}
 
@@
 
+\subsection*{Survey issues}
+
+As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, the results do not reflect the general population, but over-represent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
+
 \subsection*{Privacy, health implications and ethical considerations}
@@
 
+\subsection*{Survey on Sharing Genetic Information}
+The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
+blogs which focus on genetics and DTC genetic testing and social media like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
+
+The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
+(planned) customer-ship with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
+All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
+but only if they did not share my medical information with anybody else" and No".
+
+The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
+information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
+were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
+possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
+data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
+"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
+Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
+
+The survey data was analyzed with SPSS 19.
+
 \subsection*{Technical implementation of the platform}
@@
 Dave Edwards for constructive advice in grammar, spelling and structure of this study.
-We would also like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
+We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
 

### 2012-10-26 paper_draft.tex
added citations, edited text to accommodate for PGP & genome.gov data

@@
 Philipp Bayer$^{2}$, 
-Fabian Zimmer$^{3}$,
-Julia Reda$^{4}$,
-Helge Rausch$^{5}$
+Helge Rausch$^{3}$,
+Fabian Zimmer$^{4}$,
+Julia Reda$^{5}$
 \\
-\bf{1} Goethe University, Frankfurt am Main, Germany
+\bf{1} Molecular Ecology Group, Goethe University Frankfurt am Main, Germany
 \\
-\bf{2} University of Queensland, Brisbane, Australia
+\bf{2} Applied Bioinformatics Group, School of Agriculture and Food Sciences, The University of Queensland, Australia
 \\
-\bf{3} Westf\"alische Wilhelms Universit\"at, M\"unster, Germany
+\bf{3} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
 \\
-\bf{4} Johannes-Gutenberg University, Mainz, Germany
+\bf{4} Evolutionary Genomics and Transcriptomics Lab, Department of Genetics, Evolution and Environment, University College London, England
 \\
-\bf{5} Hochschule für Technik und Wirtschaft, Berlin, Germany
+\bf{5} Johannes Gutenberg University Mayence, Germany
 \\
@@
 In the light of those results we present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publicly publish at no cost their genetic data along with phenotypic information.
-Through this "crowdsourced" effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide area of studies, including Genome-wide association studies.
-OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at github.com/gedankenstuecke/snpr
+Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide area of studies, including Genome-wide association studies.
+OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
 
@@
 \section*{Author Summary}
-Missing for now.
 
@@
 The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases – for example Alzheimers, Parkinsons or different types of cancers (needs citations) – are at least partially heritable so the genome 
+Many diseases – for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} – are at least partially heritable so the genome
 of patients can be used for diagnostic purposes.
@@
 If genetic information for more than one individual is known, the analysis of
-allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
+allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
 (GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
@@
 often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
-like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an application programming interface (API) 
+like the \textit{Personal Genome Project} \cite{Ball24072012}, still do not allow for an easy re-use of the data, as they currently lack an application programming interface (API) 
 or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.
@@
 We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes.
-The genotypes are made available to researchers via the open Creative-Commons-Zero licence.
+The genotypes are made available to researchers via the open Creative-Commons-Zero license.
 
@@
 In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
-A total number of 15,229 documents \bastian{number needs to get updated}relevant to the SNPs listed in openSNP could be found in the publication databases of Mendeley, the Public Library of Science and in the crowdsourced SNPedia.
-Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}). For usability reasons, 
+A total number of 15,229 documents \bastian{number needs to get updated}relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia.
+Of the primary literature listed on Mendeley \& the Public Library of Science, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}).
+For usability reasons,
 SNPs are ranked by the amount of information gathered through the external services. The external services themselves are ranked by how easily non-scientists can understand information 
-from these sources and available this information.
+from these sources and how available this information is to the public.
 The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of
-the Public Library of Science. Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments. 
-An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
+the Public Library of Science and the curated databases of the \emph{GET Evidence System} and the \emph{NHGRI GWAS Catalog}.
+Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments.
+An entry on SNPedia is valued 2.5 times as high as a PLoS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.
 
@@
 the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-of publishing their data. To further improve this process we are looking forward to implement an informed consent-processes 
-by a provider like \textit{Consent for Research} (http://www.weconsent.us).
+of publishing their data. Further versions of openSNP may optionally include further consent-processes.
 
@@
 The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
-blogs which focus on genetics and DTC genetic testing and social media like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
+blogs which focus on genetics and DTC genetic testing and social media websites flike \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
 
@@
 the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, 
-through querying the content via the MediaWiki API. All databases are queried using the unique identifier of each SNP as the search term.
+through querying the content via the MediaWiki API.
+The \emph{NHGRI GWAS Catalog} and the \emph{GET Evidence System} provide complete dumps in plain text formats.
+Those are regularly downloaded and parsed.
+All databases are queried or parsed using the unique identifier of each SNP as the search term.
 
@@
 Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources 
-are handled using the ruby gem Resque and the standalone server Redis. Search features on the platform itself are implemented using Solr and the ruby gem Sunspot. 
+are handled using the ruby gem Resque and the standalone key-value storage server Redis. Search features on the platform itself are implemented using Solr and the ruby gem Sunspot. 
 Additionally, data can be requested from openSNP using the Distributed Annotation System. The required data is stored in a PostgreSQL database.  
@@
 Dave Edwards for constructive advice in grammar, spelling and structure of this study.
-We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
+Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser.
+For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams.
+Last but not least: We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
 

### 2012-10-27 paper_draft.tex
PNG-versions of the graphs, updated most stats (except # UserSNPs, still counting), moved tables, inserted new graphs

--- 2012-10-26 paper_draft.tex
@@
 
-\begin{table}
-\begin{tabular}{|l|l|l|}
-\hline
-& Turkey's HSD & \\ \cline{2-3}
-& Mean difference & SE \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & \\
-\textbf{who are already genotyped} & & \\ \hline 
-... curious & 1.159 & 0.193 \\ 
-... want to help scientists & 0.465 & 0.128 \\
-... for personal benefits & 0.448 & 0.183 \\ \hline
-\textbf{Motivation for not sharing in participants} & & \\ 
-\textbf{who are not planning to get genotyped} & & \\ \hline
-... fear of discrimination & 1.06 & 0.195 \\
-... breach of privacy & 0.821 & 0.225 \\
-... fear of personalized advertising & 0.848 & 0.208 \\ 
-... negative consequences for family members & 0.733 & 0.21 \\ \hline 
-\end{tabular}
-\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
-}
-\label{tab:motivations1}
-\end{table}
-
-\begin{table}
-\begin{tabular}{|l|l|l|}
-\hline
-& Turkey's HSD & \\ \cline{2-3}
-& Mean difference & SE \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & \\
-\textbf{who would share with their DTC provider} & & \\ \hline
-... curiosity & 1.99 & 0.321 \\
-... want to help science & 1.57 & 0.199 \\
-... for personal benefits & 0.951 & 0.308 \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & \\
-\textbf{who would not share with their DTC provider} & & \\ \hline
-... fear of discrimination & 1.52 & 0.322 \\
-... fear of consequences for family members & 1.146 & 0.32 \\
-... fear of personalized advertising & 1.112 & 0.357 \\ \hline
-\end{tabular}
-\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with their DTC provider with participants who would not share their data with their DTC provider.}
-\label{tab:motivations2}
-\end{table}
-
 Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
@@
 allows a complete reuse of the data without any constraints.
-Between the start of openSNP on 09/27/2011 and 12/18/2012, 214 people have signed 
-up with openSNP, and 79 genotyping files were made available. The openSNP 
-database lists 69,486,471 genotypes which are distributed over 1,938,603 unique SNPs.
-Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.\bastian{update all numbers}
+Between the start of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
+up with openSNP, and 270 genetic datasets were made available. The openSNP 
+database lists X genotypes which are distributed over 2,140,643 unique SNPs.
+Figures \ref{Figure1_label} and \ref{Figure2_label} depicts the increase of users and genotyping files over time.\bastian{update all numbers}
 
@@
 In the same timeframe as above, all users combined have 
-entered a total of 675 variations on 47 different phenotypes with those variations being 
-the different values on a given trait or phenotype. See figure \ref{Figure1_label} for the increase of phenotypic information over time.
-
-The mean number of users that have entered their variations for a single phenotype 
-is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
-entered their data per phenotype can be seen in figure \ref{Figure2_label}. The phenotype provided 
-by the most users is the eye color, which has been entered by 54 users. There are 
-two phenotypes which have so far only been provided by a single user: 
-the SAT Writing score and triglyceride-levels.\bastian{update all numbers and graphs}
+entered a total of 4743 variations on 130 different phenotypes with those variations being 
+the different values on a given trait or phenotype. The mean number of users that have entered their variations for a single phenotype 
+is 36.48. The distribution of how many users have 
+entered their data per phenotype , compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}.
+The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype.
 
@@
 A total number of 15,229 documents \bastian{number needs to get updated}relevant to the SNPs listed in openSNP could be found in the publication databases of Mendeley, the Public Library of Science and in the crowdsourced SNPedia.
-Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}). For usability reasons, 
+Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge. For usability reasons, 
 SNPs are ranked by the amount of information gathered through the external services. The external services themselves are ranked by how easily non-scientists can understand information 
@@
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.
+
+Users are also able to link their Fitbit-accounts to their user-accounts.
+Fitbit is a commercial service which lets their customers track their BMI, movement-data and sleep data.
+This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body- and sleep-developments over time.
 
@@
 	\begin{center}
-		\includegraphics[scale=0.35]{chart_growth.png}
+		\includegraphics[scale=0.35]{25_10_2012_Graphs/users.png}
 	\end{center}
 	\caption{
-	{\bf Growth of openSNP.} The increase in numbers for users, genotyping-files, phenotypes and their variation from 27.09.2011 to 16.12.2011 is shown.} 
+	{\bf Growth of openSNP-user-accounts.} The increase in numbers for users from 27.09.2011 to 27.10.2012 is shown.} 
 	\label{Figure1_label}
+\end{figure}
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.png}
+	\end{center}
+	\caption{
+	{\bf Growth of available genotypings.} The increase in numbers for genotyping-files from 27.09.2011 to 27.10.2012 is shown.} 
+	\label{Figure2_label}
+\end{figure}
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.png}
+	\end{center}
+	\caption{
+	{\bf Development of unique phenotypes and phenotypic information over time.} The x-axis shows the time-frame from start of the project until October 2012, the left y-axis shows how many unique phenotypes have been entered, and the right y-axis shows the amount of phenotypes users entered.}
+	\label{pheno}
 \end{figure}
@@
 	\begin{center}
-		\includegraphics[scale=0.40]{histogram_phenotypes.png}
-	\end{center}
-	\caption{
-	{\bf Histogram of users/phenotype-distribution.} The x-axis shows the minimum number of users who provide information for a phenotype, the y-axis shows how many phenotypes have at least that many users.}
-	\label{Figure2_label}
-\end{figure}
 
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.50]{paper_distribution.png}
-	\end{center}
-	\caption{
-	{\bf Distribution of external information gathered for SNPs in the openSNP-database.} Data on PLoS and SNPedia is openly available for every user.
-Publications on Mendeley are either Open Access (OA) or Closed Access (CA).}
-	\label{Figure3_label}
-\end{figure}
-
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.60]{uml_diagram.png}
 	\end{center}
@@
 \section*{Tables}
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who are already genotyped} & & \\ \hline 
+... curious & 1.159 & 0.193 \\ 
+... want to help scientists & 0.465 & 0.128 \\
+... for personal benefits & 0.448 & 0.183 \\ \hline
+\textbf{Motivation for not sharing in participants} & & \\ 
+\textbf{who are not planning to get genotyped} & & \\ \hline
+... fear of discrimination & 1.06 & 0.195 \\
+... breach of privacy & 0.821 & 0.225 \\
+... fear of personalized advertising & 0.848 & 0.208 \\ 
+... negative consequences for family members & 0.733 & 0.21 \\ \hline 
+\end{tabular}
+\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
+}
+\label{tab:motivations1}
+\end{table}
+
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would share with their DTC provider} & & \\ \hline
+... curiosity & 1.99 & 0.321 \\
+... want to help science & 1.57 & 0.199 \\
+... for personal benefits & 0.951 & 0.308 \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would not share with their DTC provider} & & \\ \hline
+... fear of discrimination & 1.52 & 0.322 \\
+... fear of consequences for family members & 1.146 & 0.32 \\
+... fear of personalized advertising & 1.112 & 0.357 \\ \hline
+\end{tabular}
+\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with their DTC provider with participants who would not share their data with their DTC provider.}
+\label{tab:motivations2}
+\end{table}
 %\begin{table}[!ht]

### 2012-10-27 paper_draft.tex

--- 2012-10-26 paper_draft.tex
@@
 An entry on SNPedia is valued 2.5 times as high as a PLoS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.
+
+Users are also able to link their Fitbit-accounts to their user-accounts.
+Fitbit is a commercial service which lets their customers track their BMI, movement-data and sleep data.
+This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body- and sleep-developments over time.
 

### 2012-10-27 paper_draft.tex
Added some more stats, changed title, only missing thing is the user-snps-counter
Finished work on all stats-numbers about SNPs.

@@
 {\Large
-\textbf{openSNP - Crowdsourcing Genome-Wide Association Studies}
+\textbf{openSNP - a crowdsourced web resource for personal genomics}
 }
@@
 Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide area of studies, including Genome-wide association studies.
-OpenSNP is hosted at www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
+OpenSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
 
@@
 up with openSNP, and 270 genetic datasets were made available. The openSNP 
-database lists X genotypes which are distributed over 2,140,643 unique SNPs.
-Figures \ref{Figure1_label} and \ref{Figure2_label} depicts the increase of users and genotyping files over time.\bastian{update all numbers}
+database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
+Figures \ref{Figure1_label} and \ref{Figure2_label} depicts the increase of users and genotyping files over time.
 
@@
 is 36.48. The distribution of how many users have 
-entered their data per phenotype , compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}.
-The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype.
+entered their data per phenotype, compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}.
+The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype (retrieved 10/27/2012).
 
@@
 In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
-A total number of 15,229 documents \bastian{number needs to get updated}relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia.
+A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia.
 Of the primary literature listed on Mendeley \& the Public Library of Science, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}).
+
 For usability reasons, 
@@
 The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.
+
 \section*{Discussion}
@@
 For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams.
-Last but not least: We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
+We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
 

### 2012-10-29 25_10_2012_Graphs/papers_pie.R
Added piechart and dotchart on distribution of annotational sources.

@@ -1,0 +1,9 @@
+names <- c("SNPedia", "PLOS", "Mendeley", "PGP", "Genome.gov")
+numbers <- c(8257, 1771, 11116, 289, 3095)
+svg("papers_pie.svg")
+pie(numbers, labels=names)
+dev.off()
+
+svg("papers_dotchart.svg")
+dotchart(numbers, labels=names, xlab="Amount of annotational entries", xlim=c(0, 12000))
+dev.off()

### 2012-10-29 paper_draft.tex

@@
 However, this data is not public for a variety of reasons and thus cannot be used in research.
-It seems reasonable to create a central open data-repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
+It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
 Here we present a survey which evaluates whether people are willing to publicly share their genetic information.
-In the light of those results we present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publicly publish at no cost their genetic data along with phenotypic information.
-Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide area of studies, including Genome-wide association studies.
+In the light of those results we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish at no cost their genetic data along with phenotypic information.
+Through this crowdsourced effort of collecting genetic and phenotypic information, openSNP has become a valuable resource, which can be used in a wide area of studies, including Genome-wide association studies.
 OpenSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
@@
 The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases – for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} – are at least partially heritable so the genome
+Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome
 of patients can be used for diagnostic purposes.
@@
 
-If genetic information for more than one individual is known, the analysis of
+If genetic information on more than one individual is known, the analysis of
 allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
@@
 - the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
-traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
+traits and diseases. With the consent of the customer, the data is used for association studies. 23andMe has published several 
 articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
-So far, over 30,000 23andme-customers have participated in 23andme's association studies, which proves that this data-source has a lot of potential for other researchers.
+So far, over 30,000 23andMe-customers have participated in 23andMe's association studies, which proves that this data source has a lot of potential for other researchers.
 
@@
 We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes.
-The genotypes are made available to researchers via the open Creative-Commons-Zero license.
+The genotypes are made available to researchers via the open Creative Commons Zero license.
 
@@
 \subsection*{Survey on Sharing Genetic Information}
-In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
+In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX, participated in our survey on sharing genetic information with the public. 
 The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
@@
 
-67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
+67.7 \% of all participants would share their data with their DTC company without any constraints, 25.8 \% would do so given the company 
 didn't share the data with third parties.
@@
 the following motivations for sharing than those who would not share their data with their DTC company.
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the some motivations 
-for not sharing their data when compared to those who would share their data with their DTC company under any circumstances, for an overview of the motivations of both groups, see table \ref{tab:motivations2}.
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with some motivations 
+for not sharing their data when compared to those who would share their data with their DTC company under any circumstances.
+For an overview of the motivations of both groups, see table \ref{tab:motivations2}.
 
@@
 
-These survey-results indicate that there is a definite interest in customers of DTC-companies to share their results with other scientists.
+These survey results indicate that there is a definite interest in customers of DTC companies to share their results with other scientists.
 
@@
 
-We created the openSNP project (http://opensnp.org) as an open, crowdsourced online platform for DTC customers interested in sharing their raw data and for researchers interested in performing GWAS or perform other types of analysis with the data.
+We created the openSNP project (http://opensnp.org) as an open, crowdsourced online platform for DTC customers interested in sharing their raw data and for researchers interested in performing GWAS or other types of analysis with the data.
 Customers of DTC testing are encouraged to share their genotyping results along with their phenotypic traits to enable easy access for researchers.
@@
 \subsection*{Sharing genotypic information}
-Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a webinterface to the openSNP 
+Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a web interface to the openSNP 
 project.
@@
 So far only the SNPs of the exome data sets are visualized on openSNP, but the downloads include all variation found in the exome.
-The uploaded data is published under the Creative Commons Zero-license, 
+The uploaded data is published under the Creative Commons Zero license, 
 which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
-allows a complete reuse of the data without any constraints.
+allows a complete re-use of the data without any constraints.
 Between the start of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
-up with openSNP, and 270 genetic datasets were made available. The openSNP 
+up with openSNP, and 270 genetic datasets have been made available. The openSNP 
 database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
-Figures \ref{Figure1_label} and \ref{Figure2_label} depicts the increase of users and genotyping files over time.
+Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files over time.
 
@@
 the expression and spelling of the same phenotype or variation. We implemented an 
-autocompletion-feature, which helps users to reuse already entered phenotypes.
+autocompletion feature, which helps users reuse already entered phenotypes.
 Users are encouraged to list as many phenotypes as possible through a simple 
 achievement system, rewarding users that upload their data and enter phenotypic 
-information with small badges that are shown on their profile pages.
+information with badges that are shown on their profile pages.
 
-In the same timeframe as above, all users combined have 
+In the same timeframe mentioned above, all users combined have 
 entered a total of 4743 variations on 130 different phenotypes with those variations being 
@@
 
-Users are also able to link their Fitbit-accounts to their user-accounts.
-Fitbit is a commercial service which lets their customers track their BMI, movement-data and sleep data.
-This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body- and sleep-developments over time.
+Users are also able to link their Fitbit accounts to their user-accounts.
+Fitbit is a commercial service which lets its customers track their BMI, movement and sleep data.
+This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body and sleep developments over time.
 
@@
 accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
-An example of how the DAS can be used can be found on openSNP where users genotypes are visualized inside a genome browser.
-So far all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now.
+An example of how the DAS can be used is implemented on openSNP, where users' genotypes are visualized inside a genome browser.
+So far, all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now.
 \bastian{should this maybe include an image as well?}
 
-The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format.
+The data is additionally available over a JSON API, which allows users to directly access data in the JSON format.
 The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.
@@
 %The advent of DTC genetic testing has led to new ethical and social issues. 
-Much of the critique on DTC genetic testing focusses on the practice 
+Much of the criticism of DTC genetic testing focusses on the practice 
 of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
@@
 the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-of publishing their data. Further versions of openSNP may optionally include further consent-processes.
+of publishing their data. Further versions of openSNP may optionally include further consent processes.
 
@@
 inference on the cause for any correlation.
-Furthermore, for a statistically sound analysis GWAS need a large enough sample size.
+Furthermore, for a statistically sound analysis, GWAS need a large enough sample size.
 Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
@@
 One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. 
-Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high 
+Data produced by DTC genetic testing companies is a promising source for such results, as those companies already have high 
 numbers of customers which are willing to pay for the genotyping by themselves.
@@
 The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
-blogs which focus on genetics and DTC genetic testing and social media websites flike \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
+blogs which focus on genetics and DTC genetic testing and social media websites like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
 
 The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
-(planned) customer-ship with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
-All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
+(planned) customership with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
+All participants were asked if they would be willing to share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
 but only if they did not share my medical information with anybody else" and No".
 
-The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
+The survey also asked some scaled questions, which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
 information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
@@
 OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
-two methods are implemented: \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
-based on the user's query. The second method is \textit{sources}, which advertises all DAS-sources for all genotypes present in openSNP.
+two methods are implemented: firstly \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
+based on the user's query. The second method is \textit{sources}, which advertises all DAS sources for all genotypes present in openSNP.
 
 A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
-published under the MIT-license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
+published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
 % Do NOT remove this, even if you are not including acknowledgments
@@
 For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams.
-We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.
+We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.
 

### 2012-10-29 paper_draft.tex
Changed my name slightly (because PB is all over pubmed, PEB is not.). Added dotchart, linked to in text.

@@
 Bastian Greshake$^{1,\ast}$, 
-Philipp Bayer$^{2}$, 
+Philipp E. Bayer$^{2}$, 
 Helge Rausch$^{3}$,
@@
 	\begin{center}
+		\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.png}
+	\end{center}
+	\caption{
+	{\bf Distribution of annotation-sources at openSNP.} Currently, SNP-annotations from SNPedia, PLOS, Mendeley, the \emph{GET Evidence System} and the \emph{NHGRI GWAS Catalog} are being collected.}
+	\label{Figure3_label}
+\end{figure}
 
+\begin{figure}[!ht]
+	\begin{center}
 	\end{center}

### 2012-10-29 25_10_2012_Graphs/papers_pie.R

@@ -1,5 +1,5 @@
-names <- c("SNPedia", "PLOS", "Mendeley", "PGP", "Genome.gov")
-numbers <- c(8257, 1771, 11116, 289, 3095)
+numbers <- c(289, 1771, 3095, 8257, 11116)
+names <- c("PGP", "PLOS", "NHGRI Catalog", "SNPedia", "Mendeley")
 svg("papers_pie.svg")
 pie(numbers, labels=names)
 dev.off()

### 2012-10-29 paper_draft.tex
Added note on insufficiently evaluated PGP-SNPs not being stored
work on tables

@@
 Those are regularly downloaded and parsed.
+SNPs that are described as 'Insufficiently evaluated' in the \emph{GET Evidence System} are not stored.
 All databases are queried or parsed using the unique identifier of each SNP as the search term.
@@
 \begin{table}
-\begin{tabular}{|l|l|l|}
+\begin{tabular}{|l|l|l|l|l|}
 \hline
-& Turkey's HSD & \\ \cline{2-3}
-& Mean difference & SE \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & \\
-\textbf{who are already genotyped} & & \\ \hline 
-... curious & 1.159 & 0.193 \\ 
-... want to help scientists & 0.465 & 0.128 \\
-... for personal benefits & 0.448 & 0.183 \\ \hline
-\textbf{Motivation for not sharing in participants} & & \\ 
-\textbf{who are not planning to get genotyped} & & \\ \hline
-... fear of discrimination & 1.06 & 0.195 \\
-... breach of privacy & 0.821 & 0.225 \\
-... fear of personalized advertising & 0.848 & 0.208 \\ 
-... negative consequences for family members & 0.733 & 0.21 \\ \hline 
+& Mean & Mean & Tukey's HSD & \\ 
+& \emph{Genotyped} & \emph{Not Genotyped} & Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & & & \\
+\textbf{who are already genotyped} & & & & \\ \hline 
+... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ 
+... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\
+... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline
+\textbf{Motivation for not sharing in participants} & & & & \\ 
+\textbf{who are not planning to get genotyped} & & & & \\ \hline
+... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\
+... breach of privacy & 3.01 & 3.68  & 0.821 & 0.225 \\
+... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ 
+... negative consequences for family members & 2.93 & 3.57  & 0.733 & 0.21 \\ \hline 
 \end{tabular}
@@
 \begin{table}
-\begin{tabular}{|l|l|l|}
+\begin{tabular}{|l|l|l|l|l|}
 \hline
-& Turkey's HSD & \\ \cline{2-3}
-& Mean difference & SE \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & \\
-\textbf{who would share with their DTC provider} & & \\ \hline
-... curiosity & 1.99 & 0.321 \\
-... want to help science & 1.57 & 0.199 \\
-... for personal benefits & 0.951 & 0.308 \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & \\
-\textbf{who would not share with their DTC provider} & & \\ \hline
-... fear of discrimination & 1.52 & 0.322 \\
-... fear of consequences for family members & 1.146 & 0.32 \\
-... fear of personalized advertising & 1.112 & 0.357 \\ \hline
+& Mean & Mean & Tukey's HSD & \\ 
+& \emph{Sharing} & \emph{Not Sharing} & Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & & & \\
+\textbf{who would share} & & & & \\ \hline
+... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\
+... want to help science & 4.44 & 3.19 & 1.57 & 0.199 \\
+... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & & & \\
+\textbf{who would not share} & & & & \\ \hline
+... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\
+... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\
+... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline
 \end{tabular}
-\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with their DTC provider with participants who would not share their data with their DTC provider.}
+\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
 \label{tab:motivations2}

### 2012-10-29 paper_draft.tex
finished tables & graphs

@@
 Through this crowdsourced effort of collecting genetic and phenotypic information, openSNP has become a valuable resource, which can be used in a wide area of studies, including Genome-wide association studies.
-OpenSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
+openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
 
@@
 In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
-A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia.
-Of the primary literature listed on Mendeley \& the Public Library of Science, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}).
+A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia (Figure \ref{Figure3_label}).
+Of the primary literature listed on Mendeley, the \emph{NHGRI GWAS Catalog} \& the Public Library of Science, about 20 \% are released in open access journals and can be accessed free of charge (Figure \ref{oa_label}), although probably not all publications on Mendeley are correctly flagged and the \emph{NHGRI GWAS Catalog} does not give details on whether a publication is Open Access or not.
+So the total number of Open Access publications might be higher.
 
@@
 
-Users are also able to link their Fitbit accounts to their user-accounts.
+Users are also able to link their Fitbit\cite{fitbit} accounts to their user-accounts.
 Fitbit is a commercial service which lets its customers track their BMI, movement and sleep data.
@@
 \subsection*{Data access}
-OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
+openSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
 grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
@@
 
-OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, once during registration for openSNP and once during 
+openSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, once during registration for openSNP and once during 
 the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
@@
 
-OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
+openSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
 two methods are implemented: firstly \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
@@
 
-A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
+A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in Figure \ref{Figure4_label}. The source code of openSNP is 
 published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
@@
 	\begin{center}
+		\includegraphics[scale=0.1]{25_10_2012_Graphs/percent_oa.png}
+	\end{center}
+	\caption{
+	{\bf Ratio of Open Access Publications.} Green pieces are Open Access. The \emph{NHGRI GWAS Catalog} doesn't give information about the Open Access status.}
+	\label{oa_label}
+\end{figure}
+
+\begin{figure}[!ht]
+	\begin{center}
+	  \includegraphics[scale=0.3]{latest_uml.png}
 	\end{center}
@@
 \begin{table}
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-& Mean & Mean & Tukey's HSD & \\ 
-& \emph{Genotyped} & \emph{Not Genotyped} & Mean difference & SE \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & & & \\
-\textbf{who are already genotyped} & & & & \\ \hline 
-... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ 
-... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\
-... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline
-\textbf{Motivation for not sharing in participants} & & & & \\ 
-\textbf{who are not planning to get genotyped} & & & & \\ \hline
-... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\
-... breach of privacy & 3.01 & 3.68  & 0.821 & 0.225 \\
-... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ 
-... negative consequences for family members & 2.93 & 3.57  & 0.733 & 0.21 \\ \hline 
-\end{tabular}
 \caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
 }
+\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
+\hline
+& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
+\hline
+\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
+\hline 
+... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
+... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
+... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
+\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
+... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
+... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
+... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
+... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
+\end{tabular}
 \label{tab:motivations1}
@@
 \begin{table}
-\begin{tabular}{|l|l|l|l|l|}
+\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
+\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
 \hline
-& Mean & Mean & Tukey's HSD & \\ 
-& \emph{Sharing} & \emph{Not Sharing} & Mean difference & SE \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & & & \\
-\textbf{who would share} & & & & \\ \hline
-... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\
-... want to help science & 4.44 & 3.19 & 1.57 & 0.199 \\
-... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline
-\textbf{Motivation for sharing genotypings in participants} & & & & \\
-\textbf{who would not share} & & & & \\ \hline
-... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\
-... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\
-... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline
+& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
+\hline
+\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
+\hline
+... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
+... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
+... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
+\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
+\hline
+... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
+... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
+... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
 \end{tabular}
-\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
 \label{tab:motivations2}

### 2012-10-29 paper_draft.tex
Added tif-conversions of my svg-plots. Changed in text, doesn't compile?
added wikimedia

@@
 Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP: only a subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, which can potentially lead to skewed GWAS-results.
+21 people, mainly from underrepresented demographics, have been offered free genotyping using funding provided by the Wikimedia Germany association in order to mitigate this bias.
 
@@
 	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/users.png}
+		\includegraphics[scale=0.35]{25_10_2012_Graphs/users.tif}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.png}
+		\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.tif}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.png}
+		\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.tif}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.png}
+		\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.tif}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.1]{25_10_2012_Graphs/percent_oa.png}
+		\includegraphics[scale=0.1]{25_10_2012_Graphs/percent_oa.tif}
 	\end{center}

### 2012-10-30 paper_draft.tex
for viewing changed images to png
Slight change of Dave Edwards' name - who knew US/UK/AUS lists academic titles different than Germany :P

@@
 Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP: only a subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, which can potentially lead to skewed GWAS-results.
-21 people, mainly from underrepresented demographics, have been offered free genotyping using funding provided by the Wikimedia Germany association in order to mitigate this bias.
 
@@
 Manuel Corpas and Prof.
-Dr.
-Dave Edwards for constructive advice in grammar, spelling and structure of this study.
+David Edwards for constructive advice in grammar, spelling and structure of this study.
 Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser.

### 2012-10-30 paper_draft.tex
Fixed two tiny typos, good to go from my end

@@
 The first group is likely to agree more strongly, on a five-point scale, with motivations for sharing genotypic information.
-On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations
-for not sharing their data, see table \ref{tab:motivations1}.
+On the other hand, those people who are not planning on getting genotyped are more likely to agree with the several motivations
+for not sharing their data (for an overview of these motivations, see table \ref{tab:motivations1}).
 
@@
 database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
-Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files over time.
+Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files since September 2011.
 
@@
 %The advent of DTC genetic testing has led to new ethical and social issues. 
-Much of the criticism of DTC genetic testing focusses on the practice 
-of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
+Much of the criticism of DTC genetic testing focuses on the practice 
+of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information 
 and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.

### 2013-05-15 paper_draft.tex
Recompiled, with new pie-chart
Removed (by comments) the survey.

@@
 It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
-Here we present a survey which evaluates whether people are willing to publicly share their genetic information.
-In the light of those results we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish at no cost their genetic data along with phenotypic information.
-Through this crowdsourced effort of collecting genetic and phenotypic information, openSNP has become a valuable resource, which can be used in a wide area of studies, including Genome-wide association studies.
+Here
+%we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
+we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish at no cost their genetic data along with phenotypic information.
+Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a
+%valuable 
+resource, which can be used in a wide area of studies, including Genome-wide association studies.
 openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
@@
 
-Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform.
-We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes.
+Here, we present 
+%the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform. We also present 
+openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes.
 The genotypes are made available to researchers via the open Creative Commons Zero license.
@@
 \section*{Results}
-\subsection*{Survey on Sharing Genetic Information}
-In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX, participated in our survey on sharing genetic information with the public. 
-The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
-customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
-becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).
-
-67.7 \% of all participants would share their data with their DTC company without any constraints, 25.8 \% would do so given the company 
-didn't share the data with third parties.
+%\subsection*{Survey on Sharing Genetic Information}
+%In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX, participated in our survey on sharing genetic information with the public. 
+%The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
+%customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
+%becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077). 
+%
+%67.7 \% of all participants would share their data with their DTC company without any constraints, 25.8 \% would do so given the company 
+%didn't share the data with third parties.
 6.6 \% of the participants would not share their data.
@@
 Those who are customers of a DTC company or are planning on becoming one in
-the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331).
-
-There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
-and those who are not planning on getting genotyped.
+%the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331). 
+%
+%
+%There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
+%and those who are not planning on getting genotyped.
 The first group is likely to agree more strongly, on a five-point scale, with motivations for sharing genotypic information.
 On the other hand, those people who are not planning on getting genotyped are more likely to agree with the several motivations
-for not sharing their data (for an overview of these motivations, see table \ref{tab:motivations1}).
-
-Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
-the following motivations for sharing than those who would not share their data with their DTC company.
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with some motivations 
-for not sharing their data when compared to those who would share their data with their DTC company under any circumstances.
+%for not sharing their data (for an overview of these motivations, see table \ref{tab:motivations1}).
+%
+%Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
+%the following motivations for sharing than those who would not share their data with their DTC company.
+%Those participants who are not willing to share data with their DTC company are likely to agree more strongly with some motivations 
+%for not sharing their data when compared to those who would share their data with their DTC company under any circumstances.
 For an overview of the motivations of both groups, see table \ref{tab:motivations2}.
-
-In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).
-
-In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist.
+%
+%In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).
+%
+%In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist.
 Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346).
 Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).
-
-These survey results indicate that there is a definite interest in customers of DTC companies to share their results with other scientists.
+%
+%These survey results indicate that there is a definite interest in customers of DTC companies to share their results with other scientists.
 
@@
 
-\subsection*{Survey issues}
-
-As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, the results do not reflect the general population, but over-represent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
+%\subsection*{Survey issues}
+%
+%As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, the results do not reflect the general population, but over-represent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
+%
 
@@
 
-As we have found in our survey on sharing such results (see supplementary methods), many DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
-There is a variety of ethical and privacy implications when it comes to DTC genetic testing\cite{Caulfield2011,Joh2011}.
+%As we have found in our survey on sharing such results (see supplementary methods), many 
+DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
+There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011}.
 
-Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
-or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
+%Our survey has shown that 
+People are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information.
+Policy makers start to react to those changes by introducing laws like the
 \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
@@
 
-\subsection*{Survey on Sharing Genetic Information}
-The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
-blogs which focus on genetics and DTC genetic testing and social media websites like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
-
-The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
-(planned) customership with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
-All participants were asked if they would be willing to share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
-but only if they did not share my medical information with anybody else" and No".
-
-The survey also asked some scaled questions, which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
-information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
-were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
-possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
-data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
-"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
-Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
-
-The survey data was analyzed with SPSS 19.
+%\subsection*{Survey on Sharing Genetic Information}
+%The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
+%blogs which focus on genetics and DTC genetic testing and social media websites like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.  
+%
+%The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
+%(planned) customership with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
+%All participants were asked if they would be willing to share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
+%but only if they did not share my medical information with anybody else" and No".
+%
+%The survey also asked some scaled questions, which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
+%information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
+%were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
+%possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
+%data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
+%"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
+%Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
+%
+%The survey data was analyzed with SPSS 19.
 
@@
 	\begin{center}
-		\includegraphics[scale=0.1]{25_10_2012_Graphs/percent_oa.png}
+		\includegraphics[scale=0.6]{25_10_2012_Graphs/new_pie.png}
 	\end{center}
@@
 \section*{Tables}
-\begin{table}
-\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
+%\begin{table}
+%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
 }
-\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
-\hline
-& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
-\hline
-\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
-\hline 
-... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
-... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
-... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
-\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
-... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
-... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
-... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
-... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
-\end{tabular}
-\label{tab:motivations1}
-\end{table}
-
-\begin{table}
-\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
-\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
-\hline
-& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
-\hline
-\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
-\hline
-... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
-... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
-... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
-\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
-\hline
-... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
-... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
-... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
-\end{tabular}
-\label{tab:motivations2}
-\end{table}
+%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
+%\hline
+%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
+%\hline
+%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
+%\hline 
+%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
+%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
+%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
+%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
+%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
+%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
+%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
+%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
+%\end{tabular}
+%\label{tab:motivations1}
+%\end{table}
+%
+%\begin{table}
+%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
+%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
+%\hline
+%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
+%\hline
+%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
+%\hline
+%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
+%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
+%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
+%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
+%\hline
+%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
+%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
+%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
+%\end{tabular}
+%\label{tab:motivations2}
+%\end{table}
 %\begin{table}[!ht]

### 2013-05-15 paper_draft.tex
Added my second affiliation.

@@
 Bastian Greshake$^{1,\ast}$, 
-Philipp E. Bayer$^{2}$, 
-Helge Rausch$^{3}$,
-Fabian Zimmer$^{4}$,
-Julia Reda$^{5}$
+Philipp E. Bayer$^{2,3}$, 
+Helge Rausch$^{4}$,
+Fabian Zimmer$^{5}$,
+Julia Reda$^{6}$
 \\
@@
 \\
-\bf{2} Applied Bioinformatics Group, School of Agriculture and Food Sciences, The University of Queensland, Australia
+\bf{2} School of Land, Crop and Food Sciences and Australian Centre for Plant Functional Genomics, University of Queensland, Brisbane, Australia
 \\
-\bf{3} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
+\bf{3} Australian Centre for Plant Functional Genomics, School of Agriculture and Food Sciences, University of Queensland, Brisbane, Australia
 \\
-\bf{4} Evolutionary Genomics and Transcriptomics Lab, Department of Genetics, Evolution and Environment, University College London, England
+\bf{4} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
 \\
-\bf{5} Johannes Gutenberg University Mayence, Germany
+\bf{5} Evolutionary Genomics and Transcriptomics Lab, Department of Genetics, Evolution and Environment, University College London, England
+\\
+\bf{6} Johannes Gutenberg University Mayence, Germany
 \\

### 2013-05-17 paper_draft.tex
Changed some minor formulations - added one paragraph about data validity

@@
 %we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
-we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish at no cost their genetic data along with phenotypic information.
+we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information.
 Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a
 %valuable 
-resource, which can be used in a wide area of studies, including Genome-wide association studies.
+resource for a wide area of studies, including Genome-wide association studies.
 openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
@@
 The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome
-of patients can be used for diagnostic purposes.
+Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome of patients can be used for diagnostic purposes.
 Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.
@@
 or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.
+
+%Enter here: definition, benefits of crowdsourcing
 
@@
 \subsection*{Sharing genotypic information}
-Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a web interface to the openSNP 
+Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme} and \textit{FamilyTreeDNA} via a web interface to the openSNP 
 project.
 There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers.
-So far only the SNPs of the exome data sets are visualized on openSNP, but the downloads include all variation found in the exome.
+So far only the SNPs of the exome data sets are stored on openSNP, but the downloads include all variation found in the exome.
 The uploaded data is published under the Creative Commons Zero license, 
@@
 \subsection*{Data access}
-openSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
+openSNP offers complete access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
 grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
@@
 An example of how the DAS can be used is implemented on openSNP, where users' genotypes are visualized inside a genome browser.
-So far, all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now.
-\bastian{should this maybe include an image as well?}
+All chromosomal positions are based on the human reference genome NCBI37, as this is the standard reference used by DTC providers right now.
 
-The data is additionally available over a JSON API, which allows users to directly access data in the JSON format.
+The data is additionally available over a JSON API which allows users to directly access data in the JSON format.
 The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.
+
+% Compare openSNP to other platforms here
+
+\subsection*{Other crowdsourcing platforms}
+
+Compare to SNPedia, PGP
 
@@
 21 people, mainly from underrepresented demographics, have been offered free genotyping using funding provided by the Wikimedia Germany association in order to mitigate this bias.
+
+Furthermore, it is impossible to verify whether users who have uploaded data are actually the sources for that data.
+This opens the venue to potentially malicious usage, as genotypings from strangers can be uploaded, as well as misinformation about phenotypes can be entered.
+The openSNP project has currently no means to verify the validity of data given to us by users.
+Of course, users can always delete their data or contact the team to delete stored data.
+No copies are kept at any time.
 

### 2013-05-18 paper_draft.tex
did some work on the discussion of different platforms and crowdsourcing

@@
 Helge Rausch$^{4}$,
-Fabian Zimmer$^{5}$,
-Julia Reda$^{6}$
+Julia Reda$^{5}$
 \\
@@
 \\
-\bf{5} Evolutionary Genomics and Transcriptomics Lab, Department of Genetics, Evolution and Environment, University College London, England
-\\
-\bf{6} Johannes Gutenberg University Mayence, Germany
+\bf{5} Johannes Gutenberg University Mayence, Germany
 \\
@@
 %valuable 
-resource for a wide area of studies, including Genome-wide association studies.
-openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
+resource for a wide area of studies, including Genome-Wide Association S
+tudies. openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
 
@@
 
-%Enter here: definition, benefits of crowdsourcing
+%Enter here: definition, benefits of sourcing
+Crowdsourcing, giving a task into the hands of a potentially large number of Ð mainly intrinsically motivated Ð people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well, with Galaxy Zoo and FoldIt (CITATIONS NEEDED) being two of the most-well known examples.
+One of the main benefits of crowdsourcing is that small contributions to a project pile up to create a larger work, which would have been virtually impossible to create otherwise. 
+Galaxy Zoo enabled thousands of amateur astronomers to walk through thousands of telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
+Similarly crowdsourcing can not only be applied to analyzing data, but also to collecting data.
+Ornithology has been doing this for years (CITATION NEEDED).
+With the advent of DTC genetic testing and the internet a similar approach can now be brought to human genetics.
 
@@
 \subsection*{Other crowdsourcing platforms}
+The most similar projects to openSNP are the SNPedia and the Personal Genome Project.
+The focus of the SNPedia is the aggregation and summarization of primary scientific literature on SNPs.
+The project uses a Wiki to store and display the data collected by volunteers contributing to the project.
+The data is mainly organized by the unique Rs-ID, as given by dbSNP.
+If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data.
+For individual SNPs pages may list scientific literature and summaries on the found impact can be given.
+As those pages are largely created manually and not automatically through database access these summaries may not be complete.
+openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has.
+While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access the different data sets.
 
-Compare to SNPedia, PGP
+The Personal Genome Project (PGP) has its focus on collecting and hosting genetic data as well as phenotypic one.
+Unlike openSNP they do not offer a completely open enrollment.
+For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent.
+This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g.
+non-US citizens).
+Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes.
+While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be extracted from their database in a highly manual fashion.
+The annotation database of the PGP is not aimed at delivering specific publications but instead focusses more on specific traits (IS THAT RIGHT?!).
+Their annotation data is incorporated into openSNP as well.
+
+COMPARE MORE. DO A TABLE. SHOULD THIS MAYBE BE PUT IN THE DISCUSSION INSTEAD OF RESULTS?
 

### 2013-06-03 paper_draft.tex
small changes in discussion, added citation for crowdsourced data

@@
 
+% for checkmarks
+\usepackage{pifont}% http://ctan.org/pkg/pifont
+\newcommand{\cmark}{\ding{51}}%
+\newcommand{\xmark}{\ding{55}}%
 % graphicx package, useful for including eps and pdf graphics
@@
 \section*{Abstract}
-Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants.
+Genome-Wide Association Studies are widely used to correlate phenotypic traits with genetic variants.
 These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups.
@@
 %Enter here: definition, benefits of sourcing
-Crowdsourcing, giving a task into the hands of a potentially large number of Ð mainly intrinsically motivated Ð people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well, with Galaxy Zoo and FoldIt (CITATIONS NEEDED) being two of the most-well known examples.
+Crowdsourcing, giving a task into the hands of a potentially large number of - mainly intrinsically motivated - people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well.
 One of the main benefits of crowdsourcing is that small contributions to a project pile up to create a larger work, which would have been virtually impossible to create otherwise.
+This especially benefits scientists who might not have enough funding or time to create data.
+Or in cases where the amounts of data are too large to be analyzed by researchers alone.
+% added this, because neither galaxy zoo or foldit crowdsource data creating but only the analysis
+ Galaxy Zoo and FoldIt \cite{Eiben2012, GalaxyZoo} are two of the best known examples. 
 Galaxy Zoo enabled thousands of amateur astronomers to walk through thousands of telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
 Similarly crowdsourcing can not only be applied to analyzing data, but also to collecting data.
-Ornithology has been doing this for years (CITATION NEEDED).
+This approach has been shown to work when it comes to tracking bird migration \cite{CrowdsourcingReview2010}.
 With the advent of DTC genetic testing and the internet a similar approach can now be brought to human genetics.
+
+There have been studies investigating how likely customers of such companies are to share their data.
+\cite{Darst2013} investigated the likelihood of 2,024 individuals to share their test results with their health-care providers and found that 26.5\% (540 individuals) did share their results with physician or health-care provider.
+Those that shared where older, had a higher income and were less concerned about testing or the privacy implications of this sharing compared to customers that didn't share their data.
+
+Other studies have shown that DTC-customers see themselves as being well-informed.
+Interviews with early adopters have shown that these customers are better informed and more skeptical about the capabilities of genotypings than expected \cite{McGowan2010}.
+However, in another study of early adopters 32\% of customers had misperceptions about personal genomic testing \cite{Gollust2012}.
+Of these participants, 92\% intended to share their results with physicians in order to receive medical recommendations.
+In general, in both studies participants chose this technology to be better informed about genetic risks and to satisfy their own curiosity.
 
@@
 allows a complete re-use of the data without any constraints.
-Between the start of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
-up with openSNP, and 270 genetic datasets have been made available. The openSNP 
-database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
+Between the launch of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
+up with openSNP, and 270 genetic datasets have been made available. As of 10/27/2012, the openSNP database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
 Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files since September 2011.
@@
 
-% Compare openSNP to other platforms here
-
-\subsection*{Other crowdsourcing platforms}
-The most similar projects to openSNP are the SNPedia and the Personal Genome Project.
-The focus of the SNPedia is the aggregation and summarization of primary scientific literature on SNPs.
-The project uses a Wiki to store and display the data collected by volunteers contributing to the project.
-The data is mainly organized by the unique Rs-ID, as given by dbSNP.
-If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data.
-For individual SNPs pages may list scientific literature and summaries on the found impact can be given.
-As those pages are largely created manually and not automatically through database access these summaries may not be complete.
-openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has.
-While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access the different data sets.
-
-The Personal Genome Project (PGP) has its focus on collecting and hosting genetic data as well as phenotypic one.
-Unlike openSNP they do not offer a completely open enrollment.
-For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent.
-This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g.
-non-US citizens).
-Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes.
-While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be extracted from their database in a highly manual fashion.
-The annotation database of the PGP is not aimed at delivering specific publications but instead focusses more on specific traits (IS THAT RIGHT?!).
-Their annotation data is incorporated into openSNP as well.
-
-COMPARE MORE. DO A TABLE. SHOULD THIS MAYBE BE PUT IN THE DISCUSSION INSTEAD OF RESULTS?
-
 \section*{Discussion}
@@
 
+\subsection*{Other crowdsourcing platforms}
+Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview).
+The focus of the SNPedia is the aggregation and summary of primary scientific literature on SNPs.
+The project uses a Wiki to store and display the data collected by volunteers contributing to the project.
+The data is mainly organized by the unique Rs-ID, as given by dbSNP.
+If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data.
+For individual SNPs pages may list scientific literature and summaries on the found impact can be given.
+As those pages are largely created manually and not automatically through database access these summaries may not be complete.
+openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has.
+While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access SNPs or data subsets in the different data sets.
+Similarly there is no way for users of SNPedia to share their phenotypes in a machine readable format.
+
+The Personal Genome Project (PGP) has its focus on collecting and hosting genetic as well as phenotypic data.
+Unlike openSNP they do not offer a completely open enrollment.
+For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent.
+This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g.
+non-US citizens).
+Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes.
+While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be manually extracted from their database.
+The annotation database of the PGP is not aimed at delivering specific publications but instead focuses more on specific traits.
+The annotation data stored by the PGP is incorporated into openSNP as well.
+
+PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.
+This helps patients in understanding their diseases better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%).
+Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}.
+Alternatively access to the data they store can be licensed by researchers for a fee.
+
+There are some projects that use gamification to let players work with crowdsourced scientific data.
+For example, FoldIt is a puzzle game that lets players fold protein structures in order to achieve optimal structures.
+Players of Foldit have been able to identify protein structure and were even able to improve the activity of existing proteins structures \cite{Eiben2012}.
+Another example is Galaxy Zoo \cite{GalaxyZoo}, which allows everyone to perform classification tasks for galaxies based on images collected by the Sloan Digital Sky Survey.
+
+Unlike PatientsLikeMe, the PGP, or openSNP - which give the task of collecting the data into the hands of the crowd - Foldit and Galaxy Zoo limit themselves to analyzing data which was previously collected by scientists.
+
 \subsection*{Privacy, health implications and ethical considerations}
@@
 Much of the criticism of DTC genetic testing focuses on the practice 
-of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information 
-and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.
+of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.
 
 %As we have found in our survey on sharing such results (see supplementary methods), many 
-DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
 There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011}.
+But studies show that DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information \cite{Darst2013}.
 
 %Our survey has shown that 
-People are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information.
-Policy makers start to react to those changes by introducing laws like the
+While people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information \cite{Wolinsky2005}, policy makers start to react to those changes by introducing laws like the
 \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
-widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
+widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data \cite{Lee2009}.
 
@@
 Furthermore, for a statistically sound analysis, GWAS need a large enough sample size.
-Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
+Nevertheless, GWAS are still frequently used and new associations are still being discovered \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
 
@@
 By including the option of uploading exome data sets the platform already is capable of adjusting for changes in the type of data generated by DTC genetic testing.
+Future improvements made on openSNP will address interoperability with other platforms and tools in Personal Genomics, amongst others: The standardization of phenotypes, the inclusion of further annotation sources and support for a wider range of data sets, including full genome data.
 
@@
 \section*{Materials and Methods}
+\subsection*{Ethics Statement}
+The director of the ethics commission of the Goethe University Frankfurt am Main, Germany teaching hospital confirmed that this study does not fall within their remit.
 
@@
 David Edwards for constructive advice in grammar, spelling and structure of this study.
-Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser.
+Further thanks go to Samantha Clark, Fabian Zimmer and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser.
 For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams.
@@
 \section*{Tables}
+\begin{table}
+\caption{Comparison of crowdsourced genetics platforms.}
+\begin{tabular}{|l|l|l|l|l|l|l|}
+\hline
+Name & Provides  & Provides & Open  & API & IRB approval & License \\ 
+& annotation & data & enrollment & & & \\
+\hline
+SNPedia & \checkmark & \xmark & N/A & \checkmark & N/A & CC-NC-SA 3.0 \\ 
+\hline
+PGP & \checkmark & \checkmark & \xmark & \xmark & \checkmark & CC-BY 1.0, CC0 \\
+\hline
+PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & \checkmark & Closed, CC-BY-SA 3.0 \\
+\hline
+openSNP & \checkmark & \checkmark & \checkmark & \checkmark & \xmark & CC-BY 3.0 \\
+\hline
+\end{tabular}
+\label{tab:platforms}
+\end{table}
 %\begin{table}

### 2013-06-04 paper_draft.tex
added thanks to reviewers
whops, linebreak in abstract
Many minor changes to formulations. Replaced disease by illness. Normalized some spellings. Added short explanatory sentences.

@@
 However, this data is not public for a variety of reasons and thus cannot be used in research.
-It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
-Here
+It seems reasonable to create a central open data repository for such data.
+%, but it was previously unknown if and how people would submit their data to such a repository.
 %we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
-we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information.
+Here we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information.
 Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a
 %valuable 
-resource for a wide area of studies, including Genome-Wide Association S
-tudies. openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
+resource for a wide area of studies, including Genome-Wide Association Studies.
+openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
 
@@
 The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome of patients can be used for diagnostic purposes.
+Many illnesses - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome of patients can be used for diagnostic purposes.
 Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.
@@
 If genetic information on more than one individual is known, the analysis of
-allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
+allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with illnesses and other inheritable traits. Genome-Wide Association Studies
 (GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
-enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or disease.
+enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or illness.
 This method does not allow inference of causal differences but merely identifies correlations. 
@@
 such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
-5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}.
+5000 SNPs have been linked to different illnesses and traits \cite{Hindorff2009}.
 
@@
 These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
-receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
+receive an analysis of the results, as well as a raw file that includes the customer's individual genotypes. In 2011, 23andMe 
 alone had over 100,000 customers \cite{23andMe2011}
 - the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
-traits and diseases. With the consent of the customer, the data is used for association studies. 23andMe has published several 
-articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
+traits and illnesses. With the consent of the customer, the data is used for association studies. 23andMe has published several 
+studies in which known findings are replicated together with new associations for disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
 So far, over 30,000 23andMe-customers have participated in 23andMe's association studies, which proves that this data source has a lot of potential for other researchers.
@@
 One of the main benefits of crowdsourcing is that small contributions to a project pile up to create a larger work, which would have been virtually impossible to create otherwise.
-This especially benefits scientists who might not have enough funding or time to create data.
-Or in cases where the amounts of data are too large to be analyzed by researchers alone.
-% added this, because neither galaxy zoo or foldit crowdsource data creating but only the analysis
+This approach especially benefits scientists who might not have enough funding or time to create data, or in cases where the amounts of data are too large to be analyzed by researchers alone.
  Galaxy Zoo and FoldIt \cite{Eiben2012, GalaxyZoo} are two of the best known examples. 
-Galaxy Zoo enabled thousands of amateur astronomers to walk through thousands of telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
+Galaxy Zoo enables amateur astronomers to walk through telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
 Similarly crowdsourcing can not only be applied to analyzing data, but also to collecting data.
 This approach has been shown to work when it comes to tracking bird migration \cite{CrowdsourcingReview2010}.
-With the advent of DTC genetic testing and the internet a similar approach can now be brought to human genetics.
+With the advent of DTC genetic testing and the internet a similar approach can now be applied to human genetics.
 
 There have been studies investigating how likely customers of such companies are to share their data.
-\cite{Darst2013} investigated the likelihood of 2,024 individuals to share their test results with their health-care providers and found that 26.5\% (540 individuals) did share their results with physician or health-care provider.
+\cite{Darst2013} investigated the likelihood of 2,024 individuals to share their test results with their health-care providers and found that 26.5\% (540 individuals) did share their results with their physician or health-care provider.
 Those that shared where older, had a higher income and were less concerned about testing or the privacy implications of this sharing compared to customers that didn't share their data.
@@
 There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers.
-So far only the SNPs of the exome data sets are stored on openSNP, but the downloads include all variation found in the exome.
+Due to space constraints openSNP currently only stores the SNPs of the exome data sets.
 The uploaded data is published under the Creative Commons Zero license, 
@@
 A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia (Figure \ref{Figure3_label}).
-Of the primary literature listed on Mendeley, the \emph{NHGRI GWAS Catalog} \& the Public Library of Science, about 20 \% are released in open access journals and can be accessed free of charge (Figure \ref{oa_label}), although probably not all publications on Mendeley are correctly flagged and the \emph{NHGRI GWAS Catalog} does not give details on whether a publication is Open Access or not.
-So the total number of Open Access publications might be higher.
+Of the primary literature listed on Mendeley, the \emph{NHGRI GWAS Catalog} \& the Public Library of Science, about 20 \% are released in open access journals and can be accessed free of charge (Figure \ref{oa_label}), although probably not all publications on Mendeley are correctly flagged and the \emph{NHGRI GWAS Catalog} does not give details on whether a publication is open access or not.
+So the total number of open access publications might be higher.
 
@@
 Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments.
-An entry on SNPedia is valued 2.5 times as high as a PLoS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.
+An entry on SNPedia is valued 2.5 times as high as a PLOS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.
 
@@
 openSNP offers complete access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
-grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
+grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table.
+For privacy reasons openSNP does not log any IPs.
+The genetic data is also
 accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
@@
 
-\subsection*{Other crowdsourcing platforms}
+\subsection*{Comparing openSNP to other crowdsourcing platforms}
 Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview).
@@
 PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.
-This helps patients in understanding their diseases better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%).
+This helps patients in understanding their illnesses better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%).
 Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}.
@@
 there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
-There are two possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes.
-The other one
+There are three possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes.
+The second one
 would be to make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
 variations that have not been available from the start.
+The third one is to exclude users from the phenotype-creation process by allowing users to select their phenotypes from a pre-given set of possible variations.
 
-In both cases it makes it harder 
-for users to enter their data which raises the bar for participation.
+In the first two cases it makes it harder 
+for users to enter their data which raises the bar for participation, and the third case doesn't let users participate at all.
 We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform additional quality control.
@@
 This opens the venue to potentially malicious usage, as genotypings from strangers can be uploaded, as well as misinformation about phenotypes can be entered.
-The openSNP project has currently no means to verify the validity of data given to us by users.
+The openSNP project has currently no means to verify the validity of data uploaded by users.
 Of course, users can always delete their data or contact the team to delete stored data.
-No copies are kept at any time.
+Old backups of the database are deleted so that at any given time, there are only two backups.
+This means that deleted data disappears from the webpage immediately and will disappear after two months in the backend where it isn't accessible to the public.
 
@@
 \subsection*{Technical implementation of the platform}
-The main platform is implemented using the web framework Ruby on Rails 3.0.10. Postgres 9.2 is used as the main database backend for Rails. 
+The main platform is implemented using the web framework Ruby on Rails 3.2.13. Postgres 9.2 is used as the main database backend for Rails. 
 The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs 
@@
 For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams.
+Thanks go to Misha Angrist and the anonymous reviewers for their helpful comments on this manuscript.
 We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.
@@
 	\caption{
-	{\bf Ratio of Open Access Publications.} Green pieces are Open Access. The \emph{NHGRI GWAS Catalog} doesn't give information about the Open Access status.}
+	{\bf Ratio of open access Publications.} Green pieces are open access. The \emph{NHGRI GWAS Catalog} doesn't give information about the open access status.}
 	\label{oa_label}
@@
 \begin{table}
-\caption{Comparison of crowdsourced genetics platforms.}
+\caption{Comparison of crowdsourced genetics platforms. N/A = Not Applicable, \checkmark = Present, \xmark = Absent}
 \begin{tabular}{|l|l|l|l|l|l|l|}

### 2013-06-04 paper_draft.tex
changes julia

@@
 Here we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information.
-Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a
+Through this crowdsourced effort of collecting genetic and phenotypic information, openSNP has become a
 %valuable 
@@
 alone had over 100,000 customers \cite{23andMe2011}
-- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
+-- the company realizes the potential of performing GWAS with this amount of data by using surveys to ask their customers about 
 traits and illnesses. With the consent of the customer, the data is used for association studies. 23andMe has published several 
@@
 %Enter here: definition, benefits of sourcing
-Crowdsourcing, giving a task into the hands of a potentially large number of - mainly intrinsically motivated - people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well.
+Crowdsourcing, giving a task into the hands of a potentially large number of -- mainly intrinsically motivated -- people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well.
 One of the main benefits of crowdsourcing is that small contributions to a project pile up to create a larger work, which would have been virtually impossible to create otherwise.
@@
 Galaxy Zoo enables amateur astronomers to walk through telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
-Similarly crowdsourcing can not only be applied to analyzing data, but also to collecting data.
+Similarly, crowdsourcing can not only be applied to analyzing data, but also to collecting data.
 This approach has been shown to work when it comes to tracking bird migration \cite{CrowdsourcingReview2010}.
-With the advent of DTC genetic testing and the internet a similar approach can now be applied to human genetics.
+With the advent of DTC genetic testing and the internet, a similar approach can now be applied to human genetics.
 
@@
 \cite{Darst2013} investigated the likelihood of 2,024 individuals to share their test results with their health-care providers and found that 26.5\% (540 individuals) did share their results with their physician or health-care provider.
-Those that shared where older, had a higher income and were less concerned about testing or the privacy implications of this sharing compared to customers that didn't share their data.
+Those that shared were older, had a higher income and were less concerned about testing or the privacy implications of sharing their data compared to customers that didn't share their data.
 
 Other studies have shown that DTC-customers see themselves as being well-informed.
-Interviews with early adopters have shown that these customers are better informed and more skeptical about the capabilities of genotypings than expected \cite{McGowan2010}.
-However, in another study of early adopters 32\% of customers had misperceptions about personal genomic testing \cite{Gollust2012}.
+Interviews with early adopters have shown that these customers are better informed and more skeptical about the capabilities of genotyping than expected \cite{McGowan2010}.
+However, in another study of early adopters, 32\% of customers had misperceptions about personal genomic testing \cite{Gollust2012}.
 Of these participants, 92\% intended to share their results with physicians in order to receive medical recommendations.
-In general, in both studies participants chose this technology to be better informed about genetic risks and to satisfy their own curiosity.
+In both studies, participants generally chose this technology to be better informed about genetic risks and to satisfy their own curiosity.
 
@@
 There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers.
-Due to space constraints openSNP currently only stores the SNPs of the exome data sets.
+Due to space constraints, openSNP currently only stores the SNPs of the exome data sets.
 The uploaded data is published under the Creative Commons Zero license, 
-which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
+which -- in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} -- 
 allows a complete re-use of the data without any constraints.
@@
 grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table.
-For privacy reasons openSNP does not log any IPs.
+For privacy reasons, openSNP does not log any IPs.
 The genetic data is also
@@
 The data is mainly organized by the unique Rs-ID, as given by dbSNP.
-If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data.
-For individual SNPs pages may list scientific literature and summaries on the found impact can be given.
-As those pages are largely created manually and not automatically through database access these summaries may not be complete.
-openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has.
-While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access SNPs or data subsets in the different data sets.
-Similarly there is no way for users of SNPedia to share their phenotypes in a machine readable format.
+If Rs-IDs are missing, the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data.
+For individual SNPs, pages may list scientific literature and summaries on the found impact can be given.
+As those pages are largely created manually and not automatically through database access, these summaries may not be complete.
+openSNP utilizes the SNPedia by crawling their data for SNPs, the summary of the impact and the magnitude a SNP has.
+While they offer a page listing download-URLs, the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access SNPs or data subsets in the different data sets.
+Similarly, there is no way for users of SNPedia to share their phenotypes in a machine readable format.
 
 The Personal Genome Project (PGP) has its focus on collecting and hosting genetic as well as phenotypic data.
-Unlike openSNP they do not offer a completely open enrollment.
-For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent.
+Unlike openSNP, they do not offer a completely open enrollment.
+For each participant of the PGP, eligibility has to be established and participants have to give IRB approved informed consent.
 This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g.
 non-US citizens).
-Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes.
-While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be manually extracted from their database.
-The annotation database of the PGP is not aimed at delivering specific publications but instead focuses more on specific traits.
+Depending on the specific use one has for the data, the PGPs enrollment policy might be preferable to the open approach openSNP takes.
+While the PGP stores genotyping data as well as exome and genome data sets, it is currently impossible to access this data through an API, instead data has to be manually extracted from their database.
+The annotation database of the PGP is not aimed at delivering specific publications, but instead focuses more on specific traits.
 The annotation data stored by the PGP is incorporated into openSNP as well.
@@
 PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.
-This helps patients in understanding their illnesses better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%).
-Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}.
-Alternatively access to the data they store can be licensed by researchers for a fee.
+This helps patients in gaining a better understanding of their illnesses -- 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%).
+Some subsets of data stored in PatientsLikeMe are open to the public and have been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}.
+Alternatively, access to the data they store can be licensed by researchers for a fee.
 
@@
 For example, FoldIt is a puzzle game that lets players fold protein structures in order to achieve optimal structures.
-Players of Foldit have been able to identify protein structure and were even able to improve the activity of existing proteins structures \cite{Eiben2012}.
+Players of Foldit have been able to identify protein structures and were even able to improve the activity of existing protein structures \cite{Eiben2012}.
 Another example is Galaxy Zoo \cite{GalaxyZoo}, which allows everyone to perform classification tasks for galaxies based on images collected by the Sloan Digital Sky Survey.
 
-Unlike PatientsLikeMe, the PGP, or openSNP - which give the task of collecting the data into the hands of the crowd - Foldit and Galaxy Zoo limit themselves to analyzing data which was previously collected by scientists.
+Unlike PatientsLikeMe, the PGP, or openSNP -- which give the task of collecting the data into the hands of the crowd -- Foldit and Galaxy Zoo limit themselves to analyzing data which was previously collected by scientists.
 
@@
 the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-of publishing their data. Further versions of openSNP may optionally include further consent processes.
+of publishing their data. Further versions of openSNP may include further consent processes.
 
@@
 
-Furthermore, it is impossible to verify whether users who have uploaded data are actually the sources for that data.
+Furthermore, it is impossible to verify whether users who have uploaded data are actually the sources of that data.
 This opens the venue to potentially malicious usage, as genotypings from strangers can be uploaded, as well as misinformation about phenotypes can be entered.
-The openSNP project has currently no means to verify the validity of data uploaded by users.
+The openSNP project has currently no means of verifying the validity of data uploaded by users.
 Of course, users can always delete their data or contact the team to delete stored data.
@@
 While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise.
-By including the option of uploading exome data sets the platform already is capable of adjusting for changes in the type of data generated by DTC genetic testing.
+By including the option of uploading exome data sets, the platform is already capable of adjusting for changes in the type of data generated by DTC genetic testing.
 Future improvements made on openSNP will address interoperability with other platforms and tools in Personal Genomics, amongst others: The standardization of phenotypes, the inclusion of further annotation sources and support for a wider range of data sets, including full genome data.

### 2013-06-12 paper_draft.tex
fixed linebreaks
well, made the F for figure uppercase so they might find it.

@@
 is 36.48. The distribution of how many users have 
-entered their data per phenotype, compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}.
+entered their data per phenotype, compared to the amount of unique phenotypes, can be seen in Figure \ref{pheno}.
 The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype (retrieved 10/27/2012).
@@
 \begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/users.png}
-	\end{center}
+	%\begin{center}
+	%	\includegraphics[scale=0.35]{25_10_2012_Graphs/users.png}
+	%\end{center}
 	\caption{
@@
 \begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.png}
-	\end{center}
+	%\begin{center}
+	%	\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.png}
+	%\end{center}
 	\caption{
@@
 \begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.png}
-	\end{center}
+	%\begin{center}
+	%	\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.png}
+	%\end{center}
 	\caption{
@@
 \begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.png}
-	\end{center}
+	%\begin{center}
+	%	\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.png}
+	%\end{center}
 	\caption{
@@
 \begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.6]{25_10_2012_Graphs/new_pie.png}
-	\end{center}
+	%\begin{center}
+	%	\includegraphics[scale=0.6]{25_10_2012_Graphs/new_pie.png}
+	%\end{center}
 	\caption{
@@
 \begin{figure}[!ht]
-	\begin{center}
-	  \includegraphics[scale=0.3]{latest_uml.png}
-	\end{center}
+	%\begin{center}
+	%  \includegraphics[scale=0.3]{latest_uml.png}
+	%\end{center}
 	\caption{

### 2013-09-07 paper_draft.tex
first edit for 2nd major revision

@@
 enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or illness.
-This method does not allow inference of causal differences but merely identifies correlations. 
+While method does not allow inference of causal differences but merely identifies correlations they can serve as a valuable tool for the unbiased discovery of candidate loci which then can be checked up in functional follow up studies \cite{10.1371/journal.pgen.1002584}, leading to a deeper understanding of diseases and thus potentially to new drug targets.
 The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
@@
 like the \textit{Personal Genome Project} \cite{Ball24072012}, still do not allow for an easy re-use of the data, as they currently lack an application programming interface (API) 
-or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.
+or other methods by which researchers could download the data.
+Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}, as their consent forms only allow residents of the United States to apply.
 
@@
 %As we have found in our survey on sharing such results (see supplementary methods), many 
-There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011}.
-But studies show that DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information \cite{Darst2013}.
+There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011,Joly2013}.
+But studies show that DTC customers are willing to share their results given the right circumstances and personal benefits gained through the sharing, while being aware that sharing genetic data can lead to misuse of the data and consequences like genetic discrimination  \cite{Darst2013}.
 
 %Our survey has shown that 
-While people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information \cite{Wolinsky2005}, policy makers start to react to those changes by introducing laws like the
-\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
-widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data \cite{Lee2009}.
+As people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information \cite{Wolinsky2005}, policy makers start to react to those changes by having introduced laws like the
+\textit{Genetic Information Non-Discrimination Act} (GINA) in the United States or the \emph{Gendiagnostikgesetz} (GenDG) in Germany to minimize the impact of
+widely available genetic information.
+DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data \cite{Lee2009}.
+Neither GINA nor the GenDG offer a complete protection of genetic discrimination, as certain things, such as life insurances, are not covered by those laws.
 
@@
 of publishing their data. Further versions of openSNP may include further consent processes.
+
+For users of openSNP the biggest potential problem is legal genetic discrimination in fields not covered by laws as GINA or GenDG once their public data is re-identified.
+While users can register pseudonymously this should not be seen as ultimate protection against re-identification.
+A recent study again showed that meta data, potentially attached to genetic profiles, such as date of birth, gender and postal code, can be be used to re-identify individuals on a name basis \cite{Sweeney2013}.
+A similar approach utilized genetic markers on the Y chromosome along with genealogical databases and meta data such as age and state to infer sure names and from there on the individuals \cite{Gymrek18012013}.
+Thus users need to be aware of the potential of re-idenification through providing meta data along with their genetic information and the genetic discrimination that could follow.
 
@@
 
-By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other 
+By offering a platform where customers of DTC genetic testing can publish their data openSNP for the general public tries to enable and facilitate the re-use of this already generated data.
+Allowing interested parties to use the data for studies on their own allows people to perform studies without the need to generate genetic data on their own.
+Additionally the data can be used to enrich other data sets in order to overcome limited sample sizes, which is especially of interest for rare diseases.
+
+In crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other 
 open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP. 

### 2013-09-12 paper_draft.tex
Added a few sentences, fixed some minor things

@@
 %
+Here, we present openSNP, a crowdsourced resource that enables customers of DTC testing companies to share their genotypings with researchers and receive new annotation for their genetic variants.
+We have shown that customers of DTC companies are willing to share their data at no cost to researchers around the world and are willing to annotate their data with phenotypes.
 
@@
 widely available genetic information.
-DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data \cite{Lee2009}.
+DTC genetic testing companies themselves also try to create communities that help in educating their customers about the risks of releasing genetic data \cite{Lee2009}.
 Neither GINA nor the GenDG offer a complete protection of genetic discrimination, as certain things, such as life insurances, are not covered by those laws.
@@
 Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP: only a subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, which can potentially lead to skewed GWAS-results.
-21 people, mainly from underrepresented demographics, have been offered free genotyping using funding provided by the Wikimedia Germany association in order to mitigate this bias.
+21 people, mainly from underrepresented demographics, have been offered free genotypings using funding provided by the Wikimedia Germany association in order to mitigate this bias.
 

### 2013-09-12 paper_draft.tex
minor changes to philipps revision

--- 2013-09-07 paper_draft.tex
@@
 %
+Here, we present openSNP, a crowdsourced resource that enables customers of DTC testing companies to share their genotypings with researchers and receive new annotations for their genetic variants.
+With the number of active users already present on openSNP we have shown that customers of DTC companies are willing to share their data at no cost to researchers around the world and are willing to annotate their data with phenotypes.
 
@@
 widely available genetic information.
-DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data \cite{Lee2009}.
+DTC genetic testing companies themselves also try to create online communities - like the 23andMe community forums - that help in educating their customers about the risks of releasing genetic data \cite{Lee2009}.
 Neither GINA nor the GenDG offer a complete protection of genetic discrimination, as certain things, such as life insurances, are not covered by those laws.
@@
 inference on the cause for any correlation.
-Furthermore, for a statistically sound analysis, GWAS need a large enough sample size.
+Furthermore, for a statistically sound analysis, GWAS need a large enough sample size, which often isn't too easy to obtain.
+Either because generating the needed amount of data still is a cost factor or because it is hard to find enough participants for the case conditions, for example if rare diseases are to be studied.
 Nevertheless, GWAS are still frequently used and new associations are still being discovered \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
@@
 
-By offering a platform where customers of DTC genetic testing can publish their data openSNP for the general public tries to enable and facilitate the re-use of this already generated data.
-Allowing interested parties to use the data for studies on their own allows people to perform studies without the need to generate genetic data on their own.
+openSNP tries to enable and facilitate the re-use of this already generated data by offering a platform where customers of DTC genetic testing can publish their results into the public domain.
+Allowing interested parties to use the data for their own research allows scientists to perform studies without the need to generate genetic data sets on their own.
 Additionally the data can be used to enrich other data sets in order to overcome limited sample sizes, which is especially of interest for rare diseases.

### 2013-09-12 paper_draft.tex
Changed one minor formulation
typos, some grammar etc.

@@
 enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or illness.
-While method does not allow inference of causal differences but merely identifies correlations they can serve as a valuable tool for the unbiased discovery of candidate loci which then can be checked up in functional follow up studies \cite{10.1371/journal.pgen.1002584}, leading to a deeper understanding of diseases and thus potentially to new drug targets.
+While the method does not allow inference of causal differences but merely identifies correlations, it can serve as a valuable tool for the unbiased discovery of candidate loci, which then can be checked up in functional follow-up studies \cite{10.1371/journal.pgen.1002584}, leading to a deeper understanding of diseases and thus potentially to new drug targets.
 The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
@@
 receive an analysis of the results, as well as a raw file that includes the customer's individual genotypes. In 2011, 23andMe 
-alone had over 100,000 customers \cite{23andMe2011}
--- the company realizes the potential of performing GWAS with this amount of data by using surveys to ask their customers about 
+alone had over 100,000 customers \cite{23andMe2011}. The company realizes the potential of performing GWAS with this amount of data by using surveys to ask their customers about 
 traits and illnesses. With the consent of the customer, the data is used for association studies. 23andMe has published several 
@@
 
-Other studies have shown that DTC-customers see themselves as being well-informed.
+Other studies have shown that DTC customers see themselves as being well-informed.
 Interviews with early adopters have shown that these customers are better informed and more skeptical about the capabilities of genotyping than expected \cite{McGowan2010}.
@@
 Here, we present openSNP, a crowdsourced resource that enables customers of DTC testing companies to share their genotypings with researchers and receive new annotations for their genetic variants.
-With the number of active users already present on openSNP we have shown that customers of DTC companies are willing to share their data at no cost to researchers around the world and are willing to annotate their data with phenotypes.
+With the number of active users already present on openSNP, we have shown that customers of DTC companies are willing to share their data at no cost to researchers around the world and are willing to annotate their data with phenotypes.
 
@@
 There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011,Joly2013}.
-But studies show that DTC customers are willing to share their results given the right circumstances and personal benefits gained through the sharing, while being aware that sharing genetic data can lead to misuse of the data and consequences like genetic discrimination  \cite{Darst2013}.
+Nevertheless, studies show that DTC customers are willing to share their results given the right circumstances and personal benefits gained through the sharing, while being aware that sharing genetic data can lead to misuse of the data and consequences such as genetic discrimination  \cite{Darst2013}.
 
@@
 DTC genetic testing companies themselves also try to create online communities - like the 23andMe community forums - that help in educating their customers about the risks of releasing genetic data \cite{Lee2009}.
-Neither GINA nor the GenDG offer a complete protection of genetic discrimination, as certain things, such as life insurances, are not covered by those laws.
+Neither GINA nor the GenDG offer complete protection from genetic discrimination, as certain areas, such as life insurances, are not covered by those laws.
 
@@
 
-For users of openSNP the biggest potential problem is legal genetic discrimination in fields not covered by laws as GINA or GenDG once their public data is re-identified.
-While users can register pseudonymously this should not be seen as ultimate protection against re-identification.
-A recent study again showed that meta data, potentially attached to genetic profiles, such as date of birth, gender and postal code, can be be used to re-identify individuals on a name basis \cite{Sweeney2013}.
-A similar approach utilized genetic markers on the Y chromosome along with genealogical databases and meta data such as age and state to infer sure names and from there on the individuals \cite{Gymrek18012013}.
-Thus users need to be aware of the potential of re-idenification through providing meta data along with their genetic information and the genetic discrimination that could follow.
+For users of openSNP, the biggest potential problem is legal genetic discrimination, in fields not covered by laws such as GINA or GenDG, once their public data is re-identified.
+As the genetic information itself is highly personalized the anonymous sharing of genetic data is impossible.
+And while users can register pseudonymously, this should not be seen as ultimate protection against re-identification.
+A recent study once again showed that metadata, potentially attached to genetic profiles, such as date of birth, gender and postal code, can be be used to re-identify individuals on a name basis \cite{Sweeney2013}.
+A similar approach utilized genetic markers on the Y chromosome along with genealogical databases and metadata such as age and state to infer surnames and from there on the individuals \cite{Gymrek18012013}.
+Thus users need to be aware of the potential of re-identification through providing metadata along with their genetic information and the genetic discrimination that could follow.
 
@@
 inference on the cause for any correlation.
-Furthermore, for a statistically sound analysis, GWAS need a large enough sample size, which often isn't too easy to obtain.
+Furthermore, for a statistically sound analysis, GWAS need a large enough sample size, which often is not easy to obtain.
 Either because generating the needed amount of data still is a cost factor or because it is hard to find enough participants for the case conditions, for example if rare diseases are to be studied.
@@
 Data produced by DTC genetic testing companies is a promising source for such results, as those companies already have high 
-numbers of customers which are willing to pay for the genotyping by themselves.
+numbers of customers who are willing to pay for the genotyping by themselves.
 

### 2013-09-13 paper_draft.tex
fixed conflicts
the version and rebuttal sent back to plos1

@@
 There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers.
-Due to space constraints, openSNP currently only stores the SNPs of the exome data sets.
+Due to space constraints on the database level, openSNP currently only displays the SNPs of the exome data sets on the website but the whole VCF files can be downloaded.
 The uploaded data is published under the Creative Commons Zero license, 
@@
 Here, we present openSNP, a crowdsourced resource that enables customers of DTC testing companies to share their genotypings with researchers and receive new annotations for their genetic variants.
-With the number of active users already present on openSNP, we have shown that customers of DTC companies are willing to share their data at no cost to researchers around the world and are willing to annotate their data with phenotypes.
+Through a of number of active users already present on openSNP, we have shown that at least some customers of DTC companies are willing to share their data at no cost to researchers around the world and are willing to annotate their data with phenotypes.
 
@@
 Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper.
-However, GWAS can only detect correlations of SNPs with those traits and do not allow
-inference on the cause for any correlation.
+However, GWAS can only detect correlations of SNPs with those traits and do not allow inference on the cause for any correlation.
 Furthermore, for a statistically sound analysis, GWAS need a large enough sample size, which often is not easy to obtain.

### 2013-12-12 paper_draft.tex
made change to clarify SNPedia API
added citation for ethics approval and minor rewrite

@@
 \subsection*{Ethics Statement}
-The director of the ethics commission of the Goethe University Frankfurt am Main, Germany teaching hospital confirmed that this study does not fall within their remit.
+In line with the German regulations and the ethics approval system for biomedical studies \cite{Pinkerton2002} we contacted the ethics commission of the Goethe University Frankfurt am Main, Germany teaching hospital.
+The director of it confirmed that this study does not fall within their remit.
 
@@
 \begin{table}
-\caption{Comparison of crowdsourced genetics platforms. N/A = Not Applicable, \checkmark = Present, \xmark = Absent}
+\caption{Comparison of crowdsourced genetics platforms.
+N/A = Not Applicable, \checkmark = Present, \xmark = Absent *SNPedia only provides an API to webpages of individual SNPs, not access to genetic data of individuals}
 \begin{tabular}{|l|l|l|l|l|l|l|}
@@
 \hline
-SNPedia & \checkmark & \xmark & N/A & \checkmark & N/A & CC-NC-SA 3.0 \\ 
+SNPedia & (\checkmark*) & \xmark & N/A & \checkmark & N/A & CC-NC-SA 3.0 \\ 
 \hline

### 2013-12-13 paper_draft.tex
last changes to affiliations & text. with rebuttal

@@
 \\
-Bastian Greshake$^{1,\ast}$, 
-Philipp E. Bayer$^{2,3}$, 
-Helge Rausch$^{4}$,
-Julia Reda$^{5}$
+Bastian Greshake$^{1,2,\ast}$, 
+Philipp E. Bayer$^{3,4}$, 
+Helge Rausch$^{5}$,
+Julia Reda$^{6}$
 \\
@@
 \\
-\bf{2} School of Land, Crop and Food Sciences and Australian Centre for Plant Functional Genomics, University of Queensland, Brisbane, Australia
+\bf{2} Department for Applied Bioinformatics, Institute for Cell Biology and Neuroscience, Goethe University Frankfurt am Main, Germany 
 \\
-\bf{3} Australian Centre for Plant Functional Genomics, School of Agriculture and Food Sciences, University of Queensland, Brisbane, Australia
+\bf{3} School of Land, Crop and Food Sciences and Australian Centre for Plant Functional Genomics, University of Queensland, Brisbane, Australia
 \\
-\bf{4} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
+\bf{4} Australian Centre for Plant Functional Genomics, School of Agriculture and Food Sciences, University of Queensland, Brisbane, Australia
 \\
-\bf{5} Johannes Gutenberg University Mayence, Germany
+\bf{5} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
+\\
+\bf{6} Johannes Gutenberg University Mainz, Germany
 \\
@@
 In line with the German regulations and the ethics approval system for biomedical studies \cite{Pinkerton2002} we contacted the ethics commission of the Goethe University Frankfurt am Main, Germany teaching hospital.
-The director of it confirmed that this study does not fall within their remit.
+Its director confirmed that this study does not fall within their remit.
 

### 2014-01-22 paper_draft.tex
edited according to PLOS publishing guidelines for final proof

@@
 \newcommand{\philipp}[1]{\authornote{PB: #1}}
+\usepackage[utf8x]{inputenc}
+\DeclareUnicodeCharacter{2717}{✗}
+%\usepackage{newunicodechar}
+%\newunicodechar{✓}{\checkmark}
+%\newunicodechar{✘}{\xmark}
+
 %% END MACROS SECTION
@@
 \\
-\bf{1} Molecular Ecology Group, Goethe University Frankfurt am Main, Germany
+\bf{1} Molecular Ecology Group, Biodiversity \& Climate Research Centre, Frankfurt am Main, Germany
 \\
-\bf{2} Department for Applied Bioinformatics, Institute for Cell Biology and Neuroscience, Goethe University Frankfurt am Main, Germany 
+\bf{2} Department for Applied Bioinformatics, Institute for Cell Biology and Neuroscience, Goethe University, Frankfurt am Main, Germany 
 \\
@@
 \\
-\bf{5} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
+\bf{5} Hochschule f\"ur Technik und Wirtschaft, Berlin, Germany 
 \\
-\bf{6} Johannes Gutenberg University Mainz, Germany
+\bf{6} Johannes Gutenberg University, Mainz, Germany
 \\
@@
 openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step.
-% Author Summary not valid for PLoS ONE submissions.   
-\section*{Author Summary}
 
@@
 \caption{Comparison of crowdsourced genetics platforms.
-N/A = Not Applicable, \checkmark = Present, \xmark = Absent *SNPedia only provides an API to webpages of individual SNPs, not access to genetic data of individuals}
+N/A = Not Applicable, x = Present, - = Absent *SNPedia only provides an API to webpages of individual SNPs, not access to genetic data of individuals}
 \begin{tabular}{|l|l|l|l|l|l|l|}
@@
 \hline
-SNPedia & (\checkmark*) & \xmark & N/A & \checkmark & N/A & CC-NC-SA 3.0 \\ 
+SNPedia & (x*) & - & N/A & x & N/A & CC-NC-SA 3.0 \\ 
 \hline
-PGP & \checkmark & \checkmark & \xmark & \xmark & \checkmark & CC-BY 1.0, CC0 \\
+PGP & x & x & - & - & x & CC-BY 1.0, CC0 \\
 \hline
-PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & \checkmark & Closed, CC-BY-SA 3.0 \\
+PatientsLikeMe & - & x & x & - & x & Closed, CC-BY-SA 3.0 \\
 \hline
-openSNP & \checkmark & \checkmark & \checkmark & \checkmark & \xmark & CC-BY 3.0 \\
+openSNP & x & x & x & x & - & CC-BY 3.0 \\
 \hline

### 2014-01-22 paper_draft_tif.tex
Added new paper-version with tif instead of png
removed \includegraphics
worked on images. submitted :)

--- 2012-10-30 paper_draft.tex
@@
 So far, all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now.
-\bastian{should this maybe include an image as well?}
 
@@
 
+\subsection*{Ethics Statement}
+The survey was taken anonymously by the participants and analyzed anonymously, thus IRB approval was deemed unnecessary according to US regulations in \textit{45 CFR 46.101(b)} and in accordance the Hessian data protection officer (http://www.datenschutz.hessen.de/wf001.htm\#entry2223).
+
 \subsection*{Survey on Sharing Genetic Information}
@@
 	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/users.png}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.png}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.png}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.png}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.1]{25_10_2012_Graphs/percent_oa.png}
 	\end{center}
@@
 	\begin{center}
-	  \includegraphics[scale=0.3]{latest_uml.png}
 	\end{center}

### 2014-01-22 supplementary_methods.tex

--- 2012-10-30 paper_draft.tex
@@
 {\Large
-\textbf{openSNP - a crowdsourced web resource for personal genomics}
+\textbf{openSNP - Crowdsourcing Genome-Wide Association Studies - Supplementary Methods}
 }
@@
 Bastian Greshake$^{1,\ast}$, 
-Philipp E. Bayer$^{2}$, 
-Helge Rausch$^{3}$,
-Fabian Zimmer$^{4}$,
-Julia Reda$^{5}$
+Philipp Bayer$^{2}$, 
+Fabian Zimmer$^{3}$,
+Julia Reda$^{4}$
 \\
-\bf{1} Molecular Ecology Group, Goethe University Frankfurt am Main, Germany
+\bf{1} Goethe University, Frankfurt am Main, Germany
 \\
-\bf{2} Applied Bioinformatics Group, School of Agriculture and Food Sciences, The University of Queensland, Australia
+\bf{2} University of Queensland, Brisbane, Australia
 \\
-\bf{3} Hochschule f\"ur Technik und Wirtschaft Berlin, Germany 
+\bf{3} Westf\"alische Wilhelms Universit\"at, M\"unster, Germany
 \\
-\bf{4} Evolutionary Genomics and Transcriptomics Lab, Department of Genetics, Evolution and Environment, University College London, England
-\\
-\bf{5} Johannes Gutenberg University Mayence, Germany
+\bf{4} Johannes-Gutenberg University, Mainz, Germany
 \\
@@
 
-% Please keep the abstract between 250 and 300 words
-\section*{Abstract}
-Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants.
-These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups.
-However, it is necessary to have a large enough sample size to find statistically significant correlations.
-Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist.
-Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies.
-However, this data is not public for a variety of reasons and thus cannot be used in research.
-It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository.
-Here we present a survey which evaluates whether people are willing to publicly share their genetic information.
-In the light of those results we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish at no cost their genetic data along with phenotypic information.
-Through this crowdsourced effort of collecting genetic and phenotypic information, openSNP has become a valuable resource, which can be used in a wide area of studies, including Genome-wide association studies.
-openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr
+\section*{Introduction}
+To evaluate whether customers of Direct-To-Consumer (DTC) genetic testing would be willing to share their raw data, along with phenotypic annotations, freely on the web a survey on the topic was set up and spread among customers and potential customers of DTC testing.
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step.
-% Author Summary not valid for PLoS ONE submissions.   
-\section*{Author Summary}
+\section*{Materials and Methods}
+\subsection*{Survey on Sharing Genetic Information}
+The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
+blogs which focus on genetics and DTC genetic testing and social media like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
 
-\section*{Introduction}
+The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
+(planned) customer-ship with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
+All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
+but only if they did not share my medical information with anybody else" and No".
 
-The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
-Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome
-of patients can be used for diagnostic purposes.
-Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.
+The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
+information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
+were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
+possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
+data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
+"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
+Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
 
-%Kicked out the following due to Corpas' argument
-%The comparison of DNA sequences from individuals in a population can reveal variable sites, which are of major interest in investigating diseases. 
-%A variation of only one nucleotide length at a given site is called a single Nucleotide Polymorphisms (SNP). 
-%The different nucleotides at this site are defined as an allele.
-%For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
-%Different methods have been developed to read SNPs in an individual, a
-%process called genotyping. A widespread used method for genotyping individuals is the use of microarrays.
-%In comparison to whole genome sequencing only sites which are known to be variable are analysed, making microarrays cheaper and faster \cite{Brown1999}.
-
-If genetic information on more than one individual is known, the analysis of
-allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
-(GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
-enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or disease.
-This method does not allow inference of causal differences but merely identifies correlations. 
-The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
-to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
-such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
-5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}.
-
-GWAS are not only performed inside the traditional scientific community. 
-Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
-These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
-receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
-alone had over 100,000 customers \cite{23andMe2011}
-- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
-traits and diseases. With the consent of the customer, the data is used for association studies. 23andMe has published several 
-articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
-So far, over 30,000 23andMe-customers have participated in 23andMe's association studies, which proves that this data source has a lot of potential for other researchers.
-
-The generation of biomedical data by private companies raises concerns about privacy \cite{23andMe2012}, 
-liability and consent \cite{Caulfield2011}. 
-Nevertheless, in some instances individual customers are willingly sharing their data. Most do so by uploading their data to
-their personal website or to open software repositories like \textit{GitHub}. 
-This data is scattered and unorganized, making it hard to use in studies. While projects like SNPedia try to keep track of all 
-the publicly available genotyping files \cite{Cariaso2011}, they usually do not provide the information necessary to perform GWAS, as the phenotypic information is 
-often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
-like the \textit{Personal Genome Project} \cite{Ball24072012}, still do not allow for an easy re-use of the data, as they currently lack an application programming interface (API) 
-or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.
-
-Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform.
-We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes.
-The genotypes are made available to researchers via the open Creative Commons Zero license.
+The survey data was analyzed with SPSS 19.
 
@@
 \section*{Results}
+
 \subsection*{Survey on Sharing Genetic Information}
-In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX, participated in our survey on sharing genetic information with the public. 
+In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
 The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
@@
 
-67.7 \% of all participants would share their data with their DTC company without any constraints, 25.8 \% would do so given the company 
+67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
 didn't share the data with third parties.
@@
 
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who are already genotyped} & & \\ \hline 
+... curious & 1.159 & 0.193 \\ 
+... want to help scientists & 0.465 & 0.128 \\
+... for personal benefits & 0.448 & 0.183 \\ \hline
+\textbf{Motivation for not sharing in participants} & & \\ 
+\textbf{who are not planning to get genotyped} & & \\ \hline
+... fear of discrimination & 1.06 & 0.195 \\
+... breach of privacy & 0.821 & 0.225 \\
+... fear of personalized advertising & 0.848 & 0.208 \\ 
+... negative consequences for family members & 0.733 & 0.21 \\ \hline 
+\end{tabular}
+\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
+}
+\label{tab:motivations1}
+\end{table}
+
+\begin{table}
+\begin{tabular}{|l|l|l|}
+\hline
+& Turkey's HSD & \\ \cline{2-3}
+& Mean difference & SE \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would share with their DTC provider} & & \\ \hline
+... curiosity & 1.99 & 0.321 \\
+... want to help science & 1.57 & 0.199 \\
+... for personal benefits & 0.951 & 0.308 \\ \hline
+\textbf{Motivation for sharing genotypings in participants} & & \\
+\textbf{who would not share with their DTC provider} & & \\ \hline
+... fear of discrimination & 1.52 & 0.322 \\
+... fear of consequences for family members & 1.146 & 0.32 \\
+... fear of personalized advertising & 1.112 & 0.357 \\ \hline
+\end{tabular}
+\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with their DTC provider with participants who would not share their data with their DTC provider.}
+\label{tab:motivations2}
+\end{table}
+
 Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
 the following motivations for sharing than those who would not share their data with their DTC company.
-Those participants who are not willing to share data with their DTC company are likely to agree more strongly with some motivations 
-for not sharing their data when compared to those who would share their data with their DTC company under any circumstances.
-For an overview of the motivations of both groups, see table \ref{tab:motivations2}.
+Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the some motivations 
+for not sharing their data when compared to those who would share their data with their DTC company under any circumstances, for an overview of the motivations of both groups, see table \ref{tab:motivations2}.
 
@@
 
-These survey results indicate that there is a definite interest in customers of DTC companies to share their results with other scientists.
-
-\subsubsection*{Sharing genotypic information}
-
-We created the openSNP project (http://opensnp.org) as an open, crowdsourced online platform for DTC customers interested in sharing their raw data and for researchers interested in performing GWAS or other types of analysis with the data.
-Customers of DTC testing are encouraged to share their genotyping results along with their phenotypic traits to enable easy access for researchers.
-Users of openSNP can create a personal profile, discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.
-
-People interested in using the data of openSNP can download complete dumps of the genotypic and phenotypic information or use query API endpoints utilizing JavaScript Object Notation (JSON) objects or the Distributed Annotation System (DAS) \cite{Dowell2001}.
-
-\subsection*{Sharing genotypic information}
-Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a web interface to the openSNP 
-project.
-There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers.
-So far only the SNPs of the exome data sets are visualized on openSNP, but the downloads include all variation found in the exome.
-The uploaded data is published under the Creative Commons Zero license, 
-which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
-allows a complete re-use of the data without any constraints.
-Between the start of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
-up with openSNP, and 270 genetic datasets have been made available. The openSNP 
-database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
-Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files over time.
-
-\subsection*{Crowdsourcing phenotypes}
-Users are able to create new phenotypes that are not yet 
-listed by openSNP. 
-The specification of these phenotypes is open and not limited 
-to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
-the expression and spelling of the same phenotype or variation. We implemented an 
-autocompletion feature, which helps users reuse already entered phenotypes.
-Users are encouraged to list as many phenotypes as possible through a simple 
-achievement system, rewarding users that upload their data and enter phenotypic 
-information with badges that are shown on their profile pages.
-
-In the same timeframe mentioned above, all users combined have 
-entered a total of 4743 variations on 130 different phenotypes with those variations being 
-the different values on a given trait or phenotype. The mean number of users that have entered their variations for a single phenotype 
-is 36.48. The distribution of how many users have 
-entered their data per phenotype, compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}.
-The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype (retrieved 10/27/2012).
-
-\subsection*{Connection to external services}
-In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
-A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia (Figure \ref{Figure3_label}).
-Of the primary literature listed on Mendeley, the \emph{NHGRI GWAS Catalog} \& the Public Library of Science, about 20 \% are released in open access journals and can be accessed free of charge (Figure \ref{oa_label}), although probably not all publications on Mendeley are correctly flagged and the \emph{NHGRI GWAS Catalog} does not give details on whether a publication is Open Access or not.
-So the total number of Open Access publications might be higher.
-
-For usability reasons, 
-SNPs are ranked by the amount of information gathered through the external services. The external services themselves are ranked by how easily non-scientists can understand information 
-from these sources and how available this information is to the public.
-The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of
-the Public Library of Science and the curated databases of the \emph{GET Evidence System} and the \emph{NHGRI GWAS Catalog}.
-Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments.
-An entry on SNPedia is valued 2.5 times as high as a PLoS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.
-
-Users are also able to link their Fitbit\cite{fitbit} accounts to their user-accounts.
-Fitbit is a commercial service which lets its customers track their BMI, movement and sleep data.
-This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body and sleep developments over time.
-
-\subsection*{Data access}
-openSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
-grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
-accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
-An example of how the DAS can be used is implemented on openSNP, where users' genotypes are visualized inside a genome browser.
-So far, all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now.
-\bastian{should this maybe include an image as well?}
-
-The data is additionally available over a JSON API, which allows users to directly access data in the JSON format.
-The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.
+These survey-results indicate that there is a definite interest in customers of DTC-companies to share their results with other scientists.
 
 \section*{Discussion}
+As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community,
+and through social media the results do not necessarily reflect the general population. 
+Instead it's expected that it over-represents those people most likely to be interested in a project such as openSNP: 
+Customers of DTC genetic testing companies and people with a high interest in biology.
 
-\subsection*{Survey issues}
-
-As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, the results do not reflect the general population, but over-represent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.
-
-\subsection*{Privacy, health implications and ethical considerations}
-
-%The advent of DTC genetic testing has led to new ethical and social issues. 
-Much of the criticism of DTC genetic testing focusses on the practice 
-of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
-and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.
-
-As we have found in our survey on sharing such results (see supplementary methods), many DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information.
-There is a variety of ethical and privacy implications when it comes to DTC genetic testing\cite{Caulfield2011,Joh2011}.
-
-Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
-or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
-\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
-widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.
-
-openSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, once during registration for openSNP and once during 
-the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
-of publishing their data. Further versions of openSNP may optionally include further consent processes.
-
-\subsection*{GWAS and Open Data}
-Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper.
-However, GWAS can only detect correlations of SNPs with those traits and do not allow
-inference on the cause for any correlation.
-Furthermore, for a statistically sound analysis, GWAS need a large enough sample size.
-Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
-
-One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. 
-Data produced by DTC genetic testing companies is a promising source for such results, as those companies already have high 
-numbers of customers which are willing to pay for the genotyping by themselves.
-
-By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other 
-open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP. 
-Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, 
-with users entering data in different measurement systems.
-Another problem with user-entered data is the frequent switching between categorical and continuous phenotypes - for example, some users entered the specific value of their height, while other users entered their height according to a category like "150cm to 160cm".
-
-While we try to suggest similar entries to the users, 
-there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.
-There are two possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes.
-The other one
-would be to make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
-variations that have not been available from the start.
-
-In both cases it makes it harder 
-for users to enter their data which raises the bar for participation.
-We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform additional quality control.
-
-Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP: only a subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, which can potentially lead to skewed GWAS-results.
-21 people, mainly from underrepresented demographics, have been offered free genotyping using funding provided by the Wikimedia Germany association in order to mitigate this bias.
-
-With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
-data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to freely use in their studies.
-Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
-While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise.
-By including the option of uploading exome data sets the platform already is capable of adjusting for changes in the type of data generated by DTC genetic testing.
-
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Materials and Methods}
-
-\subsection*{Survey on Sharing Genetic Information}
-The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
-blogs which focus on genetics and DTC genetic testing and social media websites like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.
-
-The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
-(planned) customership with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
-All participants were asked if they would be willing to share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
-but only if they did not share my medical information with anybody else" and No".
-
-The survey also asked some scaled questions, which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
-information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
-were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
-possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
-data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
-"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
-Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.
-
-The survey data was analyzed with SPSS 19.
-
-\subsection*{Technical implementation of the platform}
-The main platform is implemented using the web framework Ruby on Rails 3.0.10. Postgres 9.2 is used as the main database backend for Rails. 
-The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs 
-which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of 
-the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, 
-through querying the content via the MediaWiki API.
-The \emph{NHGRI GWAS Catalog} and the \emph{GET Evidence System} provide complete dumps in plain text formats.
-Those are regularly downloaded and parsed.
-SNPs that are described as 'Insufficiently evaluated' in the \emph{GET Evidence System} are not stored.
-All databases are queried or parsed using the unique identifier of each SNP as the search term.
-
-SNPs are catalogued by their unique identifier, which consists of a prefix (mostly \textit{rs}, rarely \textit{i}) and a unique number. This is a common format, 
-which is employed by the NCBI dbSNP database \cite{Sherry2001} and is also widely used and easily parsed from different literature sources. Publications from the different databases as 
-well as the users' genotypes are associated with individual SNPs by the Rs-ID. Allele and genotype frequencies are updated regularly, based on the data present in openSNP.
-
-Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources 
-are handled using the ruby gem Resque and the standalone key-value storage server Redis. Search features on the platform itself are implemented using Solr and the ruby gem Sunspot. 
-Additionally, data can be requested from openSNP using the Distributed Annotation System. The required data is stored in a PostgreSQL database.  
-Requested data is delivered in XML-format to facilitate parsing. Additionally, users can request data in the JSON-format, using a system not specified in any standard.
-
-openSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
-two methods are implemented: firstly \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
-based on the user's query. The second method is \textit{sources}, which advertises all DAS sources for all genotypes present in openSNP.
-
-A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in Figure \ref{Figure4_label}. The source code of openSNP is 
-published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
-% Do NOT remove this, even if you are not including acknowledgments
-\section*{Acknowledgments}
-We thank Dr.
-Manuel Corpas and Prof.
-David Edwards for constructive advice in grammar, spelling and structure of this study.
-Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser.
-For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams.
-We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.
-
-%\section*{References}
-% The bibtex filename
-\bibliography{papers}
-
-\section*{Figure Legends}
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/users.png}
-	\end{center}
-	\caption{
-	{\bf Growth of openSNP-user-accounts.} The increase in numbers for users from 27.09.2011 to 27.10.2012 is shown.} 
-	\label{Figure1_label}
-\end{figure}
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.35]{25_10_2012_Graphs/genotypes.png}
-	\end{center}
-	\caption{
-	{\bf Growth of available genotypings.} The increase in numbers for genotyping-files from 27.09.2011 to 27.10.2012 is shown.} 
-	\label{Figure2_label}
-\end{figure}
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.40]{25_10_2012_Graphs/phenotypes_vs_userphenotypes.png}
-	\end{center}
-	\caption{
-	{\bf Development of unique phenotypes and phenotypic information over time.} The x-axis shows the time-frame from start of the project until October 2012, the left y-axis shows how many unique phenotypes have been entered, and the right y-axis shows the amount of phenotypes users entered.}
-	\label{pheno}
-\end{figure}
-
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.5]{25_10_2012_Graphs/papers_dotchart.png}
-	\end{center}
-	\caption{
-	{\bf Distribution of annotation-sources at openSNP.} Currently, SNP-annotations from SNPedia, PLOS, Mendeley, the \emph{GET Evidence System} and the \emph{NHGRI GWAS Catalog} are being collected.}
-	\label{Figure3_label}
-\end{figure}
-
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[scale=0.1]{25_10_2012_Graphs/percent_oa.png}
-	\end{center}
-	\caption{
-	{\bf Ratio of Open Access Publications.} Green pieces are Open Access. The \emph{NHGRI GWAS Catalog} doesn't give information about the Open Access status.}
-	\label{oa_label}
-\end{figure}
-
-\begin{figure}[!ht]
-	\begin{center}
-	  \includegraphics[scale=0.3]{latest_uml.png}
-	\end{center}
-	\caption{
-	{\bf Flow of data inside openSNP.} External databases and user-provided data are used as input. Output of data is done using the website, the \emph{Distributed Annotation System} and a JSON-API.} 
-	\label{Figure4_label}
-\end{figure}
-
-%\begin{figure}[!ht]
-%\begin{center}
-%%\includegraphics[width=4in]{figure_name.2.eps}
-%\end{center}
-%\caption{
-%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption 
-%should be left justified, as specified by the options to the caption 
-%package.
-%}
-%\label{Figure_label}
-%\end{figure}
-
-\section*{Tables}
-\begin{table}
-\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped.
-}
-\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
-\hline
-& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
-\hline
-\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
-\hline 
-... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
-... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
-... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
-\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
-... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
-... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
-... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
-... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
-\end{tabular}
-\label{tab:motivations1}
-\end{table}
-
-\begin{table}
-\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
-\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
-\hline
-& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
-\hline
-\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
-\hline
-... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
-... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
-... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
-\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
-\hline
-... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
-... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
-... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
-\end{tabular}
-\label{tab:motivations2}
-\end{table}
-%\begin{table}[!ht]
-%\caption{
-%\bf{Table title}}
-%\begin{tabular}{|c|c|c|}
-%table information
-%\end{tabular}
-%\begin{flushleft}Table caption
-%\end{flushleft}
-%\label{tab:label}
-% \end{table}
+Due to those limitations it's not possible to get estimates of how many potential users could be acquired for a platform like openSNP.
 
